### 2017-08-25 agusuppinfosample.tex
First import for FITS paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supporting Information
%% (Optional)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% OVERVIEW
%
% Please note that all supporting information will be peer reviewed with your manuscript.
% In general, the purpose of the supporting information is to enable
% authors to provide and archive auxiliary information such as data
% tables, method information, figures, video, or computer software,
% in digital formats so that other scientists can use it.

% The key criteria are that the data:
% 1. supplement the main scientific conclusions of the paper but are not essential to the conclusions (with the exception of
%    including data so the experiment can be reproducible);
% 2. are likely to be usable or used by other scientists working in the field;
% 3. are described with sufficient precision that other scientists can understand them, and
% 4. are not exe files.
%

% All Supporting text and figures should be included in this document.

% Data sets, large tables, movie files,
% and audio files should be uploaded separately, following AGU naming
% conventions. Include their captions in this document and list the
% file name with the caption. You will be prompted to upload these
% files on the Upload Files tab during the submission process, using
% file type “Supporting Information (SI)”

\documentclass[draft]{agujournal}

% Please type in the journal name: \journalname{<Journal Name>}
% ie,
\journalname{Earth and Space Science}

%% Choose from this list of Journals:
%
% Journal of Geophysical Research
% JGR-Biogeosciences
% JGR-Earth Surface
% JGR-Planets
% JGR-Solid Earth
% JGR-Space Physics
% Global Biochemical Cycles
% Geophysical Research Letters
% Paleoceanography
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resource Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science

\begin{document}

%% This command needs article title as argument to \supportinginfo{}:
\supportinginfo{Trends in the predictive performance of raw ensemble weather forecasts}

 \authors{A. B. Smith\affil{1}\thanks{Current address, McMurdo Station,
 Antartica},
 Eric Brown\affil{1,2}, Rick Williams\affil{3},
 John B. McDougall\affil{4}, and S. Visconti\affil{5}\thanks{Also
 funded by Monsanto.}}

\affiliation{1}{Department of Hydrology and Water Resources,
University of Arizona, Tucson, Arizona, USA.}
\affiliation{2}{Department of Geography, Ohio State University,
Columbus, Ohio, USA.}
\affiliation{3}{Department of Space Sciences, University of
Michigan, Ann Arbor, Michigan, USA.}
\affiliation{4}{Division of Hydrologic Sciences, Desert Research
Institute, Reno, Nevada, USA.}
\affiliation{5}{Dipartimento di Idraulica, Trasporti ed
Infrastrutture Civili, Politecnico di Torino, Turin, Italy.}

%% Corresponding Author
%(include name and email addresses of the corresponding author.  More
%than one corresponding author is allowed in this Word file and for
%publication; but only one corresponding author is allowed in our
%editorial system.)

\correspondingauthor{A. B. Smith}{email@address.edu}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

\section*{Contents}
%%%Remove or add items as needed%%%
\begin{enumerate}
\item Text S1 to Sx
\item Figures S1 to Sx
\item Tables S1 to Sx
\end{enumerate}

\section*{Additional Supporting Information (Files uploaded separately)}

\begin{enumerate}
\item Captions for Datasets S1 to Sx
\item Captions for large Tables S1 to Sx (if larger than 1 page, upload as separate excel file)
\item Captions for Movies S1 to Sx
\item Captions for Audio S1 to Sx
\end{enumerate}

\section*{Introduction}

%Delete all unused file types below. 
%Copy/paste for multiples of each file type as needed.

\section*{Text S1.}

Type or paste text here. This should be additional explanatory text,
such as: extended descriptions of results, full details of models,
extended lists of acknowledgements etc.  It should not be
additional discussion, analysis, interpretation or critique. It
should not be an additional scientific experiment or paper.

Repeat for any additional Supporting Text

%%Enter Data Set, Movie, and Audio captions here
%%EXAMPLE CAPTIONS

\section*{Data Set S1.} %Type or paste caption here.

Upload your dataset(s) to AGU's journal submission site and select
"Supporting Information (SI)" as the file type. Following naming
convention: ds01.

Repeat for any additional Supporting data sets

\section*{Movie S1.}

Type or paste caption here.

Upload your movie(s) to AGU's journal submission site and select,
"Supporting Information (SI)" as the file type. Following naming convention: ms01.

Repeat any additional Supporting movies

\section*{Audio S1.}

Type or paste caption here.

Upload your audio file(s) to AGU's journal submission site and select
Supporting Information ``(SI)" as the file type. Following naming
convention: auds01.

Repeat for any additional Supporting audio files

\end{document}

### 2017-08-25 epss_marmo.tex
First import for FITS paper

%% 2/18/2016
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalSample.tex: this sample file is for articles formatted with LaTeX
%
% This sample file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements.
%
% PLEASE DO NOT USE YOUR OWN MACROS
% DO NOT USE \newcommand, \renewcommand, or \def.
%
% FOR FIGURES, DO NOT USE \psfrag or \subfigure.
% DO NOT USE \psfrag or \subfigure commands.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% All questions should be e-mailed to latex@agu.org.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Step 1: Set the \documentclass
%
% There are two options for article format:
%
% 1) PLEASE USE THE DRAFT OPTION TO SUBMIT YOUR PAPERS.
% The draft option produces double spaced output.
% 
% 2) numberline will give you line numbers.

% Tip:
%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

%% To submit your paper:
\documentclass[linenumbers,draft]{agujournal}

% Now, type in the journal name: \journalname{<Journal Name>}
% ie,
\journalname{JGR-Atmospheres}

%% Choose from this list of Journals:
%
% JGR-Atmospheres
% JGR-Biogeosciences
% JGR-Earth Surface
% JGR-Oceans
% JGR-Planets
% JGR-Solid Earth
% JGR-Space Physics
% Global Biochemical Cycles
% Geophysical Research Letters
% Paleoceanography
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resource Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science

%% ------------------------------------------------------------------------ %%
%
%  ENTER Title Page commands:
%
%% ------------------------------------------------------------------------ %%

% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general terms then specific results are optimized in
% searches)

% Example: \title{This is a test title}

% (List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.  
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% \affiliation{1}{First Affiliation}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

%% Keypoints, final entry on title page.
% Example: 
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 100 characters or less with no special
% characters or punctuation 
% \end{keypoints}

%% \begin{abstract} begins second page

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}
% To explain why change was made: \explain{<explanation>}

% At the end of the document, use \listofchanges, which will list the
% changes and the page and line number where the change was made.

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.
% \explain will not print anything.

% Optional argument:
% You can also add additional information to be printed with the list
% of changes, to indicate the initials of the person changing the text,
% and the time and/or date of the change, or any other comment by using
% the optional [] argument:
% \added[AH, 3:30pm, Feb 18, 2016]{added term}
% will yield 
% [AH, 3:30pm, Feb 18, 2016] added term on page...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% ------------------------------------------------------------------------ %%
%
%  TITLE
%
%% ------------------------------------------------------------------------ %%

\title{Trends in the predictive performance of raw ensemble weather forecasts}

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

 \authors{A. B. Smith\affil{1}\thanks{Current address, McMurdo Station,
 Antartica},
 Eric Brown\affil{1,2}, Rick Williams\affil{3},
 John B. McDougall\affil{4}, and S. Visconti\affil{5}\thanks{Also
 funded by Monsanto.}}

\affiliation{1}{Department of Hydrology and Water Resources,
University of Arizona, Tucson, Arizona, USA.}
\affiliation{2}{Department of Geography, Ohio State University,
Columbus, Ohio, USA.}
\affiliation{3}{Department of Space Sciences, University of
Michigan, Ann Arbor, Michigan, USA.}
\affiliation{4}{Division of Hydrologic Sciences, Desert Research
Institute, Reno, Nevada, USA.}
\affiliation{5}{Dipartimento di Idraulica, Trasporti ed
Infrastrutture Civili, Politecnico di Torino, Turin, Italy.}

%% Corresponding Author
%(include name and email addresses of the corresponding author.  More
%than one corresponding author is allowed in this Word file and for
%publication; but only one corresponding author is allowed in our
%editorial system.)

\correspondingauthor{A. B. Smith}{email@address.edu}

%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 100 characters or less with no special characters or punctuation

\begin{keypoints}
\item Evolution of raw ensemble forecast \replaced{skill}{skills}
\item Future benefits from statistical post-processing
\item Global distribution of forecast skill development
\end{keypoints}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT
%
%% ------------------------------------------------------------------------ %%

\begin{abstract}
This study applies statistical post-processing to ensemble
forecasts of near-surface temperature, 24-hour 
%% below are examples of track changes:
\deleted[date/time, etc.]{precipitation},
\added{all} totals,
and near-surface wind speed from the global model of the European
Centre for Medium-Range Weather Forecasts (ECMWF). The main objective
is to evaluate the evolution of the difference in skill between the
raw ensemble and the post-processed forecasts. Reliability and
sharpness, and 
%% another example of track changes.
\replaced{hence}{therefore} skill,
 of the former is expected to improve over
time. Thus, the gain by post-processing is expected to decrease. Based
on ECMWF forecasts from January 2002 to March 2014 and corresponding
observations from globally distributed stations we generate
post-processed forecasts by ensemble model output statistics
%% another example of track changes.
\added{abbreviated as} (EMOS)
for each station and variable. Given the higher average skill of the
post-processed forecasts, we analyse the evolution of the difference
in skill between raw ensemble and EMOS. This skill gap remains almost
constant over time indicating that post-processing will keep adding
skill in the foreseeable future.
\end{abstract}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

%%% Suggested section heads:
% \section{Introduction}
% 
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.

% \subsection{A descriptive heading about methods}
% More about Methods.
% 
% \section{Data} (Or section title might be a descriptive heading about data)
% 
% \section{Results} (Or section title might be a descriptive heading about the
% results)
% 
% \section{Conclusions}

\section{Introduction}
\label{sec:intro}
Over the last two decades the paradigm in weather forecasting has
shifted from being deterministic to probabilistic \citep[see
e.g.][]{Palm00,Hami00}. Accordingly, numerical weather prediction
(NWP) models have been run increasingly as ensemble forecasting
systems. The goal of such ensemble forecasts is to approximate the
forecast probability distribution by a finite sample of scenarios
\citet{Leit74}\footnote{See \citet{Leit74} for a more in-depth
description of these issues and their complex implications.}
Global ensemble forecast systems, like the European
Centre for Medium-Range Weather Forecasts (ECMWF) ensemble, are prone
to probabilistic biases, and are therefore not reliable. They
particularly tend to be underdispersive for surface weather parameters
\citet{Boug10, Park08}. In order to correct for forecast
underdispersion and bias in NWP ensembles different
statistical
post-processing methods have been developed, of which ensemble model
output statistics (EMOS) \citep{Gneit05} is among the most widely
applied. EMOS yields a parametric forecast distribution by linking its
parameters to ensemble statistics. Due to its simplicity and low
computational cost, we focus on EMOS for this study.

The ECMWF ensemble is under continuous development, and hence its
forecast skill improves over time \citep{Buiz98,Buiz07,Rich13,Haid14}.
Parts of these improvements may be due to a reduction of probabilistic
biases. From this we deduce the following hypothesis: As the raw
forecasts continuously improve, it is hypothesized that the gap in
skill between raw ensemble and post-processed forecasts narrows,
because systematic errors typically captured by post-processing are
reduced 
by those improvements. 
%%
%% Use \explain{} before change, so that it ends up on the right line.
\explain[Jon, 2/16/16]{Redundant sentence, better without it} 
\deleted{In other words, probabilistic biases,
which can be reduced by statistical post-processing methods, decrease
over time.}
Assuming that the raw ensemble forecasts continue to
improve in the future, the gap in skill may eventually be closed when
the raw ensemble forecasts become reliable and unbiased.  In this work
we analyse the evolution of the global performance of the operational
ECMWF raw ensemble and the corresponding post-processed EMOS forecasts
for 2 metre temperature (T2M), 24-hour precipitation (PPT24), and 10-m
wind speed (V10). We verify the forecasts against globally distributed
surface synoptic observations (SYNOP) data over a period of about 10
years. We firstly evaluate the monthly average skill in terms of CRPS
for both the raw and the EMOS forecasts. In order to assess the extent
to which the results depend on the choice of the post-processing
method, Bayesian model averaging (BMA), \citet{Raft05,Fra10} is
additionally applied to the T2M raw ensemble forecasts. We will use
the negatively oriented (i.e. the lower the value the higher the
skill) continuous ranked probability score (CRPS) \citep{Math76} as a
measure of skill. As the CRPS assesses both reliability and sharpness
and is a proper score \citep{Gneit07b}, we rely on it for model fitting
and verification throughout this study. Note that skill and
reliability are linked in that given constant sharpness an improvement
in reliability leads to an improvement in skill and vice versa. We
finally analyse the evolution of the gap in CRPS between raw ensemble
and post-processed forecasts.

After presenting the dataset in section~\ref{sec:intro} we summarize
the methods for post-processing and for the assessment of the global
skill evolution in section~\ref{sec:meth}. In section~\ref{sec:res}
the results are shown.  This is followed by a discussion in
section~\ref{sec:disc} along with some concluding
remarks. These
analyses have been performed using the statistical software \texttt{R}
\citep{R2013}.

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}
% \end{figure}

\begin{figure}[ht!]
% \centering
% when using pdflatex, use pdf file:
% \includegraphics[width=20pc]{figsamp.pdf}
%
% when using dvips, use .eps file:
% \includegraphics[width=20pc]{figsamp.eps}
%
% If you don't specify the file extension, 
% \includegraphics will insert the right file:
\centerline{\includegraphics[height=1.5in]{figsamp}}
\caption{Short caption}
\label{figone}
\end{figure}

\begin{figure}[ht!]
\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
\caption{The figure caption should begin with an overall descriptive
statement of the figure followed by additional text. They should be
immediately after each figure.  Figure parts are indicated with
lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
both the figures and captions in the text near where they are cited.}
\label{fig:mean_and_slope}
\end{figure}

\begin{table}
\caption{Start this caption with a short description of your
table.
Large tables
especially presenting rich data should be presented as separate excel
or .cvs files, not as part of the main text.}
== Table Here ==
\end{table}

%
% ---------------
% EXAMPLE TABLE
%
\begin{table}
\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
\centering
\begin{tabular}{l c}
\hline
 Run  & Time (min)  \\
\hline
  $l1$  & 260   \\
  $l2$  & 300   \\
  $l3$  & 340   \\
  $h1$  & 270   \\
  $h2$  & 250   \\
  $h3$  & 380   \\
  $r1$  & 370   \\
  $r2$  & 390   \\
\hline
\multicolumn{2}{l}{$^{a}$Table note text here.}
\end{tabular}
\end{table}

% See below for how to make sideways figures or tables.

% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
 \begin{sidewaysfigure}
 \includegraphics[width=20pc]{figsamp}
 \caption{Caption here}
 \label{newfig}
 \end{sidewaysfigure}
%
 \begin{sidewaystable}
 \caption{Caption here}
\label{tab:signif_gap_clos}
 \begin{tabular}{ccc}
one&two&three\\
four&five&six
 \end{tabular}
 \end{sidewaystable}

\clearpage

\section{Methods}
\label{sec:meth}

\subsection{Post-processing Using EMOS}
Post-processing using EMOS converts a raw ensemble of discrete
forecasts into a probability distribution. Let $y$ be the variable to
be forecast (here: T2M, PPT24 or V10) and let ${f}=(f_{1}, f_{2},
\dots, f_{K})^{T}$ be the vector of the $K$ member raw ensemble
forecasts (here: HRES, ENS, and CTRL). Then the EMOS
\added{predictive}
density can be written as:
\begin{linenomath*}
\begin{equation}
y|{f} \sim g(m, \sigma),
\end{equation}
\end{linenomath*}
where $g(\cdot)$ is a parametric density function with location and
scale parameters $m$ and $\sigma$, respectively, which depend on the
raw ensemble.

\subsubsection{Temperature}
For T2M forecasts $g(\cdot)$ is a normal density distribution with
mean $m$ and variance $\sigma^{2}$. Here, we use a variant of the
original EMOS approach similar to the one proposed by
\citet{ScheueBue13} where the departures of observed temperatures from
their climatological means are related to those of the forecasts.
Specifically, let $T=\{t_{1},\ldots,t_{n}\}$ be a training period of
$n$ days preceding the forecast initialization and denote by $f_{tk}$
the forecast of the k-th ensemble member and by $y_{t}$ the
observation on day $t\in T$. As a first step, we fit a regression
model
%%% \begin{linenomath*} ...\end{linenomath*} puts a line number in equation:
\begin{linenomath*}
\begin{equation}
y_{t_{j}} = c_{0} + c_{1}\sin\left(\frac{2\pi j}{365}\right) +
c_{2}\cos\left(\frac{2\pi j}{365}\right) + \varepsilon_{t_{j}}, \quad
j=1,\ldots,n
\end{equation}
\end{linenomath*}
which captures the seasonal variation of T2M. The residual terms
$\varepsilon_{t_{j}}$ are likely correlated over time, but for
simplicity an ordinary least squares fit is performed. We denote by
$\tilde{y}_{t}$ the fitted value of this periodic regression model on
day $t$ and interpret it as the climatological mean temperature on
this day. This model can easily be extrapolated to future days
$t_{d+1}, t_{d+2}, \ldots$ The above regression includes both a sine
and a cosine term which is equivalent to a cosine model with variable
phase and amplitude. Since $j=1,\ldots,n$ is just a numbering of the
days in $T$, different training periods have different phase
parameters and hence $c_{1}$ and $c_{2}$ evolve over the calendar
year. We fit the same type of model also to the ensemble mean,
control, and high resolution run and obtain climatological means
$\tilde{f}_{\overline{\mathrm{ENS}},t}, \tilde{f}_{\mathrm{CTRL},t}$,
and $\tilde{f}_{\mathrm{HRES},t}$. The mean of the forecast
distribution is then:
\begin{linenomath*}
\begin{equation}
m = \tilde{y} + a_{1}(f_{\mathrm{HRES}} - \tilde{f}_{\mathrm{HRES}}) +
a_{2}(f_{\mathrm{CTRL}} - \tilde{f}_{\mathrm{CTRL}}) +
a_{3}(f_{\overline{\mathrm{ENS}}} -
\tilde{f}_{\overline{\mathrm{ENS}}}).
\end{equation}
\end{linenomath*}
The variance of the forecast distribution is linked to the raw
ensemble by: 
\begin{linenomath*}
\begin{equation}
\sigma^{2} = b_{0} + b_{1}s^{2},
\end{equation}
\end{linenomath*}
where $s^{2} = \frac{1}{K}\sum_{k=1}^{K}(f_{k} -
\frac{1}{K}\sum_{k=1}^{K}f_{k})^{2}$. The parameters
${\theta}_{T2M}=(a_{1}, a_{2}, a_{3}, b_{0}, b_{1})^{T}$ are
constrained to be non-negative, and hence $a_{k}/\sum_{k=1}^{K} a_{k}$
can be understood as the weight of model $k$.

\subsubsection{Precipitation}
For PPT24 we use the EMOS approach proposed by \citet{Scheu14}, where
$g(\cdot)$ is a left-censored (at zero) generalized extreme value
(GEV) distribution. While the shape parameter $\xi$ of the GEV is kept
constant ($\xi=0.2$), the location and the scale parameters $m$ and
$\sigma$ are linked to the raw ensemble via:
\begin{linenomath*}
\begin{eqnarray}
m &=& a_{0} + a_{1} f_{\mathrm{HRES}} + a_{2} f_{\mathrm{CTRL}} +
a_{3} f_{\overline{\mathrm{ENS}}} + a_{4} \pi_{0},\nonumber\\
\sigma &=& b_{0} + b_{1} \mathrm{MD}_{f},
\end{eqnarray}
\end{linenomath*}
where $\pi_{0}$ is the fraction of ensemble members predicting zero
precipitation and $\mathrm{MD}_{f}:= K^{-2}$\break
$\sum_{k, k' = 1}^{K} |
f_{k} - f_{k'} |$ is the ensemble mean difference. Again, the
parameters are denoted by $\theta_{PPT24}=(a_{0}, \dots, a_{4}, b_{0},
b_{1})^{T}$. The parameters $a_{1}, a_{2}, a_{3}, b_{0}, b_{1}$ are
constrained to be non-negative, and hence the normalized parameters
$a_{1}$ to $a_{3}$ can be understood as weights.

\subsection{Global CRPS Analysis}
 \label{sec:meth_glob_crps}

As stated in the introduction, the main objective of this study is to
analyse whether the gap in CRPS between the raw ensemble and the
post-processed forecast narrows over time. This is assessed
station-wise using both a parametric and a non-parametric approach.
For the former, we fit the following regression model to the monthly
time series of CRPS differences ($\Delta \mathrm{CRPS}_{t} =
\mathrm{CRPS}_{\mathrm{raw},t} - \mathrm{CRPS}_{\mathrm{EMOS}, t}$):
\begin{linenomath*}
\begin{equation}
\Delta \mathrm{CRPS}_{t} = \beta_{0} + \beta_{1} t + \beta_{2} \sin\left(\frac{2\pi t}{12}\right) + \beta_{3} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2}) \label{eq:linseas}
\end{equation}
\end{linenomath*}
where $\Delta \mathrm{CRPS}_{t}$ is the predictand, $t$ is now the
time in months, and $\sigma^{2}$ denotes the error variance. For the
latter, we use Kendall's $\tau$ correlation coefficient and the
associated test statistics \citep{Mann45} as implemented in the
\texttt{R} package \texttt{Kendall} \citep{Kendall}. In order to
correct for seasonal effects, we calculate the $\tau$ statistics using
the residuals of the following model:
\begin{linenomath*}
\begin{equation}
\Delta \mathrm{CRPS}_{t} = \gamma_{0} + \gamma_{1} \sin\left(\frac{2\pi t}{12}\right) + \gamma_{2} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2})
\label{eq:tauseas}
\end{equation}
\end{linenomath*}
Note that negative $\tau$ values indicate a negative trend and
positive values a positive one. Figure~\ref{figone} a) and b)
show the regression lines estimated by model (\ref{eq:linseas}) for
monthly averages of $\Delta \mathrm{CRPS}$ and the corresponding
Kendall's $\tau$ test statistics for an example with decreasing and
increasing gap.

\section{Results}
\label{sec:res}

\subsection{Are There Any Significant Temporal Trends?}

The above results indicate a tendency of a decrease in $\Delta$CRPS
over time at least for T2M and PPT24. In the following we check the
percentages of stations with decreasing, an absence of, or increasing
trend in $\Delta$CRPS over time at a significance level of 0.05. In
order to be more confident about the results this analysis is
performed using both the parametric regression model and the
non-parametric Kendall's $\tau$ correlation coefficient test. As
already mentioned both approaches correct for seasonal effects.
Furthermore, in case of T2M the same analysis has been performed
additionally using BMA instead of EMOS in order to relax the
dependence on one particular post-processing method. As shown in
Table~\ref{tab:signif_gap_clos} the stations with no significant trend
outnumber the stations with either negative or positive trend for all
three variables and lead times considered. Note that the percentage of
stations without any significant trend increases with increasing lead
time. In line with the results shown in
Figure~\ref{fig:mean_and_slope}, significantly negative trends are
more common than positive ones for T2M and PPT24. The difference
between the number of stations with negative and those with positive
trend reduces with increasing lead time, but is still greater than
zero for a 10 day forecast. Note that the high number of
non-significant stations in case of PPT24 is likely to be due to the
high variability of precipitation amounts, and hence variability of
CRPS values, which leads to a large residual standard error in case of
the parametric regression model and to a lot of pairs (a pair denotes
here a value of $\Delta$CRPS and its associated time stamp) opposite
to the estimated direction in case of the $\tau$ test statistics. In
case of V10 the stations with a negative trend and those with a
positive trend are almost equally frequent regardless of the lead
time. Figures of the global distributions of stations with no,
significantly negative, and significantly positive trend in
$\Delta$CRPS are available as supplemental material to this paper.

\section{Discussion and Conclusions}
\label{sec:disc}

According to the above analyses the gap in CRPS between the raw
ensemble and the EMOS forecasts remains almost constant over time. For
T2M and PPT24 $\Delta$CRPS shows a slightly decreasing tendency. The
higher the lead time the less accentuated is this tendency. For V10
such a tendency cannot be detected. The parametric regression model
and the non-parametric $\tau$ test yield similar results. Hence, a
linear model that is overlaid by seasonal fluctuations seems to be
reasonable. Note that the skill of the raw ensemble and the EMOS
forecasts may sometimes be negatively affected by upgrades to the
atmospheric model. Model upgrades may deteriorate raw ensemble skill
at some individual stations. For instance, a resolution increase may
introduce new issues with statistical downscaling of the forecasts to
some specific observation sites. But more importantly, the skill of
the post-pro\-cessed forecasts can be lowered dramatically if a model
update happens between the training and the verification period. These
issues may result in positive trends in $\Delta$CRPS. Ideally,
post-processing would be based on a cascade of reforecasts. That is,
for each atmospheric model version, training of the post-processing
model would be done using a corresponding time series of reforecasts
made with that same model version. Furthermore, the observations may
be affected by measurement errors. If these errors change over time,
they may also influence the estimates of the trends in $\Delta$CRPS.
As the problems introduced by statistical downscaling may be mitigated
by verifying against model analysis, a similar study that replaces
observations by model analysis, as proposed by \citet{Ghel00} and
\citet{Papp09}, may give further insights.

From the above we conclude that the probabilistic skill of both the
raw ensembles and the EMOS forecasts improves over time. The fact that
the gap in skill has remained almost \textbf{constant}, especially for
V10, suggests that improvements to the atmospheric model have an
effect quite different from what calibration by statistical
post-processing is doing. That is, they are increasing potential
skill. Thus this study indicates that (a) further model development is
important even if one is just interested in point forecasts, and (b)
statistical post-processing is important because it will keep adding
skill in the foreseeable future.

%% ------------------------------------------------------------------------ %%
\section*{Citations}

% Please use ONLY \citet and \citep for reference citations.
% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).

\subsection*{Cites made with \tt\string\citet\string{\string}}
...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
\citet{Ghel00}, and \citet{Leit74}.

\subsection*{Cites made with \tt\string\citep\string{\string}}
...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
\citep{Ghel00, Leit74}.

...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.

%%% End of body of article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendix goes here
%
% \appendix resets counters and redefines section heads
% but doesn't print anything.
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
\appendix
\section{Here is a sample appendix}
This is an Appendix section.

\subsection{subsection}
This is an Appendix subsection.

\subsubsection{subsubsection}
This is an Appendix subsubsection.
\begin{linenomath*}
\begin{equation}asdf\end{equation}
\end{linenomath*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Optional Glossary, Notation or Acronym section goes here:
%
%%%%%%%%%%%%%%
% Glossary is only allowed in Reviews of Geophysics
\begin{glossary}
\term{Term}
 Term Definition here
\term{Term}
 Term Definition here
\term{Term}
 Term Definition here
\end{glossary}

%
%%%%%%%%%%%%%%
% Acronyms
 \begin{acronyms}
 \acro{Acronym}
 Definition here
 \acro{EMOS}
 Ensemble model output statistics 
 \acro{ECMWF}
 Centre for Medium-Range Weather Forecasts
 \end{acronyms}

%
%%%%%%%%%%%%%%
% Notation 
 \begin{notation}
 \notation{$a+b$} Notation Definition here
 \notation{$e=mc^2$} 
 Equation in German-born physicist Albert Einstein's theory of special
relativity that showed that the increased relativistic mass ($m$) of a
body comes from the energy of motion of the body—that is, its kinetic
energy ($E$)—divided by the speed of light squared ($c^2$).
 \end{notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  ACKNOWLEDGMENTS

\acknowledgments
The forecast data used in this study can be made available subject to
a handling charge. SYNOP observations are not available through ECMWF.
We are grateful to F. Rabier, D. S. Richardson, D. Lavers, and other
colleagues at ECMWF for helpful discussions and inputs. We thank David
M\"{o}ller of the University of Heidelberg for his contribution.
Furthermore, we like to thank T. Gneiting of the Heidelberg Institute
for Theoretical Studies and of the Institute for Stochastics at
Karlsruhe Institute of Technology for valuable comments, and the
colleagues at the Heidelberg Institute for Theoretical Studies.
Additionally, we are grateful to two anonymous reviewers for their
helpful comments. We are indebted to the ECMWF for granting access to
their products.

The Editor thanks Pierre Pinson and an anonymous reviewer for their assistance in evaluating this paper.

%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
%
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% Or, to use BibTeX:
%
% Follow these steps
%
% 1. Type in \bibliography{<name of your .bib file>} 
%    Run LaTeX on your LaTeX file.
%
% 2. Run BiBTeX on your LaTeX file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Run LaTeX on your new file which will produce the citations.
%

\bibliography{agusample}

% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.

% % \begin{thebibliography}{}

% % \bibitem[{\textit{Bell and Munoz}}(2008)]{Boug10} Bell, A.~H., and
% % Munoz, D.~P.  (2008). Activity in the superior colliculus reflects
% % dynamic interactions between voluntary and involuntary influences on
% % orienting behaviour. \textit{Eur. J. Neurosci.} 28, 1654--1660.

% % \bibitem[{\textit{Corbetta et~al.}}(1991)]{Buiz07} Corbetta, M.,
% % Miezin, F.~M., Dobmeyer, S., Shulman, G.~L., and Petersen, S.~E.
% % (1991). Selective and divided attention during visual discriminations
% % of shape, color, and speed: functional anatomy by positron emission
% % tomography. \textit{J.~Neurosci.} 11, 2383--2402.

% % \bibitem[{\textit{Borra et~al.}}(2014)]{Buiz98} Borra, E., Gerbella,
% % M., Rozzi, S., Tonelli, S., and Luppino, G.  (2014). Projections to
% % the superior colliculus from inferior parietal, ventral premotor, and
% % ventrolateral prefrontal areas involved in controlling goal-directed
% % hand actions in the macaque. \textit{Cereb. Cortex} 24, 1054--1065.

% % \bibitem[{\textit{Dorris et~al.}}(1997)]{Fra10}
% %  Dorris, M.~C.,
% % Par\'{e}, M., and Munoz, D.~P.  (1997). Neuronal activity in monkey
% % superior colliculus related to the initiation of saccadic eye
% % movements. \textit{J.~Neurosci.} 17, 8566--8579.

% % \bibitem[{\textit{Elsabbagh et~al.}}(2009)]{Ghel00} 
% % Elsabbagh, M.,
% % Volein, A., Holmboe, K., Tucker, L., Csibra, G., Baron-Cohen, S.,
% % et~al.  (2009). Visual orienting in the early broader autism
% % phenotype: disengagement. \textit{J.~Child Psychol. Psychiatry} 50,
% % 637--642.

% % \bibitem[{\textit{Fortin et~al.}}(1999)]{Gneit05} Fortin, S., Chabli,
% % A., Dumont, I., Shumikhina, S., Itaya, S.~K., and Molotchnikoff, S.
% % (1999). Maturation of visual receptive field properties in the rat
% % superior colliculus. \textit{bibain Res. Dev. Brain Res.} 1112,
% % 55--64.

% % \bibitem[{\textit{Felsen and Mainen}}(2008)]{Gneit07b} Felsen, G., and
% % Mainen, Z.~F.  (2008). Neural substrates of sensory-guided locomotor
% % decisions in the rat superior colliculus. \textit{Neuron} 60,
% % 137--148.

% % \bibitem[{\textit{Gattass and Desimone}}(1996)]{Haid14} Gattass, R.,
% % and Desimone, R.  (1996). Responses of cells in the superior
% % colliculus during performance of a spatial attention task in the
% % macaque. \textit{Rev. Bras. Biol.} 56, 257--279.

% % \bibitem[{\textit{Goldberg and Wurtz}}(1972)]{Hami00} Goldberg, M.~E.,
% % and Wurtz, R.~H.  (1972). Activity of superior colliculus in behaving
% % monkey. II. Effect of attention on neuronal responses.
% % \textit{J.~Neurophysiol.} 35, \hbox{560--574}.

% % \bibitem[{\textit{Krauzlis}}(2003)]{Kendall} Krauzlis, R.~J.  (2003).
% % Neuronal activity in the rostral superior colliculus related to the
% % initiation of pursuit and saccadic eye movements.
% % \textit{J.~Neurosci.} 23, 4333--4344.

% % \bibitem[{\textit{Heesy}}(2009)]{Leit74} 
% % Heesy, C.~P.  (2009). Seeing
% % in stereo: the ecology and evolution of primate binocular vision and
% % stereopsis. \textit{Evol. Anthropol.} 18, 21--35.

% % \bibitem[{\textit{Hilbig et~al.}}(2000)]{Mann45} Hilbig, H., Bidmon,
% % H.~J., Ettrich, P., and M\"{u}ller, A.  (2000). Projection neurons in
% % the superficial layers of the superior colliculus in the rat: a
% % topographic and quantitative morphometric analysis.
% % \textit{Neuroscience} 96, 109--119.

% % \bibitem[{\textit{Ignashchenkova et~al.}}(2004)]{Math76}
% % Ignashchenkova, A., Dicke, P.~W., Haarmeier, T., and Their, P.
% % (2004). Neuron-specific contribution of the superior colliculus to
% % overt and covert shifts of attention. \textit{Nat. Neurosci.} 7,
% % 56--64.

% \bibitem[{\textit{Krauzlis et~al.}}(2013)]{Palm00} Krauzlis, R.~J.,
% Lovejoy, L.~P., and Z\'{e}non, A.  (2013). Superior colliculus and
% visual spatial attention. \textit{Annu. Rev. Neurosci.} 36, 165--182.

% \bibitem[{\textit{Kustov and Robinson}}(1996)]{Papp09} Kustov, A.~A.,
% and Robinson, D.~L.  (1996). Shared neural control of attentional
% shifts and eye movements. \textit{Nature} 384, 74--77.

% \bibitem[{\textit{Landry and Bryson}}(2004)]{Park08} Landry, R., and
% Bryson, S.~E.  (2004). Impaired disengagement of attention in young
% children with autism. \textit{J.~Child Psychol. Psychiatry} 45,
% 1115--1122.

% \bibitem[{\textit{Kobayashi et~al.}}(2003)]{R2013} Kobayashi, T.,
% Tran, A.~H., Nishijo, H., Ono, T., and Matsumoto, G.  (2003).
% Contribution of hippocampal place cell activity to learning and
% formation of goal-directed navigation in rats. \textit{Neuroscience}
% 117, 1025--1035.

% \bibitem[{\textit{McHaffie et~al.}}(2005)]{Raft05} McHaffie, J.~G.,
% Stanford, T.~R., Stein, B.~E., Coizet, V., and Redgrave, P.  (2005).
% Subcortical loops through the basal ganglia. \textit{Trends Neurosci.}
% 28, 401--407.

% \bibitem[{\textit{McPeek and Keller}}(2004)]{Rich13} McPeek, R.~M.,
% and Keller, E.~L.  (2004). Deficits in saccade target selection after
% inactivation of superior colliculus. \textit{Nat. Neurosci.} 7,
% 757--763.

% \bibitem[{\textit{M\"{u}ller et~al.}}(2005)]{Scheu14} M\"{u}ller,
% J.~R., Philiastides, M.~G., and Newsome, W.~T.  (2005).
% Microstimulation of the superior colliculus focuses attention without
% moving the eyes. \textit{Proc. Natl. Acad. Sci. U.S.A.} 102, 524--529.

% \bibitem[{\textit{Munoz and Istvan}}(1998)]{ScheueBue13} Munoz, D.~P.,
% and Istvan, P.~J.  (1998). Lateral inhibitory interactions in
% the intermediate layers of the monkey superior colliculus.
% \textit{J.~Neurophysiol.} 79, 1193--1209.

% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}

% At the end of the document, use \listofchanges, which will list the
% changes and the page and line number where the change was made.

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.

%%%
\listofchanges
%%%

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supporting Information
%% (Optional) See AGUSuppInfoSamp.tex/pdf for requirements 
%% for Supporting Information.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

More Information and Advice:

%% ------------------------------------------------------------------------ %%
%
%  SECTION HEADS
%
%% ------------------------------------------------------------------------ %%

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%% ------------------------------------------------------------------------ %%
%
%  EQUATION NUMBERING: COUNTER
%
%% ------------------------------------------------------------------------ %%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

### 2017-08-25 epss_marmo.tex
Update on Overleaf.

@@
 % ie,
-\journalname{JGR-Atmospheres}
+\journalname{Earth and Space Science}
 
@@
 
-\title{Trends in the predictive performance of raw ensemble weather forecasts}
+\title{FITS format for planetary surfaces: definitions, applications and best practices.}
 
@@
 
- \authors{A. B. Smith\affil{1}\thanks{Current address, McMurdo Station,
- Antartica},
- Eric Brown\affil{1,2}, Rick Williams\affil{3},
- John B. McDougall\affil{4}, and S. Visconti\affil{5}\thanks{Also
- funded by Monsanto.}}
-
-\affiliation{1}{Department of Hydrology and Water Resources,
-University of Arizona, Tucson, Arizona, USA.}
-\affiliation{2}{Department of Geography, Ohio State University,
-Columbus, Ohio, USA.}
-\affiliation{3}{Department of Space Sciences, University of
-Michigan, Ann Arbor, Michigan, USA.}
-\affiliation{4}{Division of Hydrologic Sciences, Desert Research
-Institute, Reno, Nevada, USA.}
-\affiliation{5}{Dipartimento di Idraulica, Trasporti ed
-Infrastrutture Civili, Politecnico di Torino, Turin, Italy.}
+ \authors{C. Marmo\affil{1},
+ T. M. Hare\affil{2}, ...}
+
+\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay, Orsay, France}
+\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center, Flagstaff, AZ, USA}
+%\affiliation{3}{}
+%\affiliation{4}{}
+%\affiliation{5}{}
 
@@
 
-\correspondingauthor{A. B. Smith}{email@address.edu}
+\correspondingauthor{C. Marmo}{chiara.marmo@u-psud.fr}
 
@@
 \begin{keypoints}
-\item Evolution of raw ensemble forecast \replaced{skill}{skills}
-\item Future benefits from statistical post-processing
-\item Global distribution of forecast skill development
+\item Inter-operable data formats 
+\item 
 \end{keypoints}

### 2018-01-12 epss_marmo.tex
Update on Overleaf.

@@
 \item Inter-operable data formats 
+\item Planetary data
 \item 
@@
 \begin{abstract}
-This study applies statistical post-processing to ensemble
-forecasts of near-surface temperature, 24-hour 
-%% below are examples of track changes:
-\deleted[date/time, etc.]{precipitation},
-\added{all} totals,
-and near-surface wind speed from the global model of the European
-Centre for Medium-Range Weather Forecasts (ECMWF). The main objective
-is to evaluate the evolution of the difference in skill between the
-raw ensemble and the post-processed forecasts. Reliability and
-sharpness, and 
-%% another example of track changes.
-\replaced{hence}{therefore} skill,
- of the former is expected to improve over
-time. Thus, the gain by post-processing is expected to decrease. Based
-on ECMWF forecasts from January 2002 to March 2014 and corresponding
-observations from globally distributed stations we generate
-post-processed forecasts by ensemble model output statistics
-%% another example of track changes.
-\added{abbreviated as} (EMOS)
-for each station and variable. Given the higher average skill of the
-post-processed forecasts, we analyse the evolution of the difference
-in skill between raw ensemble and EMOS. This skill gap remains almost
-constant over time indicating that post-processing will keep adding
-skill in the foreseeable future.
+
 \end{abstract}
@@
 \label{sec:intro}
-Over the last two decades the paradigm in weather forecasting has
-shifted from being deterministic to probabilistic \citep[see
-e.g.][]{Palm00,Hami00}. Accordingly, numerical weather prediction
-(NWP) models have been run increasingly as ensemble forecasting
-systems. The goal of such ensemble forecasts is to approximate the
-forecast probability distribution by a finite sample of scenarios
-\citet{Leit74}\footnote{See \citet{Leit74} for a more in-depth
-description of these issues and their complex implications.}
-Global ensemble forecast systems, like the European
-Centre for Medium-Range Weather Forecasts (ECMWF) ensemble, are prone
-to probabilistic biases, and are therefore not reliable. They
-particularly tend to be underdispersive for surface weather parameters
-\citet{Boug10, Park08}. In order to correct for forecast
-underdispersion and bias in NWP ensembles different
-statistical
-post-processing methods have been developed, of which ensemble model
-output statistics (EMOS) \citep{Gneit05} is among the most widely
-applied. EMOS yields a parametric forecast distribution by linking its
-parameters to ensemble statistics. Due to its simplicity and low
-computational cost, we focus on EMOS for this study.
 

### 2018-01-12 epss_marmo.tex

@@
 \item Inter-operable data formats 
-\item Planetary data
-\item 
+\item Planetary data standards
+\item Data processing and visualization
 \end{keypoints}
@@
 \label{sec:intro}
-
-The ECMWF ensemble is under continuous development, and hence its
-forecast skill improves over time \citep{Buiz98,Buiz07,Rich13,Haid14}.
-Parts of these improvements may be due to a reduction of probabilistic
-biases. From this we deduce the following hypothesis: As the raw
-forecasts continuously improve, it is hypothesized that the gap in
-skill between raw ensemble and post-processed forecasts narrows,
-because systematic errors typically captured by post-processing are
-reduced 
-by those improvements. 
-%%
-%% Use \explain{} before change, so that it ends up on the right line.
-\explain[Jon, 2/16/16]{Redundant sentence, better without it} 
-\deleted{In other words, probabilistic biases,
-which can be reduced by statistical post-processing methods, decrease
-over time.}
-Assuming that the raw ensemble forecasts continue to
-improve in the future, the gap in skill may eventually be closed when
-the raw ensemble forecasts become reliable and unbiased.  In this work
-we analyse the evolution of the global performance of the operational
-ECMWF raw ensemble and the corresponding post-processed EMOS forecasts
-for 2 metre temperature (T2M), 24-hour precipitation (PPT24), and 10-m
-wind speed (V10). We verify the forecasts against globally distributed
-surface synoptic observations (SYNOP) data over a period of about 10
-years. We firstly evaluate the monthly average skill in terms of CRPS
-for both the raw and the EMOS forecasts. In order to assess the extent
-to which the results depend on the choice of the post-processing
-method, Bayesian model averaging (BMA), \citet{Raft05,Fra10} is
-additionally applied to the T2M raw ensemble forecasts. We will use
-the negatively oriented (i.e. the lower the value the higher the
-skill) continuous ranked probability score (CRPS) \citep{Math76} as a
-measure of skill. As the CRPS assesses both reliability and sharpness
-and is a proper score \citep{Gneit07b}, we rely on it for model fitting
-and verification throughout this study. Note that skill and
-reliability are linked in that given constant sharpness an improvement
-in reliability leads to an improvement in skill and vice versa. We
-finally analyse the evolution of the gap in CRPS between raw ensemble
-and post-processed forecasts.
-
-After presenting the dataset in section~\ref{sec:intro} we summarize
-the methods for post-processing and for the assessment of the global
-skill evolution in section~\ref{sec:meth}. In section~\ref{sec:res}
-the results are shown.  This is followed by a discussion in
-section~\ref{sec:disc} along with some concluding
-remarks. These
-analyses have been performed using the statistical software \texttt{R}
-\citep{R2013}.
 
@@
 
-\section{Methods}
-\label{sec:meth}
+\section{FITS format in Planetary Surface investigation}
+\label{sec:fitspss}
 
-\subsection{Post-processing Using EMOS}
-Post-processing using EMOS converts a raw ensemble of discrete
-forecasts into a probability distribution. Let $y$ be the variable to
-be forecast (here: T2M, PPT24 or V10) and let ${f}=(f_{1}, f_{2},
-\dots, f_{K})^{T}$ be the vector of the $K$ member raw ensemble
-forecasts (here: HRES, ENS, and CTRL). Then the EMOS
-\added{predictive}
-density can be written as:
-\begin{linenomath*}
-\begin{equation}
-y|{f} \sim g(m, \sigma),
-\end{equation}
-\end{linenomath*}
-where $g(\cdot)$ is a parametric density function with location and
-scale parameters $m$ and $\sigma$, respectively, which depend on the
-raw ensemble.
+\subsection{FITS format}
 
-\subsubsection{Temperature}
-For T2M forecasts $g(\cdot)$ is a normal density distribution with
-mean $m$ and variance $\sigma^{2}$. Here, we use a variant of the
-original EMOS approach similar to the one proposed by
-\citet{ScheueBue13} where the departures of observed temperatures from
-their climatological means are related to those of the forecasts.
-Specifically, let $T=\{t_{1},\ldots,t_{n}\}$ be a training period of
-$n$ days preceding the forecast initialization and denote by $f_{tk}$
-the forecast of the k-th ensemble member and by $y_{t}$ the
-observation on day $t\in T$. As a first step, we fit a regression
-model
-%%% \begin{linenomath*} ...\end{linenomath*} puts a line number in equation:
-\begin{linenomath*}
-\begin{equation}
-y_{t_{j}} = c_{0} + c_{1}\sin\left(\frac{2\pi j}{365}\right) +
-c_{2}\cos\left(\frac{2\pi j}{365}\right) + \varepsilon_{t_{j}}, \quad
-j=1,\ldots,n
-\end{equation}
-\end{linenomath*}
-which captures the seasonal variation of T2M. The residual terms
-$\varepsilon_{t_{j}}$ are likely correlated over time, but for
-simplicity an ordinary least squares fit is performed. We denote by
-$\tilde{y}_{t}$ the fitted value of this periodic regression model on
-day $t$ and interpret it as the climatological mean temperature on
-this day. This model can easily be extrapolated to future days
-$t_{d+1}, t_{d+2}, \ldots$ The above regression includes both a sine
-and a cosine term which is equivalent to a cosine model with variable
-phase and amplitude. Since $j=1,\ldots,n$ is just a numbering of the
-days in $T$, different training periods have different phase
-parameters and hence $c_{1}$ and $c_{2}$ evolve over the calendar
-year. We fit the same type of model also to the ensemble mean,
-control, and high resolution run and obtain climatological means
-$\tilde{f}_{\overline{\mathrm{ENS}},t}, \tilde{f}_{\mathrm{CTRL},t}$,
-and $\tilde{f}_{\mathrm{HRES},t}$. The mean of the forecast
-distribution is then:
-\begin{linenomath*}
-\begin{equation}
-m = \tilde{y} + a_{1}(f_{\mathrm{HRES}} - \tilde{f}_{\mathrm{HRES}}) +
-a_{2}(f_{\mathrm{CTRL}} - \tilde{f}_{\mathrm{CTRL}}) +
-a_{3}(f_{\overline{\mathrm{ENS}}} -
-\tilde{f}_{\overline{\mathrm{ENS}}}).
-\end{equation}
-\end{linenomath*}
-The variance of the forecast distribution is linked to the raw
-ensemble by: 
-\begin{linenomath*}
-\begin{equation}
-\sigma^{2} = b_{0} + b_{1}s^{2},
-\end{equation}
-\end{linenomath*}
-where $s^{2} = \frac{1}{K}\sum_{k=1}^{K}(f_{k} -
-\frac{1}{K}\sum_{k=1}^{K}f_{k})^{2}$. The parameters
-${\theta}_{T2M}=(a_{1}, a_{2}, a_{3}, b_{0}, b_{1})^{T}$ are
-constrained to be non-negative, and hence $a_{k}/\sum_{k=1}^{K} a_{k}$
-can be understood as the weight of model $k$.
+\subsubsection{Planetary Surface proposed convention}
 
-\subsubsection{Precipitation}
-For PPT24 we use the EMOS approach proposed by \citet{Scheu14}, where
-$g(\cdot)$ is a left-censored (at zero) generalized extreme value
-(GEV) distribution. While the shape parameter $\xi$ of the GEV is kept
-constant ($\xi=0.2$), the location and the scale parameters $m$ and
-$\sigma$ are linked to the raw ensemble via:
-\begin{linenomath*}
-\begin{eqnarray}
-m &=& a_{0} + a_{1} f_{\mathrm{HRES}} + a_{2} f_{\mathrm{CTRL}} +
-a_{3} f_{\overline{\mathrm{ENS}}} + a_{4} \pi_{0},\nonumber\\
-\sigma &=& b_{0} + b_{1} \mathrm{MD}_{f},
-\end{eqnarray}
-\end{linenomath*}
-where $\pi_{0}$ is the fraction of ensemble members predicting zero
-precipitation and $\mathrm{MD}_{f}:= K^{-2}$\break
-$\sum_{k, k' = 1}^{K} |
-f_{k} - f_{k'} |$ is the ensemble mean difference. Again, the
-parameters are denoted by $\theta_{PPT24}=(a_{0}, \dots, a_{4}, b_{0},
-b_{1})^{T}$. The parameters $a_{1}, a_{2}, a_{3}, b_{0}, b_{1}$ are
-constrained to be non-negative, and hence the normalized parameters
-$a_{1}$ to $a_{3}$ can be understood as weights.
+\subsubsection{FITS to PDS3 and PDS4 dictionaries}
 
-\subsection{Global CRPS Analysis}
- \label{sec:meth_glob_crps}
+\subsubsection{Best Practices and Examples}
 
-As stated in the introduction, the main objective of this study is to
-analyse whether the gap in CRPS between the raw ensemble and the
-post-processed forecast narrows over time. This is assessed
-station-wise using both a parametric and a non-parametric approach.
-For the former, we fit the following regression model to the monthly
-time series of CRPS differences ($\Delta \mathrm{CRPS}_{t} =
-\mathrm{CRPS}_{\mathrm{raw},t} - \mathrm{CRPS}_{\mathrm{EMOS}, t}$):
-\begin{linenomath*}
-\begin{equation}
-\Delta \mathrm{CRPS}_{t} = \beta_{0} + \beta_{1} t + \beta_{2} \sin\left(\frac{2\pi t}{12}\right) + \beta_{3} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2}) \label{eq:linseas}
-\end{equation}
-\end{linenomath*}
-where $\Delta \mathrm{CRPS}_{t}$ is the predictand, $t$ is now the
-time in months, and $\sigma^{2}$ denotes the error variance. For the
-latter, we use Kendall's $\tau$ correlation coefficient and the
-associated test statistics \citep{Mann45} as implemented in the
-\texttt{R} package \texttt{Kendall} \citep{Kendall}. In order to
-correct for seasonal effects, we calculate the $\tau$ statistics using
-the residuals of the following model:
-\begin{linenomath*}
-\begin{equation}
-\Delta \mathrm{CRPS}_{t} = \gamma_{0} + \gamma_{1} \sin\left(\frac{2\pi t}{12}\right) + \gamma_{2} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2})
-\label{eq:tauseas}
-\end{equation}
-\end{linenomath*}
-Note that negative $\tau$ values indicate a negative trend and
-positive values a positive one. Figure~\ref{figone} a) and b)
-show the regression lines estimated by model (\ref{eq:linseas}) for
-monthly averages of $\Delta \mathrm{CRPS}$ and the corresponding
-Kendall's $\tau$ test statistics for an example with decreasing and
-increasing gap.
+\section{Inter-operable software developments}
+\label{sec:softdev}
 
-\section{Results}
-\label{sec:res}
-
-\subsection{Are There Any Significant Temporal Trends?}
-
-The above results indicate a tendency of a decrease in $\Delta$CRPS
-over time at least for T2M and PPT24. In the following we check the
-percentages of stations with decreasing, an absence of, or increasing
-trend in $\Delta$CRPS over time at a significance level of 0.05. In
-order to be more confident about the results this analysis is
-performed using both the parametric regression model and the
-non-parametric Kendall's $\tau$ correlation coefficient test. As
-already mentioned both approaches correct for seasonal effects.
-Furthermore, in case of T2M the same analysis has been performed
-additionally using BMA instead of EMOS in order to relax the
-dependence on one particular post-processing method. As shown in
-Table~\ref{tab:signif_gap_clos} the stations with no significant trend
-outnumber the stations with either negative or positive trend for all
-three variables and lead times considered. Note that the percentage of
-stations without any significant trend increases with increasing lead
-time. In line with the results shown in
-Figure~\ref{fig:mean_and_slope}, significantly negative trends are
-more common than positive ones for T2M and PPT24. The difference
-between the number of stations with negative and those with positive
-trend reduces with increasing lead time, but is still greater than
-zero for a 10 day forecast. Note that the high number of
-non-significant stations in case of PPT24 is likely to be due to the
-high variability of precipitation amounts, and hence variability of
-CRPS values, which leads to a large residual standard error in case of
-the parametric regression model and to a lot of pairs (a pair denotes
-here a value of $\Delta$CRPS and its associated time stamp) opposite
-to the estimated direction in case of the $\tau$ test statistics. In
-case of V10 the stations with a negative trend and those with a
-positive trend are almost equally frequent regardless of the lead
-time. Figures of the global distributions of stations with no,
-significantly negative, and significantly positive trend in
-$\Delta$CRPS are available as supplemental material to this paper.
-
-\section{Discussion and Conclusions}
+\section{Discussion and Perspectives}
 \label{sec:disc}
-
-According to the above analyses the gap in CRPS between the raw
-ensemble and the EMOS forecasts remains almost constant over time. For
-T2M and PPT24 $\Delta$CRPS shows a slightly decreasing tendency. The
-higher the lead time the less accentuated is this tendency. For V10
-such a tendency cannot be detected. The parametric regression model
-and the non-parametric $\tau$ test yield similar results. Hence, a
-linear model that is overlaid by seasonal fluctuations seems to be
-reasonable. Note that the skill of the raw ensemble and the EMOS
-forecasts may sometimes be negatively affected by upgrades to the
-atmospheric model. Model upgrades may deteriorate raw ensemble skill
-at some individual stations. For instance, a resolution increase may
-introduce new issues with statistical downscaling of the forecasts to
-some specific observation sites. But more importantly, the skill of
-the post-pro\-cessed forecasts can be lowered dramatically if a model
-update happens between the training and the verification period. These
-issues may result in positive trends in $\Delta$CRPS. Ideally,
-post-processing would be based on a cascade of reforecasts. That is,
-for each atmospheric model version, training of the post-processing
-model would be done using a corresponding time series of reforecasts
-made with that same model version. Furthermore, the observations may
-be affected by measurement errors. If these errors change over time,
-they may also influence the estimates of the trends in $\Delta$CRPS.
-As the problems introduced by statistical downscaling may be mitigated
-by verifying against model analysis, a similar study that replaces
-observations by model analysis, as proposed by \citet{Ghel00} and
-\citet{Papp09}, may give further insights.
-
-From the above we conclude that the probabilistic skill of both the
-raw ensembles and the EMOS forecasts improves over time. The fact that
-the gap in skill has remained almost \textbf{constant}, especially for
-V10, suggests that improvements to the atmospheric model have an
-effect quite different from what calibration by statistical
-post-processing is doing. That is, they are increasing potential
-skill. Thus this study indicates that (a) further model development is
-important even if one is just interested in point forecasts, and (b)
-statistical post-processing is important because it will keep adding
-skill in the foreseeable future.
 
@@
 \acknowledgments
-The forecast data used in this study can be made available subject to
-a handling charge. SYNOP observations are not available through ECMWF.
-We are grateful to F. Rabier, D. S. Richardson, D. Lavers, and other
-colleagues at ECMWF for helpful discussions and inputs. We thank David
-M\"{o}ller of the University of Heidelberg for his contribution.
-Furthermore, we like to thank T. Gneiting of the Heidelberg Institute
-for Theoretical Studies and of the Institute for Stochastics at
-Karlsruhe Institute of Technology for valuable comments, and the
-colleagues at the Heidelberg Institute for Theoretical Studies.
-Additionally, we are grateful to two anonymous reviewers for their
-helpful comments. We are indebted to the ECMWF for granting access to
-their products.
-
-The Editor thanks Pierre Pinson and an anonymous reviewer for their assistance in evaluating this paper.
+This work benefits from support of VESPA/Europlanet.
+The Europlanet 2020 Research Infrastructure is funded by the European Union under the
+Horizon 2020 research and innovation program, grant agreement N.654208.
 
@@
 
-\bibliography{agusample}
+\bibliography{epss_marmo}
 

### 2018-01-15 epss_marmo.tex
Update on Overleaf.

@@
  \authors{C. Marmo\affil{1},
- T. M. Hare\affil{2}, ...}
+ T. M. Hare\affil{2},
+ S. Erard\affil{3},
+ B. Cecconi\affil{3},
+ M. Minin\affil{4},
+ A. P. Rossi\affil{4} ...}
 
@@
 \affiliation{2}{U. S. Geological Survey, Astrogeology Science Center, Flagstaff, AZ, USA}
-%\affiliation{3}{}
-%\affiliation{4}{}
+\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University, CNRS, Sorbonne Universités, UPMC Univ. Paris 06, Univ. Paris Diderot, Sorbonne Paris Cité, Meudon, France}
+\affiliation{4}{Jacobs University, Bremen, Germany}
 %\affiliation{5}{}
@@
 \begin{abstract}
-
+Planetary science is a vast field of investigation that brings together several
+research communities (geologists, astronomers, physicists, geochemists, etc.),
+and produces an impressively growing amount of heterogeneous data.
+Planetary surface research continues to evolve from mostly visual assessment
+to more automated quantitative analysis.
+Interoperability and openness of data formats and processing techniques are
+becoming a necessity, to avoid the risk of being unable to efficiently extract
+scientific information from the data, and to guarantee the reproducibility of
+the scientific results.
+This paper will describe how Flexible Image Transport System (FITS) can be used
+in planetary surface investigations, and how its metadata can easily be inserted
+in the PDS 4 metadata distribution model.
+FITS format is open and flexible, largely implemented in open and efficient
+processing tools.
+Those reasons make FITS format appropriate for large amounts of raster data
+processing.
+The purpose of this approach is to make easier data mining and re-processing in
+Planetary Surface investigations, promoting general software based on those
+standards.
+More than capturing a formal data model, FITS description for those data aims to
+simplify operational logistics.
 \end{abstract}
@@
 \label{sec:intro}
+Planetary science is a vast field of investigation that brings together several
+research communities (geologists, astronomers, physicists, geochemists, etc.),
+and produces an impressively growing amount of heterogeneous data.
+Planetary surface research continues to evolve from mostly visual assessment
+to more automated quantitative analysis.
+Interoperability and openness of data formats and processing techniques are
+becoming a necessity, to avoid the risk of being unable to efficiently extract
+scientific information from the data, and to guarantee the reproducibility of
+the scientific results.
+This paper will describe how Flexible Image Transport System (FITS) can be used
+in planetary surface investigations, and how its metadata can easily be inserted
+in the PDS 4 metadata distribution model.
+FITS format is open and flexible, largely implemented in open and efficient
+processing tools.
+Those reasons make FITS format appropriate for large amounts of raster data
+processing.
+The purpose of this approach is to make easier data mining and re-processing in
+Planetary Surface investigations, promoting general software based on those
+standards.
+More than capturing a formal data model, FITS description for those data aims to
+simplify operational logistics.
+
+The option to use FITS within the
+planetary domain could be an opportunity to allow
+more seamless sharing of data across these different
+domains and potentially homogenize methods from
+acquisition, to visualization, while giving more chances
+to optimize data processing
+
+This paper is organized as follows:
+\begin{enumerate}
+\item{a brief description of FITS format and on its pros and cons in using it in planetary science;}
+\item{a proposition for planetary surface metadata description in FITS}
+\item{Dictionaries between FITS and PDS metadata description}
+\item{examples of planetary dataset in FITS}
+\item{current developments for making FITS usable in planetary sciences}
+\end{enumerate}
+
+\section{FITS format in Planetary Surface investigation}
+\label{sec:fitspss}
+
+\subsection{FITS format}
+FITS\citep{fitsorig,fitsver3,fitsoffice} stands for 'Flexible Image Transport System'.
+It is an open digital standard, defined for data acquisition and archiving in astronomical
+observatories back in the late 70's, and it is used for spatial telescope data.
+In fact, FITS is the IAU-approved standard format for astronomical
+data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
+Therefore, FITS is one of the standard implemented formats in the Virtual
+Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html} (VO).
+Finally, FITS format is compatible with the Planetary Data System (PDS)
+3\footnote{https://pds.nasa.gov/pds4/about/} and
+4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving specifications and
+supported by a large number of open libraries and software tools.
+
+FITS is already able to propose standard formatting for some data products quite common by now in
+planetary surface investigations.
+In particular, Multi-Extension FITS (MEF) schema proposes an easy way
+to store inhomogeneous digital information (reflectance, calibration data, vector table data, etc.)
+in the same file each with relative metadata, as well as
+multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
+(e.g. from CRISM \citep{crism} or OMEGA \citep{omega}).
+FITS has been already chosen to distribute, e.g.,
+Hayabusa [10] and some of the Dawn [11] data.
+To be efficiently used in planetary surface investigations, FITS metadata must be extended in order to
+take into account the size of the reference body.
+planetary body shapes and orientations are well defined
+and standardized by the International Astronomical
+Union (IAU) Working Group on Cartographic
+Coordinates and Rotational Elements (WGCCRE).
+This group reports triennially on the preferred rotation
+rate, spin axis, prime meridian, and reference surface
+for planets and satellites which helps ensure that
+cartographic endeavors are effectively comparable [8].
+In the framework of the VESPA [12, 13] component of the
+Europlanet 2020 project an extension to FITS meta-
+data (GeoFITS) has been proposed [14].
+
+\subsubsection{Planetary Surface proposed convention}
+
+\subsubsection{FITS to PDS3 and PDS4 dictionaries}
+
+\subsubsection{Best Practices and Examples}
+
+\section{Inter-operable software developments}
+\label{sec:softdev}
+
+\section{Discussion and Perspectives}
+\label{sec:disc}
 
@@
 
-\begin{figure}[ht!]
+%\begin{figure}[ht!]
 % \centering
@@
 % \includegraphics will insert the right file:
-\centerline{\includegraphics[height=1.5in]{figsamp}}
-\caption{Short caption}
-\label{figone}
-\end{figure}
+%\centerline{\includegraphics[height=1.5in]{figsamp}}
+%\caption{Short caption}
+%\label{figone}
+%\end{figure}
 
-\begin{figure}[ht!]
-\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
-\caption{The figure caption should begin with an overall descriptive
-statement of the figure followed by additional text. They should be
-immediately after each figure.  Figure parts are indicated with
-lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
-both the figures and captions in the text near where they are cited.}
-\label{fig:mean_and_slope}
-\end{figure}
+%\begin{figure}[ht!]
+%\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
+%\caption{The figure caption should begin with an overall descriptive
+%statement of the figure followed by additional text. They should be
+%immediately after each figure.  Figure parts are indicated with
+%lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
+%both the figures and captions in the text near where they are cited.}
+%\label{fig:mean_and_slope}
+%\end{figure}
 
-\begin{table}
-\caption{Start this caption with a short description of your
-table.
-Large tables
-especially presenting rich data should be presented as separate excel
-or .cvs files, not as part of the main text.}
-== Table Here ==
-\end{table}
+%\begin{table}
+%\caption{Start this caption with a short description of your
+%table.
+%Large tables
+%especially presenting rich data should be presented as separate excel
+%or .cvs files, not as part of the main text.}
+%== Table Here ==
+%\end{table}
 
@@
 %
-\begin{table}
-\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-\centering
-\begin{tabular}{l c}
-\hline
- Run  & Time (min)  \\
-\hline
-  $l1$  & 260   \\
-  $l2$  & 300   \\
-  $l3$  & 340   \\
-  $h1$  & 270   \\
-  $h2$  & 250   \\
-  $h3$  & 380   \\
-  $r1$  & 370   \\
-  $r2$  & 390   \\
-\hline
-\multicolumn{2}{l}{$^{a}$Table note text here.}
-\end{tabular}
-\end{table}
+%\begin{table}
+%\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
+%\centering
+%\begin{tabular}{l c}
+%\hline
+% Run  & Time (min)  \\
+%\hline
+%  $l1$  & 260   \\
+%  $l2$  & 300   \\
+%  $l3$  & 340   \\
+%  $h1$  & 270   \\
+%  $h2$  & 250   \\
+%  $h3$  & 380   \\
+%  $r1$  & 370   \\
+%  $r2$  & 390   \\
+%\hline
+%\multicolumn{2}{l}{$^{a}$Table note text here.}
+%\end{tabular}
+%\end{table}
 
@@
 %
- \begin{sidewaysfigure}
- \includegraphics[width=20pc]{figsamp}
- \caption{Caption here}
- \label{newfig}
- \end{sidewaysfigure}
+% \begin{sidewaysfigure}
+% \includegraphics[width=20pc]{figsamp}
+% \caption{Caption here}
+% \label{newfig}
+% \end{sidewaysfigure}
 %
- \begin{sidewaystable}
- \caption{Caption here}
-\label{tab:signif_gap_clos}
- \begin{tabular}{ccc}
-one&two&three\\
-four&five&six
- \end{tabular}
- \end{sidewaystable}
+% \begin{sidewaystable}
+% \caption{Caption here}
+%\label{tab:signif_gap_clos}
+% \begin{tabular}{ccc}
+%one&two&three\\
+%four&five&six
+% \end{tabular}
+% \end{sidewaystable}
 
-\clearpage
-
-\section{FITS format in Planetary Surface investigation}
-\label{sec:fitspss}
-
-\subsection{FITS format}
-
-\subsubsection{Planetary Surface proposed convention}
-
-\subsubsection{FITS to PDS3 and PDS4 dictionaries}
-
-\subsubsection{Best Practices and Examples}
-
-\section{Inter-operable software developments}
-\label{sec:softdev}
-
-\section{Discussion and Perspectives}
-\label{sec:disc}
+%\clearpage
 
 %% ------------------------------------------------------------------------ %%
-\section*{Citations}
+%\section*{Citations}
 
@@
 
-\subsection*{Cites made with \tt\string\citet\string{\string}}
-...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
-\citet{Ghel00}, and \citet{Leit74}.
+%\subsection*{Cites made with \tt\string\citet\string{\string}}
+%...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
+%\citet{Ghel00}, and \citet{Leit74}.
 
-\subsection*{Cites made with \tt\string\citep\string{\string}}
-...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
-\citep{Ghel00, Leit74}.
+%\subsection*{Cites made with \tt\string\citep\string{\string}}
+%...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
+%\citep{Ghel00, Leit74}.
 
-...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.
+%...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.
 
@@
 %
-\appendix
-\section{Here is a sample appendix}
-This is an Appendix section.
+%\appendix
+%\section{Here is a sample appendix}
+%This is an Appendix section.
 
-\subsection{subsection}
-This is an Appendix subsection.
+%\subsection{subsection}
+%This is an Appendix subsection.
 
-\subsubsection{subsubsection}
-This is an Appendix subsubsection.
-\begin{linenomath*}
-\begin{equation}asdf\end{equation}
-\end{linenomath*}
+%\subsubsection{subsubsection}
+%This is an Appendix subsubsection.
+%\begin{linenomath*}
+%\begin{equation}asdf\end{equation}
+%\end{linenomath*}
 
@@
 % Glossary is only allowed in Reviews of Geophysics
-\begin{glossary}
-\term{Term}
- Term Definition here
-\term{Term}
- Term Definition here
-\term{Term}
- Term Definition here
-\end{glossary}
+%\begin{glossary}
+%\term{Term}
+% Term Definition here
+%\term{Term}
+% Term Definition here
+%\term{Term}
+% Term Definition here
+%\end{glossary}
 
@@
 % Acronyms
- \begin{acronyms}
- \acro{Acronym}
- Definition here
- \acro{EMOS}
- Ensemble model output statistics 
- \acro{ECMWF}
- Centre for Medium-Range Weather Forecasts
- \end{acronyms}
+% \begin{acronyms}
+% \acro{Acronym}
+% Definition here
+% \acro{EMOS}
+% Ensemble model output statistics 
+% \acro{ECMWF}
+% Centre for Medium-Range Weather Forecasts
+% \end{acronyms}
 
@@
 % Notation 
- \begin{notation}
- \notation{$a+b$} Notation Definition here
- \notation{$e=mc^2$} 
- Equation in German-born physicist Albert Einstein's theory of special
-relativity that showed that the increased relativistic mass ($m$) of a
-body comes from the energy of motion of the body—that is, its kinetic
-energy ($E$)—divided by the speed of light squared ($c^2$).
- \end{notation}
+% \begin{notation}
+% \notation{$a+b$} Notation Definition here
+% \notation{$e=mc^2$} 
+% Equation in German-born physicist Albert Einstein's theory of special
+%relativity that showed that the increased relativistic mass ($m$) of a
+%body comes from the energy of motion of the body—that is, its kinetic
+%energy ($E$)—divided by the speed of light squared ($c^2$).
+% \end{notation}
 
@@
 This work benefits from support of VESPA/Europlanet.
-The Europlanet 2020 Research Infrastructure is funded by the European Union under the
-Horizon 2020 research and innovation program, grant agreement N.654208.
+The Europlanet 2020 Research Infrastructure is funded by the European Union
+under the Horizon 2020 research and innovation program, grant agreement N.654208.
 
@@
 
-% % \begin{thebibliography}{}
+%\begin{thebibliography}{}
 
@@
 
-% \end{thebibliography}
+%\end{thebibliography}
 

### 2018-01-17 epss_marmo.tex

@@
 \begin{enumerate}
-\item{a brief description of FITS format and on its pros and cons in using it in planetary science;}
+\item{a brief description of FITS format and on its pros and cons in using it in
+planetary science;}
 \item{a proposition for planetary surface metadata description in FITS}
@@
 3\footnote{https://pds.nasa.gov/pds4/about/} and
-4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving specifications and
-supported by a large number of open libraries and software tools.
+4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving
+specifications and supported by a large number of open libraries and software tools.
 
-FITS is already able to propose standard formatting for some data products quite common by now in
-planetary surface investigations.
+FITS is already able to propose standard formatting for some data products quite
+common by now in planetary surface investigations.
 In particular, Multi-Extension FITS (MEF) schema proposes an easy way
-to store inhomogeneous digital information (reflectance, calibration data, vector table data, etc.)
-in the same file each with relative metadata, as well as
+to store inhomogeneous digital information (reflectance, calibration data, vector
+table data, etc.) in the same file each with relative metadata, as well as
 multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
@@
 FITS has been already chosen to distribute, e.g.,
-Hayabusa [10] and some of the Dawn [11] data.
-To be efficiently used in planetary surface investigations, FITS metadata must be extended in order to
-take into account the size of the reference body.
-planetary body shapes and orientations are well defined
-and standardized by the International Astronomical
-Union (IAU) Working Group on Cartographic
+Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
+Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
+and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
+To be efficiently used in planetary surface investigations, FITS metadata must
+be extended in order to take into account the size of the reference body.
+Planetary body shapes and orientations are well defined and standardized by the
+International Astronomical Union (IAU) Working Group on Cartographic
 Coordinates and Rotational Elements (WGCCRE).
-This group reports triennially on the preferred rotation
-rate, spin axis, prime meridian, and reference surface
-for planets and satellites which helps ensure that
-cartographic endeavors are effectively comparable [8].
+This group reports triennially on the preferred rotation rate, spin axis, prime meridian,
+and reference surface for planets and satellites which helps ensure that
+cartographic endeavors are effectively comparable \citep{wgccre}.
 In the framework of the VESPA [12, 13] component of the

### 2018-01-24 epss_marmo.tex

@@
 \subsection{FITS format}
-FITS\citep{fitsorig,fitsver3,fitsoffice} stands for 'Flexible Image Transport System'.
+FITS\footnote{https://fits.gsfc.nasa.gov/} \citep{fitsorig,fitsver3} stands for
+'Flexible Image Transport System'.
 It is an open digital standard, defined for data acquisition and archiving in astronomical
@@
 cartographic endeavors are effectively comparable \citep{wgccre}.
-In the framework of the VESPA [12, 13] component of the
-Europlanet 2020 project an extension to FITS meta-
-data (GeoFITS) has been proposed [14].
+In the framework of the VESPA\footnote{http://europlanet-vespa.eu/} \citep{vespapss}
+component of the EuroplanetRI2020 project an extension to FITS meta-
+data (GeoFITS) has been proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention}.
 
 \subsubsection{Planetary Surface proposed convention}
+When willing to use FITS format for Planetary Surfaces the issue of rigorous
+geographical representation must be addressed.
+Planetary surface researchers for their mapping and analysis needs, commonly use
+Geographical Information Systems (GIS).
+In general, GIS applications excel in data interoperability even
+though some have historically been anchored to Earth's spatial description.
+FITS World Coordinate System representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html.}
+simplifies the spatial coordinate description with respect to historical terrestrial
+references.
+There is no need for oblique projection definitions \citep{Snyder1987,calagreisenI}
+as projection parameters (projection center or reference point)
+describe the difference between the Mercator and
+Transverse Mercator or Simple Cylindrical and
+Equirectangular map projections.
+There is no need for East or West positive longitude definitions as the pixel
+conversion in world coordinates is set by a common
+oriented matrix. 
+However, if still needed, FITS does offer the possibility to fully describe those
+same parameters using alternative WCS definitions (meters, degrees,
+East and West longitude, etc.)\citep{greisencalaII}.
+Nevertheless, WCS does not record information about the body shape and
+orientation, and reference surface.
 

### 2018-01-24 epss_marmo.tex
Update on Overleaf.

@@
 oriented matrix.
+
+\begin{table}
+\label{table:proj}
+\caption{Projection names and codes already available in FITS suitable for planetary applications:
+their translation in PDS vocabulary is also provided}
+\centering
+\begin{tabular}{l l l}
+\hline
+FITS Name$^{a}$ & FITS code$^{a}$ & PDS name \\ 
+\hline
+Sanson-Flamsteed & SFL & \\
+Zenithal Equal-Area & ZEA & \\
+Conic ortomorphic & COO & \\
+Plate Carre & CAR & \\
+generalization of Lambert Equal-Area & CEA \\
+Mercator & MER & \\
+Ortographic & SIN & \\
+Zenithal perspective & AZP & \\
+Stereographic & STG & \\
+\hline
+\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}}
+\end{tabular}
+\end{table}
+
+\begin{table}
+\label{table:ctype}
+\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.}
+\centering
+\begin{tabular}{l l}
+\hline
+Body & CTYPE$^{a}$ \\ 
+\hline
+Moon & SE \\
+Venus & VE \\
+Mars & MA \\
+Jupiter & JU \\
+Saturn & SA \\
+Uranus & UR \\
+Neptune & NE \\
+Planetary bodies & B \\
+\hline
+\multicolumn{2}{l}{$^{a}$From \citet{calagreisenI}}
+\end{tabular}
+\end{table}
+
 However, if still needed, FITS does offer the possibility to fully describe those
@@
 orientation, and reference surface.
+
+\begin{sidewaystable}
+\label{table:newkeys}
+\caption{Proposed keywords are related to 3D surface shape,
+not available in astronomical celestial standards.}
+\centering
+\begin{tabular}{l l l l}
+\hline
+Keyword & Type & Status & Definition  \\
+\hline
+WGCCRECS & string & reserved & Value field contains a string referring to the WGCCRE report \\
+& & & or document defining the coordinate system describing the data. (DOI)    \\
+A\_RADIUS$^{a}$ & real & mandatory & Value field contains the semi-major axis of the ellipsoid that \\
+& & & defines the approximate shape of a target body used in projection. \\
+& & & 'A' is usually in the equatorial plane. Always in meters.  \\
+B\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the intermediate axis of \\
+& & & the ellipsoid that defines the approximate shape of a target body used in projection. \\
+& & & 'B' is usually in the equatorial plane. Always in meters.   \\
+C\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the semi-minor axis of the \\
+& & & ellipsoid that defines the approximate shape of a target body used in projection. \\
+& & & 'C' is normal to the plane defined by 'A' and 'B'. Always in meters. \\
+OGCCODE	& string & reserved, optional & Value field contains a string describing the standard OGC code \\
+& & & (if any) from \citet{hare2006} corresponding to the shape and projection used in the image. \\
+& & & This keyword will assure compatibility with standard GIS softwares. \\
+\hline
+\multicolumn{4}{l}{$^{a}$Note that IAU Working Group defines a mean radius for each Solar System body. When this Mean Radius is used in projection definition,} \\
+\multicolumn{4}{l}{  the three keywords A\_RADIUS, B\_RADIUS,C\_RADIUS, must be set to the same value of the defined Mean Radius.} \\
+\end{tabular}
+\end{sidewaystable}
 

### 2018-01-24 epss_marmo.tex

@@
 \hline
-FITS Name$^{a}$ & FITS code$^{a}$ & PDS name \\ 
+FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
 \hline
-Sanson-Flamsteed & SFL & \\
-Zenithal Equal-Area & ZEA & \\
-Conic ortomorphic & COO & \\
-Plate Carre & CAR & \\
-generalization of Lambert Equal-Area & CEA \\
-Mercator & MER & \\
-Ortographic & SIN & \\
+Sanson-Flamsteed & SFL & Sinusoidal \\
+Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
+Conic ortomorphic & COO & Lambert Conformal Conic\\
+Plate Carre & CAR & Equirectangular \\
+Cylindrical Equal-Area & CEA & \\
+Mercator & MER & Mercator, Oblique Mercator \\
+Ortographic & SIN & Orthographic \\
 Zenithal perspective & AZP & \\
@@
 \hline
-\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}}
+\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
+\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
 \end{tabular}

### 2018-01-25 epss_marmo.tex

@@
 
-\subsubsection{Planetary Surface proposed convention}
+\section{Planetary Surface proposed convention}
 When willing to use FITS format for Planetary Surfaces the issue of rigorous
@@
 describe the difference between the Mercator and
-Transverse Mercator or Simple Cylindrical and
+Oblique Mercator or Simple Cylindrical and
 Equirectangular map projections.
@@
 oriented matrix.
+
+\begin{table}
+\label{table:wcskeys}
+\caption{WCS FITS keywords in the Planetary Surface context}
+\centering
+\begin{tabular}{l l l l}
+\hline
+Keyword & Type & Definition & PDS4 \\
+\hline
+EQUINOX & real & & \\
+RADESYS	& string & & \\
+CRPIXn & real & & \\
+CTYPEn & string & & \\
+CRVALn & real & & \\
+CDi\_j & real matrix & & \\
+PCi\_j & real matrix & & \\
+CDELTn & real & & \\
+CUNITn & string & & \\
+WCSNAME & string & & \\
+CNAMEn & string (up to 68 characters) & & \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
-Conic ortomorphic & COO & Lambert Conformal Conic\\
+Conic Ortomorphic & COO & Lambert Conformal Conic\\
 Plate Carre & CAR & Equirectangular \\
-Cylindrical Equal-Area & CEA & \\
+Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
 Mercator & MER & Mercator, Oblique Mercator \\
 Ortographic & SIN & Orthographic \\
-Zenithal perspective & AZP & \\
-Stereographic & STG & \\
+Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
+Stereographic & STG & Stereographic, Polar Stereographic \\
+Gnomonic & TAN & Gnomonic \\
 \hline
@@
 
-\subsubsection{Best Practices and Examples}
+\subsection{Best Practices and Examples}
 

### 2018-02-01 epss_marmo.tex
Update on Overleaf.

@@
 
-\title{FITS format for planetary surfaces: definitions, applications and best practices.}
+\title{FITS format for planetary surfaces: definitions,
+applications and best practices.}
 
@@
 
-\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay, Orsay, France}
-\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center, Flagstaff, AZ, USA}
-\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University, CNRS, Sorbonne Universités, UPMC Univ. Paris 06, Univ. Paris Diderot, Sorbonne Paris Cité, Meudon, France}
+\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay,
+Orsay, France}
+\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center,
+Flagstaff, AZ, USA}
+\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
+CNRS, Sorbonne Universités, UPMC Univ. Paris 06, Univ. Paris Diderot,
+Sorbonne Paris Cité, Meudon, France}
 \affiliation{4}{Jacobs University, Bremen, Germany}
@@
 \begin{abstract}
-Planetary science is a vast field of investigation that brings together several
-research communities (geologists, astronomers, physicists, geochemists, etc.),
-and produces an impressively growing amount of heterogeneous data.
-Planetary surface research continues to evolve from mostly visual assessment
-to more automated quantitative analysis.
+Planetary science encompasses a broad number of research fields and
+brings together several research communities (geologists, astronomers,
+physicists, geochemists, etc.).
+Planetary missions produce an impressively growing amount of heterogeneous
+data which mining evolves those days from mostly visual assessment to more
+automated quantitative analysis.
 Interoperability and openness of data formats and processing techniques are
@@
 the scientific results.
-This paper will describe how Flexible Image Transport System (FITS) can be used
-in planetary surface investigations, and how its metadata can easily be inserted
-in the PDS 4 metadata distribution model.
-FITS format is open and flexible, largely implemented in open and efficient
-processing tools.
-Those reasons make FITS format appropriate for large amounts of raster data
-processing.
-The purpose of this approach is to make easier data mining and re-processing in
-Planetary Surface investigations, promoting general software based on those
-standards.
-More than capturing a formal data model, FITS description for those data aims to
-simplify operational logistics.
+This paper will describe how Flexible Image Transport System (FITS) can be
+used in planetary surface investigations, and how its metadata can easily be
+inserted in the PDS4 metadata distribution model.
+An evolution to the actual standard is proposed to adapt FITS to
+planetary body description.
+Some examples and implementations are then presented.
+More than imposing a formal data model, FITS description for those data aims
+to simplify the processing chain in planetary science, promoting
+interoperability from data acquisition to final visualization.
 \end{abstract}
@@
 \label{sec:intro}
-Planetary science is a vast field of investigation that brings together several
-research communities (geologists, astronomers, physicists, geochemists, etc.),
-and produces an impressively growing amount of heterogeneous data.
-Planetary surface research continues to evolve from mostly visual assessment
-to more automated quantitative analysis.
-Interoperability and openness of data formats and processing techniques are
-becoming a necessity, to avoid the risk of being unable to efficiently extract
-scientific information from the data, and to guarantee the reproducibility of
-the scientific results.
-This paper will describe how Flexible Image Transport System (FITS) can be used
-in planetary surface investigations, and how its metadata can easily be inserted
-in the PDS 4 metadata distribution model.
+Planetary science brings together several research communities
+(geologists, astronomers, physicists, geochemists, etc.),
+and planetary missions produce a growing amount of heterogeneous data.
+To efficiently extract scientific information from those data, they must
+be made integrable, thus inter-operable.
+Openness of data format and processing techniques is becoming a necessity,
+to guarantee the reproducibility of the results obtained from those data.
+
+This paper proposes to use Flexible Image Transport System\footnote{https://fits.gsfc.nasa.gov/}
+(FITS) \citep{fitsorig,fitsver3} in planetary surface investigations,
+and describes how FITS metadata can easily be inserted in the
+PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
+metadata distribution model.
 FITS format is open and flexible, largely implemented in open and efficient
@@
 processing.
-The purpose of this approach is to make easier data mining and re-processing in
-Planetary Surface investigations, promoting general software based on those
+The goal of this approach is to make easier data mining and re-processing
+in planetary surface investigations, promoting general software based on those
 standards.
-More than capturing a formal data model, FITS description for those data aims to
-simplify operational logistics.
-
-The option to use FITS within the
-planetary domain could be an opportunity to allow
-more seamless sharing of data across these different
-domains and potentially homogenize methods from
-acquisition, to visualization, while giving more chances
-to optimize data processing
+The option to use FITS within the planetary domain could potentially
+homogenize methods from acquisition, to visualization, while
+giving chances to optimize data processing.
 
@@
 \begin{enumerate}
-\item{a brief description of FITS format and on its pros and cons in using it in
-planetary science;}
-\item{a proposition for planetary surface metadata description in FITS}
-\item{Dictionaries between FITS and PDS metadata description}
-\item{examples of planetary dataset in FITS}
-\item{current developments for making FITS usable in planetary sciences}
+\item{a brief description of FITS format;}
+\item{a proposition for planetary surface metadata description in FITS;}
+\item{dictionaries between FITS and PDS metadata description;}
+\item{some examples of planetary dataset in FITS;}
+\item{current developments for making FITS usable in planetary sciences.}
 \end{enumerate}
@@
 \subsection{FITS format}
-FITS\footnote{https://fits.gsfc.nasa.gov/} \citep{fitsorig,fitsver3} stands for
-'Flexible Image Transport System'.
-It is an open digital standard, defined for data acquisition and archiving in astronomical
-observatories back in the late 70's, and it is used for spatial telescope data.
+FITS stands for 'Flexible Image Transport System'.
+It is an open digital standard, defined for data acquisition and archiving
+in astronomical observatories back in the late 70's, and it is used for
+spatial telescope data.
 In fact, FITS is the IAU-approved standard format for astronomical
@@
 Therefore, FITS is one of the standard implemented formats in the Virtual
-Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html} (VO).
+Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
+(VO).
 Finally, FITS format is compatible with the Planetary Data System (PDS)
-3\footnote{https://pds.nasa.gov/pds4/about/} and
-4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml} archiving
-specifications and supported by a large number of open libraries and software tools.
+archiving specifications and supported by a large number of open libraries
+and software tools.
 

### 2018-02-01 epss_marmo.tex

@@
 To efficiently extract scientific information from those data, they must
-be made integrable, thus inter-operable.
+be made integrable, thus interoperable.
 Openness of data format and processing techniques is becoming a necessity,
@@
 
-This paper proposes to use Flexible Image Transport System\footnote{https://fits.gsfc.nasa.gov/}
-(FITS) \citep{fitsorig,fitsver3} in planetary surface investigations,
+This paper proposes to use Flexible Image Transport
+System\footnote{https://fits.gsfc.nasa.gov/}
+(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
 and describes how FITS metadata can easily be inserted in the
@@
 processing.
-The goal of this approach is to make easier data mining and re-processing
+The goal of this approach is to make easier data mining and reprocessing
 in planetary surface investigations, promoting general software based on those
@@
 \item{a brief description of FITS format;}
-\item{a proposition for planetary surface metadata description in FITS;}
-\item{dictionaries between FITS and PDS metadata description;}
+\item{a proposition for planetary surface meta-data description in FITS;}
+\item{dictionaries between FITS and PDS meta-data description;}
 \item{some examples of planetary dataset in FITS;}
@@
 \subsection{FITS format}
-FITS stands for 'Flexible Image Transport System'.
-It is an open digital standard, defined for data acquisition and archiving
-in astronomical observatories back in the late 70's, and it is used for
+FITS is an open digital standard, defined for data acquisition and archiving
+in astronomical observatories in the late 70's, and it is used for
 spatial telescope data.
-In fact, FITS is the IAU-approved standard format for astronomical
-data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
+The International Astronomical Union approved FITS as the standard format
+for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
 Therefore, FITS is one of the standard implemented formats in the Virtual
@@
 
-FITS is already able to propose standard formatting for some data products quite
-common by now in planetary surface investigations.
-In particular, Multi-Extension FITS (MEF) schema proposes an easy way
-to store inhomogeneous digital information (reflectance, calibration data, vector
-table data, etc.) in the same file each with relative metadata, as well as
-multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
-(e.g. from CRISM \citep{crism} or OMEGA \citep{omega}).
+FITS is already able to propose standard formatting for some data products
+quite common by now in planetary surface investigations.
+In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
+to store inhomogeneous digital information (reflectance, calibration data,
+vector, table data, etc.) in the same file, each with relative metadata,
+as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
+or hyperspectral cubes with geometry information (e.g. from CRISM
+\citep{crism} or OMEGA \citep{omega} instruments).
 FITS has been already chosen to distribute, e.g.,
@@
 be extended in order to take into account the size of the reference body.
-Planetary body shapes and orientations are well defined and standardized by the
-International Astronomical Union (IAU) Working Group on Cartographic
-Coordinates and Rotational Elements (WGCCRE).
-This group reports triennially on the preferred rotation rate, spin axis, prime meridian,
-and reference surface for planets and satellites which helps ensure that
-cartographic endeavors are effectively comparable \citep{wgccre}.
-In the framework of the VESPA\footnote{http://europlanet-vespa.eu/} \citep{vespapss}
-component of the EuroplanetRI2020 project an extension to FITS meta-
-data (GeoFITS) has been proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention}.
+Planetary body shapes and orientations are well defined and standardized by
+the IAU Working Group on Cartographic Coordinates and Rotational Elements
+(WGCCRE).
+This group reports triennially \citep{wgccre} on the preferred rotation rate,
+spin axis, prime meridian, and reference surface for planets and satellites
+ensuring that cartographic endeavors are effectively comparable.
+In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
+\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
+project an extension to FITS metadata (GeoFITS) has been
+proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
+taking into account the WGCCRE recommendations.
 
-\section{Planetary Surface proposed convention}
-When willing to use FITS format for Planetary Surfaces the issue of rigorous
+\subsection{Planetary Surface proposed convention}
+\label{ssec:geofits}
+When willing to use FITS format for planetary surfaces the issue of rigorous
 geographical representation must be addressed.
-Planetary surface researchers for their mapping and analysis needs, commonly use
+For their mapping and analysis needs planetary researchers commonly use
 Geographical Information Systems (GIS).
@@
 though some have historically been anchored to Earth's spatial description.
-FITS World Coordinate System representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html.}
-simplifies the spatial coordinate description with respect to historical terrestrial
-references.
-There is no need for oblique projection definitions \citep{Snyder1987,calagreisenI}
+FITS World Coordinate System
+representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html}
+simplifies the spatial coordinate description with respect to historical
+terrestrial references.
+There is no need for oblique projection definitions \citep{calagreisenI,Snyder1987}
 as projection parameters (projection center or reference point)
-describe the difference between the Mercator and
+describe the difference between,e.g., the Mercator and
 Oblique Mercator or Simple Cylindrical and
@@
 There is no need for East or West positive longitude definitions as the pixel
-conversion in world coordinates is set by a common
-oriented matrix.
+conversion in world coordinates is set by an oriented matrix.
+
+\begin{table}
+\label{table:proj}
+\caption{Projection names and codes already available in FITS suitable for
+planetary applications: their translation in PDS4 vocabulary is
+provided. See Table A.1. in \citet{calagreisenI} for details.}
+\centering
+\begin{tabular}{l l l}
+\hline
+FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
+\hline
+Sanson-Flamsteed & SFL & Sinusoidal \\
+Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
+Conic Ortomorphic & COO & Lambert Conformal Conic\\
+Plate Carre & CAR & Equirectangular \\
+Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
+Mercator & MER & Mercator, Oblique Mercator \\
+Ortographic & SIN & Orthographic \\
+Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
+Stereographic & STG & Stereographic, Polar Stereographic \\
+Gnomonic & TAN & Gnomonic \\
+\hline
+\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
+\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
+\end{tabular}
+\end{table}
 
@@
 \hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-\label{table:proj}
-\caption{Projection names and codes already available in FITS suitable for planetary applications:
-their translation in PDS vocabulary is also provided}
-\centering
-\begin{tabular}{l l l}
-\hline
-FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
-\hline
-Sanson-Flamsteed & SFL & Sinusoidal \\
-Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
-Conic Ortomorphic & COO & Lambert Conformal Conic\\
-Plate Carre & CAR & Equirectangular \\
-Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
-Mercator & MER & Mercator, Oblique Mercator \\
-Ortographic & SIN & Orthographic \\
-Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
-Stereographic & STG & Stereographic, Polar Stereographic \\
-Gnomonic & TAN & Gnomonic \\
-\hline
-\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
-\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
 \end{tabular}

### 2018-02-01 epss_marmo.tex

@@
 metadata distribution model.
+FITS is one of the standard implemented formats in the Virtual
+Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
+(VO), therefore making obvious its connection to the Planetary Virtual
+Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
+\citep{ERARDepntap}.
 FITS format is open and flexible, largely implemented in open and efficient
@@
 \item{a proposition for planetary surface meta-data description in FITS;}
-\item{dictionaries between FITS and PDS meta-data description;}
+\item{dictionaries between FITS, PDS meta-data description and EPN-TAP2.0
+parameters;}
 \item{some examples of planetary dataset in FITS;}
@@
 terrestrial references.
-There is no need for oblique projection definitions \citep{calagreisenI,Snyder1987}
-as projection parameters (projection center or reference point)
+There is no need for oblique projection definitions
+\citep{calagreisenI,Snyder1987} as projection parameters
+(projection center or reference point)
 describe the difference between,e.g., the Mercator and
@@
 Equirectangular map projections.
-There is no need for East or West positive longitude definitions as the pixel
-conversion in world coordinates is set by an oriented matrix.
+This simplification is effective when a spherical local approximation
+is applied on planetary imagery.
+This is particularly the case with growing image resolutions and
+increasing computing capabilities: spherical projections with a local
+radius definition are easier to reproject from one system to another, than
+global spheroidal data (see for example the distribution of HiRISE images).
+In addition, there is no need for East or West positive longitude
+definitions as the pixel conversion in world coordinates is set by an
+oriented matrix.
+Table \ref{table:proj} summarizes the FITS projection codes and their
+correspondent PDS4 projection names (see table A.1. in \citet{calagreisenI}
+for details).
 
@@
 \caption{Projection names and codes already available in FITS suitable for
-planetary applications: their translation in PDS4 vocabulary is
-provided. See Table A.1. in \citet{calagreisenI} for details.}
+planetary applications: their correspondent in PDS4 vocabulary are
+provided.}
 \centering
@@
 \hline
+Zenithal Equidistant & ARC & Azimuthal Equidistant \\
+Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
+Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
+Equidistant Conic & COD & Equidistant Conic \\
+Conic Equal-Area & COE & Albers Conical Equal Area \\ 
 Sanson-Flamsteed & SFL & Sinusoidal \\
-Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
 Conic Ortomorphic & COO & Lambert Conformal Conic\\
-Plate Carre & CAR & Equirectangular \\
-Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
-Mercator & MER & Mercator, Oblique Mercator \\
+%Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
+Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
+Polyconic & PCO & Polyconic \\
 Ortographic & SIN & Orthographic \\
-Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
 Stereographic & STG & Stereographic, Polar Stereographic \\
 Gnomonic & TAN & Gnomonic \\
+Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
 \hline
 \multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
-\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
+\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}.
+We cannot find correspondences for PDS4 Robinson, Van Den Grinten and Space Oblique Mercator.}
 \end{tabular}

### 2018-02-02 epss_marmo.tex

@@
 \affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
-CNRS, Sorbonne Universités, UPMC Univ. Paris 06, Univ. Paris Diderot,
-Sorbonne Paris Cité, Meudon, France}
+CNRS, Sorbonne Universit\'es, UPMC Univ. Paris 06, Univ. Paris Diderot,
+Sorbonne Paris Cit\'e, Meudon, France}
 \affiliation{4}{Jacobs University, Bremen, Germany}
@@
 \item{a brief description of FITS format;}
-\item{a proposition for planetary surface meta-data description in FITS;}
-\item{dictionaries between FITS, PDS meta-data description and EPN-TAP2.0
+\item{a proposition for planetary surface metadata description in FITS;}
+\item{dictionaries between FITS, PDS metadata description and EPN-TAP2.0
 parameters;}
@@
 to store inhomogeneous digital information (reflectance, calibration data,
-vector, table data, etc.) in the same file, each with relative metadata,
+vector, table data, etc.) in the same file, each with correspondent metadata,
 as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
@@
 (projection center or reference point)
-describe the difference between,e.g., the Mercator and
+describe the difference between, e.g., the Mercator and
 Oblique Mercator or Simple Cylindrical and
@@
 radius definition are easier to reproject from one system to another, than
-global spheroidal data (see for example the distribution of HiRISE images).
+from or to a global spheroidal datum (see for example how the HiRISE team
+has chosen to distribute HiRISE images).
 In addition, there is no need for East or West positive longitude
@@
 oriented matrix.
+If needed, FITS does offer the possibility to fully describe the projected
+coordinate system using several alternative WCS definitions (meters, degrees,
+East and West longitude, etc.)\citep{greisencalaII}.
+
 Table \ref{table:proj} summarizes the FITS projection codes and their
@@
 Conic Equal-Area & COE & Albers Conical Equal Area \\ 
-Sanson-Flamsteed & SFL & Sinusoidal \\
 Conic Ortomorphic & COO & Lambert Conformal Conic\\
@@
 Polyconic & PCO & Polyconic \\
+Sanson-Flamsteed & SFL & Sinusoidal \\
 Ortographic & SIN & Orthographic \\
@@
 \multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}.
-We cannot find correspondences for PDS4 Robinson, Van Den Grinten and Space Oblique Mercator.}
+We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
+\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
 \end{tabular}
 \end{table}
+
+Table \ref{table:wcskeys}
 
@@
 
-However, if still needed, FITS does offer the possibility to fully describe those
-same parameters using alternative WCS definitions (meters, degrees,
-East and West longitude, etc.)\citep{greisencalaII}.
 Nevertheless, WCS does not record information about the body shape and
@@
 %
-%\appendix
-%\section{Here is a sample appendix}
+\appendix
+\section{Correspondence between reserved FITS keywords and PDS4 standard reference}
+\label{app:stdfitspds4}
 %This is an Appendix section.
-
-%\subsection{subsection}
-%This is an Appendix subsection.
-
-%\subsubsection{subsubsection}
-%This is an Appendix subsubsection.
+The correspondence between mandatory and reserved FITS keywords (as defined in
+the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
+and PDS4 attributes, is described in this Appendix.
+\begin{itemize}
+\item{\texttt{BITPIX}.
+It specifies the number of bits that represent a data value in
+the associated data array.
+PDS4 uses the attribute \texttt{data\_type}: correspondent 
+values depend on the associated object class.}
+
+\item{\texttt{NAXIS}.
+It contains a non-negative integer representing the number of axes in
+the associated data array.
+A value of zero signifies that no data follow the header.
+PDS4 uses \texttt{axes}, always defined as a non-negative integer.}
+
+\item{\texttt{NAXISn}.
+It contains a non-negative integer representing the number of
+elements along axis \texttt{n} of a data array.
+The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
+the count of the number of elements along an array.}
+
+\item{\texttt{BUNIT}.
+It contains a string, describing the physical units in which the quantities
+in the array are expressed.
+The units of all FITS header keyword values should conform with the recommendations
+in the IAU Style Manual.
+For angular measurements given as floating point values, degrees are the recommended
+units (with the units, if specified, given as 'deg').
+PDS4 specifies in the class Element (of a vector, array, or any other class
+containing Elements), a \texttt{unit\_id}} (unit abbreviation) and a \texttt{unit}
+(unit long name).
+
+\item{\texttt{BSCALE}.
+It contains a floating point number representing the coefficient of the linear term
+in the scaling equation $physical\_value = BZERO + BSCALE \times array\_value$,
+the ratio of physical value to array value at zero offset.
+For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
+attribute. Both FITS and PDS4 defaulted this value to $1$.}
+
+\item{\texttt{BZERO}.
+It contains a floating point number representing the constant term in the
+scaling equation $physical\_value = BZERO + BSCALE \times array\_value$
+For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
+attribute. Both FITS and PDS4 defaulted this value to $0$.}
+
+\item{\texttt{BLANK}.
+This keyword must be used only in integer data.
+It contains an integer that specifies the value that is used to
+represent pixels that have an undefined physical value.}
+
+\item{\texttt{DATAMAX}}
+\item{\texttt{DATAMIN}}
+\item{\texttt{DATE}}
+\item{\texttt{DATE-OBS}}
+\item{\texttt{INSTRUME}}
+\item{\texttt{TELESCOP}}
+\item{\texttt{OBJECT}}
+\item{\texttt{ORIGIN}}
+\item{\texttt{AUTHOR}}
+\item{\texttt{REFERENC}}
+
+\end{itemize}
+
 %\begin{linenomath*}

### 2018-02-05 ess_marmo.tex
Renaming files.

--- 2018-02-02 epss_marmo.tex
@@
 
-\subsubsection{FITS to PDS3 and PDS4 dictionaries}
+\subsubsection{Correspondence between reserved FITS keywords and PDS4 standard reference}
+\label{sssec:stdfitspds4}
+The correspondence between mandatory and reserved FITS keywords (as defined in
+the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
+and PDS4 attributes, is described here.
+\begin{itemize}
+\item{\texttt{BITPIX}.
+It specifies the number of bits that represent a data value in
+the associated data array.
+PDS4 uses the attribute \texttt{data\_type}: correspondent 
+values depend on the associated object class.}
+
+\item{\texttt{NAXIS}.
+It contains a non-negative integer representing the number of axes in
+the associated data array.
+A value of zero signifies that no data follow the header.
+PDS4 uses \texttt{axes}, always defined as a non-negative integer.}
+
+\item{\texttt{NAXISn}.
+It contains a non-negative integer representing the number of
+elements along axis \texttt{n} of a data array.
+The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
+the count of the number of elements along an array.}
+
+\item{\texttt{BUNIT}.
+It contains a string, describing the physical units in which the quantities
+in the array are expressed.
+The units of all FITS header keyword values should conform with the recommendations
+in the IAU Style Manual.
+For angular measurements given as floating point values, degrees are the recommended
+units (with the units, if specified, given as 'deg').
+PDS4 specifies in the class Element (of a vector, array, or any other class
+containing Elements), a \texttt{unit\_id}} (unit abbreviation) and a \texttt{unit}
+(unit long name).
+
+\item{\texttt{BSCALE}.
+It contains a floating point number representing the coefficient of the linear term
+in the scaling equation $physical\_value = BZERO + BSCALE \times array\_value$,
+the ratio of physical value to array value at zero offset.
+For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
+attribute. Both FITS and PDS4 defaulted this value to $1$.}
+
+\item{\texttt{BZERO}.
+It contains a floating point number representing the constant term in the
+scaling equation $physical\_value = BZERO + BSCALE \times array\_value$
+For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
+attribute. Both FITS and PDS4 defaulted this value to $0$.}
+
+\item{\texttt{BLANK}.
+This keyword must be used only in integer data.
+It contains an integer that specifies the value that is used to
+represent pixels that have an undefined physical value.... 
+\textit{PDS4 Special\_Constants?}}
+
+\item{\texttt{DATAMAX}}
+\item{\texttt{DATAMIN}}
+
+\item{\texttt{DATE}.
+The date on which the file was created,}
+\item{\texttt{DATE-OBS}}
+\item{\texttt{INSTRUME}}
+\item{\texttt{TELESCOP}}
+\item{\texttt{OBJECT}}
+\item{\texttt{ORIGIN}}
+\item{\texttt{AUTHOR}}
+\item{\texttt{REFERENC}}
+
+\end{itemize}
 
@@
 %
-\appendix
-\section{Correspondence between reserved FITS keywords and PDS4 standard reference}
-\label{app:stdfitspds4}
+%\appendix
+%\section{Here Is Appendix Title}
 %This is an Appendix section.
-The correspondence between mandatory and reserved FITS keywords (as defined in
-the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
-and PDS4 attributes, is described in this Appendix.
-\begin{itemize}
-\item{\texttt{BITPIX}.
-It specifies the number of bits that represent a data value in
-the associated data array.
-PDS4 uses the attribute \texttt{data\_type}: correspondent 
-values depend on the associated object class.}
-
-\item{\texttt{NAXIS}.
-It contains a non-negative integer representing the number of axes in
-the associated data array.
-A value of zero signifies that no data follow the header.
-PDS4 uses \texttt{axes}, always defined as a non-negative integer.}
-
-\item{\texttt{NAXISn}.
-It contains a non-negative integer representing the number of
-elements along axis \texttt{n} of a data array.
-The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
-the count of the number of elements along an array.}
-
-\item{\texttt{BUNIT}.
-It contains a string, describing the physical units in which the quantities
-in the array are expressed.
-The units of all FITS header keyword values should conform with the recommendations
-in the IAU Style Manual.
-For angular measurements given as floating point values, degrees are the recommended
-units (with the units, if specified, given as 'deg').
-PDS4 specifies in the class Element (of a vector, array, or any other class
-containing Elements), a \texttt{unit\_id}} (unit abbreviation) and a \texttt{unit}
-(unit long name).
-
-\item{\texttt{BSCALE}.
-It contains a floating point number representing the coefficient of the linear term
-in the scaling equation $physical\_value = BZERO + BSCALE \times array\_value$,
-the ratio of physical value to array value at zero offset.
-For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
-attribute. Both FITS and PDS4 defaulted this value to $1$.}
-
-\item{\texttt{BZERO}.
-It contains a floating point number representing the constant term in the
-scaling equation $physical\_value = BZERO + BSCALE \times array\_value$
-For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
-attribute. Both FITS and PDS4 defaulted this value to $0$.}
-
-\item{\texttt{BLANK}.
-This keyword must be used only in integer data.
-It contains an integer that specifies the value that is used to
-represent pixels that have an undefined physical value.}
-
-\item{\texttt{DATAMAX}}
-\item{\texttt{DATAMIN}}
-\item{\texttt{DATE}}
-\item{\texttt{DATE-OBS}}
-\item{\texttt{INSTRUME}}
-\item{\texttt{TELESCOP}}
-\item{\texttt{OBJECT}}
-\item{\texttt{ORIGIN}}
-\item{\texttt{AUTHOR}}
-\item{\texttt{REFERENC}}
-
-\end{itemize}
 

### 2018-02-05 ess_marmo.tex
Update on Overleaf.

@@
 represent pixels that have an undefined physical value.... 
-\textit{PDS4 Special\_Constants?}}
+\texttt{PDS4 Special\_Constants?}}
 
@@
 \item{\texttt{DATE}.
-The date on which the file was created,}
-\item{\texttt{DATE-OBS}}
-\item{\texttt{INSTRUME}}
+It contains the date on which the file was created, provided as ISO 8601
+string.
+PDS4 provides a \texttt{creation\_date\_time} attribute defining date
+and time of product creation.}
+
+\item{\texttt{DATE-OBS}.
+It contains the date of the observation, provided as ISO 8601
+string. The value shall be assumed to refer to the start of the observation,
+unless another interpretation is clearly explained in the comment field.
+It is assumed to be expressed in UTC but other systems are accepted as long as they
+are specified in the ISO string. 
+PDS4 provides a \texttt{Time\_Coordinates} class with attributes
+\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
+stop date and time appropriate to the product being labeled, always in UT
+coordinates.
+EPN-TAP also strongly recommends to fill in the \texttt{time\_min} and
+\texttt{time\_max} fields, always in UTC and formatted in Julian Days.}
+% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
+% 
+% > Julian Days
+% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
+% 
+% ^.
+
+\item{\texttt{INSTRUME}.}
 \item{\texttt{TELESCOP}}
@@
 
-\bibliography{epss_marmo}
+\bibliography{ess_marmo}
 

### 2018-02-05 ess_marmo.tex
Update on Overleaf.

@@
 
-Table \ref{table:wcskeys}
-
-\begin{table}
+Once reprojected in one of the standard projections listed in Table \ref{table:proj}
+Planetary images or maps can be described using the WCS scheme
+\citep{greisencalaII,calagreisenI}.
+In WCS context pixels are first translated in an intermediate pixel coordinate system
+via linear transformations, then mapped on the grid specified by the \texttt{CTYPE}
+keyword, containing the projection code and the coordinate system.
+The coordinate system must be defined with respect to a specific inertial reference frame.
+
+Multiple coordinate system descriptions are allowed in the same header: this is a valuable
+property for planetary surface imagery, as angular ('deg') and linear ('m') description of
+the surface are often needed at the same time.
+Further below 3D keyword will be proposed for FITS metadata, in order to standardize
+the conversion between angular and linear coordinates: alternate description will still
+be important in our framework as standard libraries (e.g.
+wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
+astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
+software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them.
+
+Table \ref{table:wcskeys} describes FITS WCS keywords and their recommended use in
+planetary surface context.
+
+\begin{sidewaystable}
 \label{table:wcskeys}
@@
 \centering
-\begin{tabular}{l l l l}
+\begin{tabular}{l l l l l}
 \hline
-Keyword & Type & Definition & PDS4 \\
+Keyword & Type & Definition & PDS4 & EPN-TAP \\
 \hline
-EQUINOX & real & & \\
-RADESYS	& string & & \\
+RADESYS	& string & Name of the inertial reference frame with respect & & \\
+& & to which the coordinate system is defined & & \\
+& & (defaulted to \texttt{'ICRS'}) & & \\
+EQUINOX & real & The value field contains the equinox in years for the celestial & & \\
+& & coordinate system in which positions are expressed & & \\
+& & (defaulted to \texttt{'J2000'}) & & \\
+WCSAXES & integer & Number of axes in the WCS description & & \\
+& & The value of WCSAXES may exceed the number of raster pixel axes & & \\
+WCSNAME & & & & \\
 CRPIXn & real & & \\
@@
 CUNITn & string & & \\
-WCSNAME & string & & \\
 CNAMEn & string (up to 68 characters) & & \\
@@
 \end{tabular}
-\end{table}
+\end{sidewaystable}
 
@@
 It contains an integer that specifies the value that is used to
-represent pixels that have an undefined physical value.... 
+represent pixels that have an undefined physical value.
+On floating point data the IEEE NaN must be used to represent undefined values.  
 \texttt{PDS4 Special\_Constants?}}
@@
 coordinates.
-EPN-TAP also strongly recommends to fill in the \texttt{time\_min} and
-\texttt{time\_max} fields, always in UTC and formatted in Julian Days.}
+The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
+they must always be formatted in UTC and Julian Days.}
 % * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
@@
 
-\item{\texttt{INSTRUME}.}
-\item{\texttt{TELESCOP}}
-\item{\texttt{OBJECT}}
-\item{\texttt{ORIGIN}}
-\item{\texttt{AUTHOR}}
-\item{\texttt{REFERENC}}
+\item{\texttt{INSTRUME}.
+It contains a character string identifying the instrument used to acquire the data.
+PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally corresponds
+to this FITS keywords.
+EPN-TAP provides with an \texttt{instrument\_name} parameter.}
+
+\item{\texttt{TELESCOP}.
+It contains a character string identifying the telescope used to acquire the data.
+The PDS4 \texttt{Instrument\_Host} class provides the description of the physical object
+mounting the instrument. Its \texttt{Name} attribute should correspond to the FITS keyword.
+PDS4 also provides for the definition of a \texttt{Telescope} class, in case of
+ground-based observations.
+In the EPN-TAP scheme this will defined the \texttt{instrument\_host\_name} parameter.
+\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}
+
+\item{\texttt{OBJECT}.
+It contains a character string giving a name for the observed object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
+The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to this
+FITS keyword.
+EPN-TAP standard also provides with a \texttt{target\_name} parameter.
+Notice that PDS4 and EPN-TAP also ask for a \texttt{target\_type} attribute, which must be
+specified from an enumerated list.}
+
+%
+%\item{\texttt{ORIGIN}.
+%It contains string identifying the organization or institution responsible for
+%creating the FITS file.
+%}
+%
+%\item{\texttt{AUTHOR}}
+
+\item{\texttt{REFERENC}.
+It contains a character string citing a reference where the data are published.
+It is recommended that either the Astrophysics Data System bibliographic databases\footnote{http://adswww.harvard.edu/}
+bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
+be used.
+PDS4 \texttt{External\_Reference} class 
+EPN-TAP \texttt{bib\_reference} has exactly the same meaning and use than the FITS keyword.}
 
@@
 
-\subsection{Best Practices and Examples}
-
+\subsection{Examples and Best Practices}
+\subsubsection{Global maps}
+\subsubsection{HiRISE mosaics}
+\subsubsection{OMEGA spectra and geometry data}
+\subsubsection{Akatsuki imaging data}
 \section{Inter-operable software developments}
@@
 \label{sec:disc}
+FITS discussion in astronomy\citep{THOMAS2015133} 
+let's astronomy learn from planetary sciences too: accent must be put on interoperability
+.... blabla
+Planetary surface is lacking open and modular software for massive processing and can benefit
+of astronomy experience. In addition to that WCS projection scheme is simpler than the
+historical one used in planetary applications. 
+Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
+Reducing data format conversion during the processing chain but
+allowing easy conversion for final products following the needed application.
 

### 2018-02-06 ess_marmo.tex
Update on Overleaf.

@@
  M. Minin\affil{4},
- A. P. Rossi\affil{4} ...}
+ A. Zinzi\affil{5},
+ A. P. Rossi\affil{4}}
 
@@
 proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
-taking into account the WGCCRE recommendations.
+taking into account the WGCCRE recommendations \citep{MarmoLPSC}.
 
@@
 astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
-software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them.
+software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
+coordinate tracking more efficient.
 
@@
 \subsection{Examples and Best Practices}
-\subsubsection{Global maps}
+
+\subsubsection{Visualization of global and projected maps}
+High level products are meant to be used for visual inspection or GIS analysis.
+They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
+format, as GeoTIFF make them GIS ready.
+Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
+and GeoTIFF are completely interchangeable.
+As an example we have downloaded some global maps from USGS
+Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
+We used ds9 to display the FITS image and were able to visualize coordinates in meters
+as in QGIS, but also in planetary latitude and longitude thanks to alternate coordinate
+description.
+
 \subsubsection{HiRISE mosaics}
+
 \subsubsection{OMEGA spectra and geometry data}

### 2018-02-06 ess_marmo.tex
Update on Overleaf.

@@
 though some have historically been anchored to Earth's spatial description.
+Several of them are based on the Geospatial Data Abstraction
+Library\footnote{http://www.gdal.org} (GDAL), released by the Open Source
+Geospatial Foundation.
+GDAL is essentially a format translation library written in C++ for geospatial raster
+and vector data and, fortunately, it also offer basic support for FITS format.
+
 FITS World Coordinate System
@@
 We used ds9 to display the FITS image and were able to visualize coordinates in meters
-as in QGIS, but also in planetary latitude and longitude thanks to alternate coordinate
-description.
+as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
+description (see figure \ref{fig:ds9qgis}, left side).
+\begin{figure}[ht!]
+\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
+\caption{left: Mercury South Pole FITS image displayed in ds9;
+right: GeoTIFF raster superposed to the virtual
+description of the FITS raster.
+Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
+Laboratory, Carnegie Science}
+\label{fig:ds9qgis}
+\end{figure}
+
+Assuming projected products distributed in FITS format a quick method to add GIS support for
+FITS images is to use a GDAL-supported detached header.
+GDAL Virtual Header files simply describe the internal structure of a FITS image, but
+also the map projection and body size in a standardized "well known text"
+(WKT) projection string.
+As a result, the FITS raster loaded into QGIS thanks to its correspondent GDAL Virtual Header
+is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).
+
+MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
 
@@
 \subsubsection{OMEGA spectra and geometry data}
+
 \subsubsection{Akatsuki imaging data}

### 2018-02-07 ess_marmo.tex
Update on Overleaf.

@@
  S. Erard\affil{3},
+ M. Minin\affil{4},
+ A. Zinzi\affil{5,6},
  B. Cecconi\affil{3},
- M. Minin\affil{4},
- A. Zinzi\affil{5},
  A. P. Rossi\affil{4}}
@@
 \affiliation{4}{Jacobs University, Bremen, Germany}
-%\affiliation{5}{}
+\affiliation{5}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
+\affiliation{6}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
 
@@
 
-Assuming projected products distributed in FITS format a quick method to add GIS support for
+Assuming projected products distributed in FITS format, a quick method to add GIS support for
 FITS images is to use a GDAL-supported detached header.
@@
 
+In order to straightforward load FITS files in GIS software, it is recommended to have
+body radii and linear coordinates filled in.
+Information about invalid values (NoData values) and minimum and maximum values are useful
+for correct display.
+
+\subsubsection{OMEGA spectra and geometry data}
+As stressed above, having planetary data distributed in FITS at any processing level is
+a way to avoid multiple conversions in the processing chain and simplify reprocessing.
+As an example of non projected planetary data we have converted in FITS format data
+acquired by the spectroimager OMEGA onboard of Mars Express.
+OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
+Their structure is described in the OMEGA Experiment Archive 
+Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
+In planetary science a first spatial calibration data often consists in providing coordinates
+on specific points on the detector (the center of the pixel, one or more pixel corners, etc.)
+
+\subsubsection{Akatsuki imaging data}
+
+\subsubsection{HiRISE mosaics}
+
+\section{Inter-operable software developments}
+VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
 MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
-
-\subsubsection{HiRISE mosaics}
-
-\subsubsection{OMEGA spectra and geometry data}
-
-\subsubsection{Akatsuki imaging data}
-\section{Inter-operable software developments}
 \label{sec:softdev}
@@
 under the Horizon 2020 research and innovation program, grant agreement N.654208.
+C. Marmo would lik to thanks Mark Calabretta from Australia Telescope National Facility
+for his WCSLIB library, and David Berry from Joint Astronomy Center, Hilo, for useful
+discussions about TAB projection in FITS.
 

### 2018-02-08 ess_marmo.tex
Update on Overleaf.

@@
  M. Minin\affil{4},
- A. Zinzi\affil{5,6},
+ F.-X. Pineau\affil{5},
+ A. Zinzi\affil{6,7},
  B. Cecconi\affil{3},
@@
 \affiliation{4}{Jacobs University, Bremen, Germany}
-\affiliation{5}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
-\affiliation{6}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
+\affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
+\affiliation{6}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
+\affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
 
@@
 As an example of non projected planetary data we have converted in FITS format data
-acquired by the spectroimager OMEGA onboard of Mars Express.
+acquired by the spectroimager OMEGA on-board of Mars Express.
 OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
@@
 In planetary science a first spatial calibration data often consists in providing coordinates
-on specific points on the detector (the center of the pixel, one or more pixel corners, etc.)
+on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
+To support such look-up table coordinate representation the \texttt{TAB} algorithm
+is defined in \citet{wcsspectral}.
 
@@
 \section{Inter-operable software developments}
+\label{sec:softdev}
 VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
 MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
-\label{sec:softdev}
+Aladin\footnote{http://aladin.u-strasbg.fr/}
 
@@
 under the Horizon 2020 research and innovation program, grant agreement N.654208.
-C. Marmo would lik to thanks Mark Calabretta from Australia Telescope National Facility
+C. Marmo would like to thanks Mark Calabretta from Australia Telescope National Facility
 for his WCSLIB library, and David Berry from Joint Astronomy Center, Hilo, for useful

### 2018-02-08 ess_marmo.tex

@@
 a way to avoid multiple conversions in the processing chain and simplify reprocessing.
+In planetary science a first spatial calibration data often consists in providing coordinates
+on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
+To support such look-up table coordinate representation the \texttt{TAB} algorithm
+has been defined in \citet{wcsspectral}.
+In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array, an indexing vector
+can be used to address coordinate array elements.
+Coordinates can then be sampled more or less coarsely depending on the behavior of the spatial
+reconstruction.
+Also, when the field is only partially covered by the planetary surface, coordinate array
+dimensions can be significantly reduced.
+
 As an example of non projected planetary data we have converted in FITS format data
@@
 Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
-In planetary science a first spatial calibration data often consists in providing coordinates
-on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
-To support such look-up table coordinate representation the \texttt{TAB} algorithm
-is defined in \citet{wcsspectral}.
 

### 2018-02-08 ess_marmo.tex
Update on Overleaf.

@@
 Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
+\begin{figure}[ht!]
+\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{60pt}\includegraphics[height=17pc]{ess_marmo3}}
+\caption{left: OMEGA data structure (from OMEGA EAICD); right: FITS Multi Digital Object
+conversion}
+\label{fig:omegascheme}
+\end{figure}
+
+TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
+available in all major linux distribution, and consequently in all
 
@@
 VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
+
 MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
+
 Aladin\footnote{http://aladin.u-strasbg.fr/}

### 2018-02-09 ess_marmo.tex
Update on Overleaf.

@@
 %% ------------------------------------------------------------------------ %%
-
-% (A title should be specific, informative, and brief. Use
-% abbreviations only if they are defined in the abstract. Titles that
-% start with general terms then specific results are optimized in
-% searches)
-
-% Example: \title{This is a test title}
 
@@
 
-%%% Suggested section heads:
-% \section{Introduction}
-% 
-% The main text should start with an introduction. Except for short
-% manuscripts (such as comments and replies), the text should be divided
-% into sections, each with its own heading.
-
-% Headings should be sentence fragments and do not begin with a
-% lowercase letter or number. Examples of good headings are:
-
-% \section{Materials and Methods}
-% Here is text on Materials and Methods.
-
-% \subsection{A descriptive heading about methods}
-% More about Methods.
-% 
-% \section{Data} (Or section title might be a descriptive heading about data)
-% 
-% \section{Results} (Or section title might be a descriptive heading about the
-% results)
-% 
-% \section{Conclusions}
-
 \section{Introduction}
@@
 
-\section{FITS format in Planetary Surface investigation}
+\section{FITS Format in Planetary Surface Investigation}
 \label{sec:fitspss}
 
-\subsection{FITS format}
+\subsection{FITS Format}
 FITS is an open digital standard, defined for data acquisition and archiving
@@
 
-\subsection{Planetary Surface proposed convention}
+\subsection{Planetary Surface Proposed Convention}
 \label{ssec:geofits}
@@
 Conic Ortomorphic & COO & Lambert Conformal Conic\\
-%Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
 Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
@@
 
-\subsubsection{Correspondence between reserved FITS keywords and PDS4 standard reference}
+\subsubsection{Correspondence Between Reserved FITS Keywords and PDS4 Standard Reference}
 \label{sssec:stdfitspds4}
@@
 and PDS4 attributes, is described here.
-\begin{itemize}
+\begin{enumerate}
 \item{\texttt{BITPIX}.
@@
 
-\end{itemize}
+\end{enumerate}
 
@@
 
-\subsubsection{Visualization of global and projected maps}
+\subsubsection{Visualization of Global and Projected Maps}
 High level products are meant to be used for visual inspection or GIS analysis.
@@
 
-\subsubsection{OMEGA spectra and geometry data}
+\subsubsection{OMEGA Spectra and Geometry Data}
 As stressed above, having planetary data distributed in FITS at any processing level is
@@
 dimensions can be significantly reduced. 
+TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
+available in all major linux distribution.
 
@@
 \caption{left: OMEGA data structure (from OMEGA EAICD); right: FITS Multi Digital Object
-conversion}
+conversion.}
 \label{fig:omegascheme}
@@
 
-TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
-available in all major linux distribution, and consequently in all
-
-\subsubsection{Akatsuki imaging data}
-
-\subsubsection{HiRISE mosaics}
-
-\section{Inter-operable software developments}
+OMEGA rasters are equally sampled all over the field: we use them to exemplify
+TAB description without indexing, then we only use spatial information on the center
+of the pixel. 
+The output result can be downloaded at the address ...
+
+The \texttt{wcsware} tool (available in the \texttt{wcslib-utils} package in Fedora
+distribution, in the \texttt{wcslib-tools} in Debian) had been run on the output file,
+validating the WCS structure and checking the ability to convert to (\texttt{wcsware -w})
+and from (\texttt{wcsware -x}) pixel and planetary surface coordinates.
+
+\subsubsection{Akatsuki Imaging Data}
+
+\subsubsection{HiRISE Mosaics}
+
+\section{Interoperable Software Developments}
 \label{sec:softdev}
@@
 Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
-Reducing data format conversion during the processing chain but
+Reducing custom data format conversions during the processing chain but
 allowing easy conversion for final products following the needed application.
-
-%% Enter Figures and Tables near as possible to where they are first mentioned:
-%
-% DO NOT USE \psfrag or \subfigure commands.
-%
-% Figure captions go below the figure.
-% Table titles go above tables;  other caption information
-%  should be placed in last line of the table, using
-% \multicolumn2l{$^a$ This is a table note.}
-%
-%----------------
-% EXAMPLE FIGURE
-%
-% \begin{figure}
-% \end{figure}
-
-%\begin{figure}[ht!]
-% \centering
-% when using pdflatex, use pdf file:
-% \includegraphics[width=20pc]{figsamp.pdf}
-%
-% when using dvips, use .eps file:
-% \includegraphics[width=20pc]{figsamp.eps}
-%
-% If you don't specify the file extension, 
-% \includegraphics will insert the right file:
-%\centerline{\includegraphics[height=1.5in]{figsamp}}
-%\caption{Short caption}
-%\label{figone}
-%\end{figure}
-
-%\begin{figure}[ht!]
-%\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
-%\caption{The figure caption should begin with an overall descriptive
-%statement of the figure followed by additional text. They should be
-%immediately after each figure.  Figure parts are indicated with
-%lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
-%both the figures and captions in the text near where they are cited.}
-%\label{fig:mean_and_slope}
-%\end{figure}
-
-%\begin{table}
-%\caption{Start this caption with a short description of your
-%table.
-%Large tables
-%especially presenting rich data should be presented as separate excel
-%or .cvs files, not as part of the main text.}
-%== Table Here ==
-%\end{table}
-
-%
-% ---------------
-% EXAMPLE TABLE
-%
-%\begin{table}
-%\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-%\centering
-%\begin{tabular}{l c}
-%\hline
-% Run  & Time (min)  \\
-%\hline
-%  $l1$  & 260   \\
-%  $l2$  & 300   \\
-%  $l3$  & 340   \\
-%  $h1$  & 270   \\
-%  $h2$  & 250   \\
-%  $h3$  & 380   \\
-%  $r1$  & 370   \\
-%  $r2$  & 390   \\
-%\hline
-%\multicolumn{2}{l}{$^{a}$Table note text here.}
-%\end{tabular}
-%\end{table}
-
-% See below for how to make sideways figures or tables.
-
-% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
-%
-% \begin{sidewaysfigure}
-% \includegraphics[width=20pc]{figsamp}
-% \caption{Caption here}
-% \label{newfig}
-% \end{sidewaysfigure}
-%
-% \begin{sidewaystable}
-% \caption{Caption here}
-%\label{tab:signif_gap_clos}
-% \begin{tabular}{ccc}
-%one&two&three\\
-%four&five&six
-% \end{tabular}
-% \end{sidewaystable}
-
-%\clearpage
-
-%% ------------------------------------------------------------------------ %%
-%\section*{Citations}
-
-% Please use ONLY \citet and \citep for reference citations.
-% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).
-
-%\subsection*{Cites made with \tt\string\citet\string{\string}}
-%...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
-%\citet{Ghel00}, and \citet{Leit74}.
-
-%\subsection*{Cites made with \tt\string\citep\string{\string}}
-%...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
-%\citep{Ghel00, Leit74}.
-
-%...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.
 

### 2018-02-09 ess_marmo.tex
Update on Overleaf.

@@
 Those reasons make FITS format appropriate for large amounts of raster data
-processing.
+automatic processing.
 The goal of this approach is to make easier data mining and reprocessing
@@
 a way to avoid multiple conversions in the processing chain and simplify reprocessing.
-In planetary science a first spatial calibration data often consists in providing coordinates
+In planetary science a first spatial calibrated product often consists in providing coordinates
 on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
@@
 of the pixel. 
-The output result can be downloaded at the address ...
+The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/orb0413\_1.IR1.fits.gz together with the python script
+used for conversion\footnote{https://github.com/cmarmo/convertofits/blob/master/omega2fits.py}.
 
@@
 \subsubsection{Akatsuki Imaging Data}
-
-\subsubsection{HiRISE Mosaics}
+The Venus Climate Orbiter Akatsuki\footnote{http://akatsuki.isas.jaxa.jp/en/} is a JAXA
+spacecraft dedicated to the study of the Venus atmosphere.
+Akatsuki imaging data are already distributed in FITS format.
+Geometry information is distributed in two different cubes of data, containing respectively 
+values at the center of the pixel and at the lower left corner of the pixel.
+Tabular coordinate representation allows us to gather geometry in one table together with
+the described raster.
+In addition to that, in Akatsuki images Venus is often far from completely filling the field:
+tabular representation avoids to store unnecessary data. 
+In that case, coordinate indexes are necessary to identify which pixels contains the spatial
+information.
+The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/ir1\_20160415\_070351\_097\_l2b\_v10\_out.fit.gz
+together with the python script
+used for conversion\footnote{https://github.com/cmarmo/convertofits/blob/master/akatsukil2b.py}.
+
+Again, we used \texttt{wcsware} to validate the WCS structure and to test pixel to world /
+world to pixel coordinate conversion.
+Even if TAB projection is not fully implemented in FITS visualization tools, this representation
+has the advantage to store the detector geometrical information together with physical
+quantities measured by the detector itself in a standardized way.
 
@@
 .... blabla
-Planetary surface is lacking open and modular software for massive processing and can benefit
+Planetary surface is lacking open and general software for massive processing and can benefit
 of astronomy experience. In addition to that WCS projection scheme is simpler than the

### 2018-02-13 ess_marmo.tex
Update on Overleaf.

@@
 brings together several research communities (geologists, astronomers,
-physicists, geochemists, etc.).
-Planetary missions produce an impressively growing amount of heterogeneous
-data which mining evolves those days from mostly visual assessment to more
-automated quantitative analysis.
-Interoperability and openness of data formats and processing techniques are
-becoming a necessity, to avoid the risk of being unable to efficiently extract
-scientific information from the data, and to guarantee the reproducibility of
-the scientific results.
-This paper will describe how Flexible Image Transport System (FITS) can be
-used in planetary surface investigations, and how its metadata can easily be
-inserted in the PDS4 metadata distribution model.
-An evolution to the actual standard is proposed to adapt FITS to
-planetary body description.
-Some examples and implementations are then presented.
-More than imposing a formal data model, FITS description for those data aims
-to simplify the processing chain in planetary science, promoting
+physicists, geochemists, etc.). Planetary missions produce an impressively growing 
+amount of diverse data requiring an evolution from a mostly manual-based visual 
+analysis to a more automated and quantitative analysis.
+Interoperability, openness of data formats, and shared processing techniques are
+becoming a necessity to efficiently extract scientific information from the data and 
+to guarantee the reproducibility of the scientific results.
+Unfortunately, the technologies 
+and data formats used by researchers for planetary surface studies and the field of astronomy 
+diverged as these related, but almost completely isolated domains evolved.  
+% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
+% 
+% > Unfortunately, the technologies 
+% > and data formats used by researchers for planetary surface studies and the field of astronomy 
+% > diverged as these related, but almost completely isolated domains evolved.  
+% 
+% Is this too strong a statement?
+% 
+% ^.
+In this paper we will will describe how a small addition to the Flexible Image Transport System (FITS) 
+standard, widely used in the astronomy investigations,  will allow FITS to be more easily used in planetary
+surface investigations , we will  show  how FITS metadata can easily be transformed in the PDS4 metadata
+archival model, and lastly provide example implementations.
+More than imposing a formal data model, a FITS description for planetary data aims
+to simplify sharing data across the planetary and astronomy domains and promoting
 interoperability from data acquisition to final visualization.
@@
 
-Table \ref{table:wcskeys} describes FITS WCS keywords and their recommended use in
-planetary surface context.
-
-\begin{sidewaystable}
-\label{table:wcskeys}
-\caption{WCS FITS keywords in the Planetary Surface context}
-\centering
-\begin{tabular}{l l l l l}
-\hline
-Keyword & Type & Definition & PDS4 & EPN-TAP \\
-\hline
-RADESYS	& string & Name of the inertial reference frame with respect & & \\
-& & to which the coordinate system is defined & & \\
-& & (defaulted to \texttt{'ICRS'}) & & \\
-EQUINOX & real & The value field contains the equinox in years for the celestial & & \\
-& & coordinate system in which positions are expressed & & \\
-& & (defaulted to \texttt{'J2000'}) & & \\
-WCSAXES & integer & Number of axes in the WCS description & & \\
-& & The value of WCSAXES may exceed the number of raster pixel axes & & \\
-WCSNAME & & & & \\
-CRPIXn & real & & \\
-CTYPEn & string & & \\
-CRVALn & real & & \\
-CDi\_j & real matrix & & \\
-PCi\_j & real matrix & & \\
-CDELTn & real & & \\
-CUNITn & string & & \\
-CNAMEn & string (up to 68 characters) & & \\
-\hline
-\end{tabular}
-\end{sidewaystable}
+\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
+\label{sssec:wcskeys}
+In this section WCS FITS keywords are inserted in the planetary surface context.
+Correspondence with PDS4 and EPN-TAP are added when available.
+\begin{enumerate}
+\item{RADESYS.
+Name of the inertial reference frame with respect to which the coordinate
+system is defined (defaulted to \texttt{'ICRS'})}
+\item{EQUINOX.
+The value field contains the equinox in years for the celestial
+coordinate system in which positions are expressed (defaulted to \texttt{'J2000'})}
+\item{WCSAXES.
+Number of axes in the WCS description.
+The value of WCSAXES may exceed the number of raster pixel axes.}
+\item{WCSNAME.
+Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic, non-rotating, inertial).
+EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
+\item{CRPIXn.
+Location of a reference point along axis $n$, in units of the axis index.
+The reference point value do not need to lie within the actual data array.}
+\item{CTYPEn.
+Name of the coordinate represented by axis $n$, inn “4–3” form: the first four characters
+specify the coordinate type, the fifth character is a '-' and the
+remaining three characters specify an algorithm code for computing the world coordinate
+value (\ref{table:proj}).
+For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
+are listed in Table \ref{table:ctype}.
+For linear coordinates a similar scheme (xPX as Projected X xPY as Projected Y or yzPX and yzPY) is proposed.
+Alternative representation of WCS are already possible in FITS, to describe images in degrees and in meters
+\citep{greisencalaII}.
+The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
+\item{CRVALn.
+Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
+\item{CDi\_j.
+The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
+world coordinates.}
+\item{PCi\_j.
+The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel coordinates (to be used with CDELT keywords).}
+\item{CDELTn.
+The coordinate increment along axis $n$ (scaling component of the linear transformation
+matrix PCi\_j).
+In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
+attributes.
+In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
+\item{CUNITn.
+Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
+and to meters (m) for linear ones.}
+\item{CNAMEn.
+The description of a particular coordinate for an axis in a particular WCS (up to 68
+characters), while the WCSNAMEa keyword names the particular WCS as a whole.
+Its default value will be blank.}
+\end{enumerate}
 
@@
 \label{table:ctype}
-\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.}
+\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
+Codes for planets are assumed in \citet{calagreisenI}. We propose to add codes for classes of targets,
+more or less corresponding to PDS4 Target type attribute and to EPN-TAP target\_class parameter.}
 \centering
@@
 Neptune & NE \\
-Planetary bodies & B \\
+Satellites (other than the Moon) & ST \\
+Asteroids & AS \\
+Dwarf Planets & DW \\
+Comets & CO \\
 \hline
@@
 
-%
-%\item{\texttt{ORIGIN}.
-%It contains string identifying the organization or institution responsible for
-%creating the FITS file.
-%}
+\item{\texttt{ORIGIN}.
+It contains string identifying the organization or institution responsible for
+creating the FITS file.
+}
 %

### 2018-02-14 ess_marmo.tex
Update on Overleaf.

@@
 to guarantee the reproducibility of the scientific results.
-Unfortunately, the technologies 
-and data formats used by researchers for planetary surface studies and the field of astronomy 
+Unfortunately, the technologies and data formats used by researchers for
+planetary surface studies and the field of astronomy 
 diverged as these related, but almost completely isolated domains evolved.  
@@
 metadata distribution model.
-FITS is one of the standard implemented formats in the Virtual
+FITS is one of the standard formats implemented in the Virtual
 Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
@@
 processing tools.
-Those reasons make FITS format appropriate for large amounts of raster data
-automatic processing.
+Those reasons make FITS format appropriate to automatically handle large amounts of raster data.
 The goal of this approach is to make easier data mining and reprocessing
@@
 standards.
-The option to use FITS within the planetary domain could potentially
+The option to use FITS within the planetary surface domain could potentially
 homogenize methods from acquisition, to visualization, while
@@
 \item{some examples of planetary dataset in FITS;}
-\item{current developments for making FITS usable in planetary sciences.}
+\item{current developments for making FITS \added{more} usable \deleted{in planetary sciences} \added{to handle data of planetary surfaces}.}
 \end{enumerate}
@@
 for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
-Therefore, FITS is one of the standard implemented formats in the Virtual
+Therefore, FITS is one of the standard formats implemented in the Virtual
 Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
@@
 terrestrial references.
-There is no need for oblique projection definitions
+For instance, there is no need for oblique projection definitions
 \citep{calagreisenI,Snyder1987} as projection parameters
@@
 has chosen to distribute HiRISE images).
-In addition, there is no need for East or West positive longitude
-definitions as the pixel conversion in world coordinates is set by an
+In addition, WCS resolves the ambiguity between east-handed and west-handed longitude systems,
+as the pixel conversion in world coordinates is set by an
 oriented matrix.
@@
 \item{WCSNAME.
-Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic, non-rotating, inertial).
+Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
+non-rotating, inertial).
 EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
@@
 \item{CTYPEn.
-Name of the coordinate represented by axis $n$, inn “4–3” form: the first four characters
+Name of the coordinate represented by axis $n$, in “4–3” form: the first four characters
 specify the coordinate type, the fifth character is a '-' and the
@@
 are listed in Table \ref{table:ctype}.
-For linear coordinates a similar scheme (xPX as Projected X xPY as Projected Y or yzPX and yzPY) is proposed.
-Alternative representation of WCS are already possible in FITS, to describe images in degrees and in meters
-\citep{greisencalaII}.
+For linear coordinates we propose a similar scheme: xPX as Projected X, xPY as Projected Y,
+or yzPX and yzPY.
+Alternative representation of WCS are already possible in FITS, to describe images in degrees
+and in meters \citep{greisencalaII}.
 The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
@@
 \item{PCi\_j.
-The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel coordinates (to be used with CDELT keywords).}
+The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
+coordinates (to be used with CDELT keywords).}
 \item{CDELTn.
@@
 elements along axis \texttt{n} of a data array.
-The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
+The \texttt{elements} attribute of the \added{PDS4} \texttt{Axis\_Array} class provides
 the count of the number of elements along an array.}
@@
 For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
-attribute. Both FITS and PDS4 defaulted this value to $1$.}
+attribute.
+Both FITS and PDS4 default this value to $1$.}
 
@@
 For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
-attribute. Both FITS and PDS4 defaulted this value to $0$.}
+attribute. Both FITS and PDS4 default this value to $0$.}
 
 \item{\texttt{BLANK}.
-This keyword must be used only in integer data.
+This keyword must be used only with integer data.
 It contains an integer that specifies the value that is used to
@@
 \textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
-stop date and time appropriate to the product being labeled, always in UT
-coordinates.
+stop date and time appropriate to the product being labeled, always in UTC.
 The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
@@
 It contains a character string identifying the instrument used to acquire the data.
-PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally corresponds
-to this FITS keywords.
+PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally \replaced{corresponds}{corresponding}
+to this FITS keyword.
 EPN-TAP provides with an \texttt{instrument\_name} parameter.}
@@
 ground-based observations.
-In the EPN-TAP scheme this will defined the \texttt{instrument\_host\_name} parameter.
+In the EPN-TAP scheme this is introduced by the \texttt{instrument\_host\_name} parameter.
 \footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}
@@
 It contains string identifying the organization or institution responsible for
-creating the FITS file.
-}
+creating the FITS file.}
+
 %
@@
 It contains a character string citing a reference where the data are published.
-It is recommended that either the Astrophysics Data System bibliographic databases\footnote{http://adswww.harvard.edu/}
+It is recommended that either the Astrophysics Data System bibliographic
+databases\footnote{http://adswww.harvard.edu/}
 bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
 be used.
-PDS4 \texttt{External\_Reference} class 
-EPN-TAP \texttt{bib\_reference} has exactly the same meaning and use than the FITS keyword.}
+PDS4 \texttt{External\_Reference} class and EPN-TAP \texttt{bib\_reference} have
+exactly the same meaning and use than the FITS keyword.}
 
@@
 format, as GeoTIFF make them GIS ready.
-Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
+Given some new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
 and GeoTIFF are completely interchangeable.
@@
 We used ds9 to display the FITS image and were able to visualize coordinates in meters
-as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
+as in QGIS, a popular open source GIS software, but also in planetary latitude and longitude
+thanks to the alternate coordinate
 description (see figure \ref{fig:ds9qgis}, left side).
@@
 
-Assuming projected products distributed in FITS format, a quick method to add GIS support for
+Assuming projected products are distributed in FITS format, a quick method to add GIS support for
 FITS images is to use a GDAL-supported detached header.
@@
 
-In order to straightforward load FITS files in GIS software, it is recommended to have
+In order to load FITS files in GIS software, it is recommended to have
 body radii and linear coordinates filled in.
 Information about invalid values (NoData values) and minimum and maximum values are useful
-for correct display.
+to correctly display the image dynamics.
 
@@
 As stressed above, having planetary data distributed in FITS at any processing level is
-a way to avoid multiple conversions in the processing chain and simplify reprocessing.
+a way to avoid multiple conversions in the processing chain and to simplify reprocessing.
 In planetary science a first spatial calibrated product often consists in providing coordinates
@@
 
-As an example of non projected planetary data we have converted in FITS format data
-acquired by the spectroimager OMEGA on-board of Mars Express.
+As an example of non projected planetary data we have converted in FITS format some data
+acquired by the imaging spectrometer OMEGA on-board of Mars Express.
 OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.

### 2018-02-14 ess_marmo.tex

@@
 \item{some examples of planetary dataset in FITS;}
-\item{current developments for making FITS \added{more} usable \deleted{in planetary sciences} \added{to handle data of planetary surfaces}.}
+\item{current developments for making FITS more usable to handle data
+of planetary surfaces.}
 \end{enumerate}
@@
 (VO).
-Finally, FITS format is compatible with the Planetary Data System (PDS)
+Finally, FITS data storage is compatible with the Planetary Data System (PDS)
 archiving specifications and supported by a large number of open libraries
@@
 to store inhomogeneous digital information (reflectance, calibration data,
-vector, table data, etc.) in the same file, each with correspondent metadata,
+vector, table data, etc.) in the same file, each with corresponding metadata,
 as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
@@
 \citep{crism} or OMEGA \citep{omega} instruments).
-FITS has been already chosen to distribute, e.g.,
-Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
-Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
-and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
+FITS has been already chosen to distribute data from, e.g.,
+Hayabusa AMICA and NIRS cameras\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
+all Akatsuki cameras\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
+(except the Lightning and Airglow Camera),
+and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
+In addition, many PDS3 datasets in the Small Bodies Node include FITS files.
 To be efficiently used in planetary surface investigations, FITS metadata must
@@
 FITS World Coordinate System
-representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html}
-simplifies the spatial coordinate description with respect to historical
+representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
+way to describe spatial dependencies in FITS metadata.
+WCS simplifies the spatial coordinate description with respect to historical
 terrestrial references.
-For instance, there is no need for oblique projection definitions
+F
+or instance, there is no need for oblique projection definitions
 \citep{calagreisenI,Snyder1987} as projection parameters
@@
 radius definition are easier to reproject from one system to another, than
-from or to a global spheroidal datum (see for example how the HiRISE team
-has chosen to distribute HiRISE images).
-In addition, WCS resolves the ambiguity between east-handed and west-handed longitude systems,
-as the pixel conversion in world coordinates is set by an
+from or to a global spheroidal datum.
+As an example, the HiRISE team has chosen to distribute HiRISE projected
+products using Equirectangular projection (CAR) on a local surface, approximated
+by the sphere with radius equivalent to the local Mars spheroid radius.
+
+In addition, WCS resolves the ambiguity between east-handed and west-handed
+longitude systems, as the pixel conversion in world coordinates is set by an
 oriented matrix.
@@
 Table \ref{table:proj} summarizes the FITS projection codes and their
-correspondent PDS4 projection names (see table A.1. in \citet{calagreisenI}
+corresponding PDS4 projection names (see table A.1. in \citet{calagreisenI}
 for details).
@@
 \caption{Projection names and codes already available in FITS suitable for
-planetary applications: their correspondent in PDS4 vocabulary are
+planetary applications: the corresponding PDS4 terminology is
 provided.}
@@
 In this section WCS FITS keywords are inserted in the planetary surface context.
-Correspondence with PDS4 and EPN-TAP are added when available.
+When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
+standardization purposes.
 \begin{enumerate}
@@
 the associated data array.
-PDS4 uses the attribute \texttt{data\_type}: correspondent 
+PDS4 uses the attribute \texttt{data\_type}: corresponding 
 values depend on the associated object class.}
@@
 The units of all FITS header keyword values should conform with the recommendations
-in the IAU Style Manual.
+in the IAU Style Manual\footnote{https://www.iau.org/publications/proceedings\_rules/units/}.
 For angular measurements given as floating point values, degrees are the recommended
@@
 It is assumed to be expressed in UTC but other systems are accepted as long as they
-are specified in the ISO string. 
+are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
 PDS4 provides a \texttt{Time\_Coordinates} class with attributes
@@
 The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
-they must always be formatted in UTC and Julian Days.}
-% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
-% 
-% > Julian Days
-% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
-% 
-% ^.
+they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}
 
@@
 It contains a character string identifying the instrument used to acquire the data.
-PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally \replaced{corresponds}{corresponding}
-to this FITS keyword.
+PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally 
+corresponds to this FITS keyword.
 EPN-TAP provides with an \texttt{instrument\_name} parameter.}
@@
 ground-based observations.
-In the EPN-TAP scheme this is introduced by the \texttt{instrument\_host\_name} parameter.
-\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}
+In the EPN-TAP scheme this is introduced by the \texttt{instrument\_host\_name}
+parameter\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}.}
 
 \item{\texttt{OBJECT}.
-It contains a character string giving a name for the observed object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
+It contains a character string giving a name for the observed
+object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
 The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to this
@@
 \item{\texttt{REFERENC}.
-It contains a character string citing a reference where the data are published.
+It contains a character string citing the bibliographic reference where the
+data are first described.
 It is recommended that either the Astrophysics Data System bibliographic
@@
 (WKT) projection string.
-As a result, the FITS raster loaded into QGIS thanks to its correspondent GDAL Virtual Header
+As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual Header
 is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).

### 2018-02-15 ess_marmo.tex
Update on Overleaf.

@@
 elements along axis \texttt{n} of a data array.
-The \texttt{elements} attribute of the \added{PDS4} \texttt{Axis\_Array} class provides
+The \texttt{elements} attribute of the PDS4 \texttt{Axis\_Array} class provides
 the count of the number of elements along an array.}
@@
 \begin{figure}[ht!]
-\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
+\centerline{\includegraphics[height=17pc]{ess_marmo2}\includegraphics[height=17pc]{ess_marmo1}}
 \caption{left: Mercury South Pole FITS image displayed in ds9;

### 2018-02-15 ess_marmo.tex
Update on Overleaf.

@@
 proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
-taking into account the WGCCRE recommendations \citep{MarmoLPSC}.
+taking into account the WGCCRE recommendations.
 
@@
 
-\item{\texttt{DATAMAX}}
-\item{\texttt{DATAMIN}}
+\item{\texttt{DATAMAX}.
+It always contains a floating-point number, giving the maximum valid physical value represented
+by the array, excluding any IEEE special values.
+PDS4 defines a \texttt{valid\_maximum} attribute in the \texttt{Special Constants} class
+having this same meaning.}
+
+\item{\texttt{DATAMIN}.
+It always contains a floating-point number, giving the minimum valid physical value represented
+by the array, excluding any IEEE special values.
+PDS4 defines a \texttt{valid\_minimum} attribute in the \texttt{Special Constants} class
+having this same meaning.}
 
@@
 \begin{figure}[ht!]
-\centerline{\includegraphics[height=17pc]{ess_marmo2}\includegraphics[height=17pc]{ess_marmo1}}
+\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
 \caption{left: Mercury South Pole FITS image displayed in ds9;
-right: GeoTIFF raster superposed to the virtual
-description of the FITS raster.
+right: GeoTIFF raster (displayed in yellow) and the virtual
+header linked to the FITS raster (displayed in red); the two layers
+are perfectly superposed.
 Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
@@
 a way to avoid multiple conversions in the processing chain and to simplify reprocessing.
-In planetary science a first spatial calibrated product often consists in providing coordinates
-on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
-To support such look-up table coordinate representation the \texttt{TAB} algorithm
-has been defined in \citet{wcsspectral}.
+In planetary science spatial information is first provided for specific points over the detector
+(the four corners of the detector, the center of each pixel, one or more pixel corners, etc.).
+To support in FITS metadata such look-up table coordinate representation the \texttt{TAB}
+algorithm has been defined in \citet{wcsspectral}.
 In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array, an indexing vector
@@
 acquired by the imaging spectrometer OMEGA on-board of Mars Express.
-OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
+OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA and PDS archives.
 Their structure is described in the OMEGA Experiment Archive 
@@
 \begin{figure}[ht!]
-\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{60pt}\includegraphics[height=17pc]{ess_marmo3}}
-\caption{left: OMEGA data structure (from OMEGA EAICD); right: FITS Multi Digital Object
-conversion.}
+\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo3}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo5}}
+\caption{left: OMEGA data structure (from OMEGA EAICD), science data and geometry description are
+distributed in different files; center: FITS Multi Digital Object scheme; as an example, we stored
+the spectral data in the first cube extension, the coordinate look-up table in the second
+extension, and other geometry parameters (incidence and emission angle, etc.) as images in the
+following extensions; right: BSQ cube visualization with ds9.}
 \label{fig:omegascheme}
@@
 OMEGA rasters are equally sampled all over the field: we use them to exemplify
-TAB description without indexing, then we only use spatial information on the center
-of the pixel. 
-The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/orb0413\_1.IR1.fits.gz together with the python script
-used for conversion\footnote{https://github.com/cmarmo/convertofits/blob/master/omega2fits.py}.
+TAB description without indexing.
+In addition, as geometrical description depends on the observation filter, this allows us to
+distribute a file for each filter, containing science and geometry data.
+The spectral cubes have been converted from Band Interleaved per Line (BIL) to Band Sequential
+(BSQ) in order to make them easily understandable when visualized using standard FITS viewer
+(as ds9 in figure \ref{fig:omegascheme})\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/orb0413\_1.IR1.fits.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/omega2fits.py}.
 
@@
 the described raster.
-In addition to that, in Akatsuki images Venus is often far from completely filling the field:
+In addition to that, in Akatsuki images Venus is sometime far from completely filling the field:
 tabular representation avoids to store unnecessary data. 
 In that case, coordinate indexes are necessary to identify which pixels contains the spatial
-information.
-The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/ir1\_20160415\_070351\_097\_l2b\_v10\_out.fit.gz
-together with the python script
-used for conversion\footnote{https://github.com/cmarmo/convertofits/blob/master/akatsukil2b.py}.
+information\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/ir1\_20160415\_070351\_097\_l2b\_v10\_out.fit.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/akatsukil2b.py}.
 
@@
 world to pixel coordinate conversion.
-Even if TAB projection is not fully implemented in FITS visualization tools, this representation
+Even though TAB projection is not fully implemented in FITS visualization tools, this representation
 has the advantage to store the detector geometrical information together with physical

### 2018-02-16 ess_marmo.tex
Update on Overleaf.

@@
  S. Erard\affil{3},
- M. Minin\affil{4},
- F.-X. Pineau\affil{5},
- A. Zinzi\affil{6,7},
+ A. Zinzi\affil{4,5},
+ M. Minin\affil{6},
+ F.-X. Pineau\affil{7},
  B. Cecconi\affil{3},
- A. P. Rossi\affil{4}}
+ A. P. Rossi\affil{6}}
 
@@
 Sorbonne Paris Cit\'e, Meudon, France}
-\affiliation{4}{Jacobs University, Bremen, Germany}
-\affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
-\affiliation{6}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
-\affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
+\affiliation{4}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
+\affiliation{5}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
+\affiliation{6}{Jacobs University, Bremen, Germany}
+\affiliation{7}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
 
@@
 Interoperability, openness of data formats, and shared processing techniques are 
-becoming a necessity to efficiently extract scientific information from the data and 
+a necessity to efficiently extract scientific information from the data and 
 to guarantee the reproducibility of the scientific results.
@@
 % ^.
-In this paper we will will describe how a small addition to the Flexible Image Transport System (FITS) 
+In this paper we will describe how a small addition to the Flexible Image Transport System (FITS) 
 standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
-surface investigations , we will  show  how FITS metadata can easily be transformed in the PDS4 metadata
-archival model, and lastly provide example implementations.
+surface investigations. We will  show  how FITS metadata can easily be transformed in the PDS4 metadata
+archival model and lastly provide example implementations.
 More than imposing a formal data model, a FITS description for planetary data aims
@@
 \label{sec:intro}
-Planetary science brings together several research communities
-(geologists, astronomers, physicists, geochemists, etc.),
-and planetary missions produce a growing amount of heterogeneous data.
-To efficiently extract scientific information from those data, they must
-be made integrable, thus interoperable.
-Openness of data format and processing techniques is becoming a necessity,
-to guarantee the reproducibility of the results obtained from those data.
+Planetary science encompasses a broad number of research fields and
+brings together several research communities (geologists, astronomers,
+physicists, geochemists, etc.). Planetary missions produce an impressively growing 
+amount of diverse data requiring an evolution from a mostly manual-based visual 
+analysis to a more automated and quantitative analysis.
+Interoperability, openness of data formats, and shared processing techniques are 
+a necessity to efficiently extract scientific information from the data and 
+to guarantee the reproducibility of the scientific results.
 
@@
 \citep{ERARDepntap}.
-FITS format is open and flexible, largely implemented in open and efficient
-processing tools.
-Those reasons make FITS format appropriate to automatically handle large amounts of raster data.
+FITS format is open, flexible, and largely implemented in open and efficient
+processing tools allowing to handle large amounts of raster data.
 The goal of this approach is to make easier data mining and reprocessing
@@
 standards.
-The option to use FITS within the planetary surface domain could potentially
+The option to use FITS within the planetary surface domain could help to
 homogenize methods from acquisition, to visualization, while
-giving chances to optimize data processing.
+giving chances to optimize data processing across the planetary and astronomy domains.
 
@@
 \subsection{FITS Format}
-FITS is an open digital standard, defined for data acquisition and archiving
-in astronomical observatories in the late 70's, and it is used for
-spatial telescope data.
+FITS is an open digital standard, defined for data acquisition and archiving telescope data
+by astronomical observatories since the late 1970's.
 The International Astronomical Union approved FITS as the standard format
@@
 
-FITS is already able to propose standard formatting for some data products
-quite common by now in planetary surface investigations.
+FITS is already capable of being used as a standard formatting for most data products
+commonly used in planetary surface investigations.
 In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
@@
 and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
-In addition, many PDS3 datasets in the Small Bodies Node include FITS files.
-To be efficiently used in planetary surface investigations, FITS metadata must
-be extended in order to take into account the size of the reference body.
-Planetary body shapes and orientations are well defined and standardized by
+In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed as FITS files with PDS3 labels.
+However, to be more efficiently used in planetary surface investigations, FITS metadata must
+be extended in order to take into account the size of the reference body and orientation as standardized by
 the IAU Working Group on Cartographic Coordinates and Rotational Elements
@@
 \label{ssec:geofits}
-When willing to use FITS format for planetary surfaces the issue of rigorous
-geographical representation must be addressed.
-For their mapping and analysis needs planetary researchers commonly use
-Geographical Information Systems (GIS).
-In general, GIS applications excel in data interoperability even
+To facilitate surface investigations, data taken of a planetary surface should be radiometrically calibrated and orthorectified (when topographic data is available) into a well-known map projection as required by Geographical Information Systems (GIS) applications.
+In general, GIS applications excel in planetary data set integration and interoperability even
 though some have historically been anchored to Earth's spatial description.
-Several of them are based on the Geospatial Data Abstraction
+Often, robust interoperability in GIS applications is achieved by using the Geospatial Data Abstraction
 Library\footnote{http://www.gdal.org} (GDAL), released by the Open Source
@@
 GDAL is essentially a format translation library written in C++ for geospatial raster
-and vector data and, fortunately, it also offer basic support for FITS format.
+and vector data and, fortunately, it also offers basic support for the FITS format.
 
@@
 terrestrial references.
-F
-or instance, there is no need for oblique projection definitions
+For instance, there is no need for oblique projection definitions
 \citep{calagreisenI,Snyder1987} as projection parameters
@@
 orientation, and reference surface.
+Table \ref{table:newkeys} summarize the keywords we propose to add to FITS metadata description.
 
@@
 \subsubsection{Visualization of Global and Projected Maps}
+\label{sssec:maps}
 High level products are meant to be used for visual inspection or GIS analysis.
@@
 As an example we have downloaded some global maps from USGS
-Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
+PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
 We used ds9 to display the FITS image and were able to visualize coordinates in meters
@@
 \label{sec:softdev}
+Basic conversion from and to FITS is already supported by GDAL.
+However, dictionaries for precisely translate geospatial metadata are not implemented yet.
+The conversions described in section \ref{sssec:maps} have been performed using python scripts
+available from the VESPA github repository\footnote{https://github.com/epn-vespa/fits2vrt}.
+Those python developments benefit from the rich and well maintained Astropy and GDAL python APIs.
+They will be implemented in planetary VO services with the goal to bridge between GIS and VO services.
+The next step to VO and GIS tool integration will be to directly implement metadata translation
+into GDAL C++ library.
+
+MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
+
 VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
-
-MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
 

### 2018-02-21 ess_marmo.tex
MATISSE description

--- 2018-02-15 ess_marmo.tex
@@
 \affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
-\affiliation{6}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
+\affiliation{6}{Space Science Data Center - ASI, Via del Politecnico snc, 00133, Rome, Italy}
 \affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
@@
 brings together several research communities (geologists, astronomers,
-physicists, geochemists, etc.). Planetary missions produce an impressively growing 
-amount of diverse data requiring an evolution from a mostly manual-based visual 
-analysis to a more automated and quantitative analysis.
-Interoperability, openness of data formats, and shared processing techniques are
-becoming a necessity to efficiently extract scientific information from the data and 
-to guarantee the reproducibility of the scientific results.
-Unfortunately, the technologies and data formats used by researchers for
-planetary surface studies and the field of astronomy 
-diverged as these related, but almost completely isolated domains evolved.  
-% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
-% 
-% > Unfortunately, the technologies 
-% > and data formats used by researchers for planetary surface studies and the field of astronomy 
-% > diverged as these related, but almost completely isolated domains evolved.  
-% 
-% Is this too strong a statement?
-% 
-% ^.
-In this paper we will will describe how a small addition to the Flexible Image Transport System (FITS) 
-standard, widely used in the astronomy investigations,  will allow FITS to be more easily used in planetary
-surface investigations , we will  show  how FITS metadata can easily be transformed in the PDS4 metadata
-archival model, and lastly provide example implementations.
-More than imposing a formal data model, a FITS description for planetary data aims
-to simplify sharing data across the planetary and astronomy domains and promoting
+physicists, geochemists, etc.).
+Planetary missions produce an impressively growing amount of heterogeneous
+data which mining evolves those days from mostly visual assessment to more
+automated quantitative analysis.
+Interoperability and openness of data formats and processing techniques are
+becoming a necessity, to avoid the risk of being unable to efficiently extract
+scientific information from the data, and to guarantee the reproducibility of
+the scientific results.
+This paper will describe how Flexible Image Transport System (FITS) can be
+used in planetary surface investigations, and how its metadata can easily be
+inserted in the PDS4 metadata distribution model.
+An evolution to the actual standard is proposed to adapt FITS to
+planetary body description.
+Some examples and implementations are then presented.
+More than imposing a formal data model, FITS description for those data aims
+to simplify the processing chain in planetary science, promoting
 interoperability from data acquisition to final visualization.
@@
 metadata distribution model.
-FITS is one of the standard formats implemented in the Virtual
+FITS is one of the standard implemented formats in the Virtual
 Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
@@
 processing tools.
-Those reasons make FITS format appropriate to automatically handle large amounts of raster data.
+Those reasons make FITS format appropriate for large amounts of raster data
+automatic processing.
 The goal of this approach is to make easier data mining and reprocessing
@@
 standards.
-The option to use FITS within the planetary surface domain could potentially
+The option to use FITS within the planetary domain could potentially
 homogenize methods from acquisition, to visualization, while
@@
 \item{some examples of planetary dataset in FITS;}
-\item{current developments for making FITS more usable to handle data
-of planetary surfaces.}
+\item{current developments for making FITS usable in planetary sciences.}
 \end{enumerate}
@@
 for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
-Therefore, FITS is one of the standard formats implemented in the Virtual
+Therefore, FITS is one of the standard implemented formats in the Virtual
 Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
 (VO).
-Finally, FITS data storage is compatible with the Planetary Data System (PDS)
+Finally, FITS format is compatible with the Planetary Data System (PDS)
 archiving specifications and supported by a large number of open libraries
@@
 to store inhomogeneous digital information (reflectance, calibration data,
-vector, table data, etc.) in the same file, each with corresponding metadata,
+vector, table data, etc.) in the same file, each with correspondent metadata,
 as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
@@
 \citep{crism} or OMEGA \citep{omega} instruments).
-FITS has been already chosen to distribute data from, e.g.,
-Hayabusa AMICA and NIRS cameras\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
-all Akatsuki cameras\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
-(except the Lightning and Airglow Camera),
-and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
-In addition, many PDS3 datasets in the Small Bodies Node include FITS files.
+FITS has been already chosen to distribute, e.g.,
+Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
+Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
+and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
 To be efficiently used in planetary surface investigations, FITS metadata must
@@
 FITS World Coordinate System
-representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
-way to describe spatial dependencies in FITS metadata.
-WCS simplifies the spatial coordinate description with respect to historical
+representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html}
+simplifies the spatial coordinate description with respect to historical
 terrestrial references.
-F
-or instance, there is no need for oblique projection definitions
+There is no need for oblique projection definitions
 \citep{calagreisenI,Snyder1987} as projection parameters
@@
 radius definition are easier to reproject from one system to another, than
-from or to a global spheroidal datum.
-As an example, the HiRISE team has chosen to distribute HiRISE projected
-products using Equirectangular projection (CAR) on a local surface, approximated
-by the sphere with radius equivalent to the local Mars spheroid radius.
-
-In addition, WCS resolves the ambiguity between east-handed and west-handed
-longitude systems, as the pixel conversion in world coordinates is set by an
+from or to a global spheroidal datum (see for example how the HiRISE team
+has chosen to distribute HiRISE images).
+In addition, there is no need for East or West positive longitude
+definitions as the pixel conversion in world coordinates is set by an
 oriented matrix.
@@
 Table \ref{table:proj} summarizes the FITS projection codes and their
-corresponding PDS4 projection names (see table A.1. in \citet{calagreisenI}
+correspondent PDS4 projection names (see table A.1. in \citet{calagreisenI}
 for details).
@@
 \caption{Projection names and codes already available in FITS suitable for
-planetary applications: the corresponding PDS4 terminology is
+planetary applications: their correspondent in PDS4 vocabulary are
 provided.}
@@
 
-\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
-\label{sssec:wcskeys}
-In this section WCS FITS keywords are inserted in the planetary surface context.
-When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
-standardization purposes.
-\begin{enumerate}
-\item{RADESYS.
-Name of the inertial reference frame with respect to which the coordinate
-system is defined (defaulted to \texttt{'ICRS'})}
-\item{EQUINOX.
-The value field contains the equinox in years for the celestial
-coordinate system in which positions are expressed (defaulted to \texttt{'J2000'})}
-\item{WCSAXES.
-Number of axes in the WCS description.
-The value of WCSAXES may exceed the number of raster pixel axes.}
-\item{WCSNAME.
-Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
-non-rotating, inertial).
-EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
-\item{CRPIXn.
-Location of a reference point along axis $n$, in units of the axis index.
-The reference point value do not need to lie within the actual data array.}
-\item{CTYPEn.
-Name of the coordinate represented by axis $n$, in “4–3” form: the first four characters
-specify the coordinate type, the fifth character is a '-' and the
-remaining three characters specify an algorithm code for computing the world coordinate
-value (\ref{table:proj}).
-For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
-are listed in Table \ref{table:ctype}.
-For linear coordinates we propose a similar scheme: xPX as Projected X, xPY as Projected Y,
-or yzPX and yzPY.
-Alternative representation of WCS are already possible in FITS, to describe images in degrees
-and in meters \citep{greisencalaII}.
-The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
-\item{CRVALn.
-Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
-\item{CDi\_j.
-The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
-world coordinates.}
-\item{PCi\_j.
-The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
-coordinates (to be used with CDELT keywords).}
-\item{CDELTn.
-The coordinate increment along axis $n$ (scaling component of the linear transformation
-matrix PCi\_j).
-In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
-attributes.
-In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
-\item{CUNITn.
-Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
-and to meters (m) for linear ones.}
-\item{CNAMEn.
-The description of a particular coordinate for an axis in a particular WCS (up to 68
-characters), while the WCSNAMEa keyword names the particular WCS as a whole.
-Its default value will be blank.}
-\end{enumerate}
+Table \ref{table:wcskeys} describes FITS WCS keywords and their recommended use in
+planetary surface context.
+
+\begin{sidewaystable}
+\label{table:wcskeys}
+\caption{WCS FITS keywords in the Planetary Surface context}
+\centering
+\begin{tabular}{l l l l l}
+\hline
+Keyword & Type & Definition & PDS4 & EPN-TAP \\
+\hline
+RADESYS	& string & Name of the inertial reference frame with respect & & \\
+& & to which the coordinate system is defined & & \\
+& & (defaulted to \texttt{'ICRS'}) & & \\
+EQUINOX & real & The value field contains the equinox in years for the celestial & & \\
+& & coordinate system in which positions are expressed & & \\
+& & (defaulted to \texttt{'J2000'}) & & \\
+WCSAXES & integer & Number of axes in the WCS description & & \\
+& & The value of WCSAXES may exceed the number of raster pixel axes & & \\
+WCSNAME & & & & \\
+CRPIXn & real & & \\
+CTYPEn & string & & \\
+CRVALn & real & & \\
+CDi\_j & real matrix & & \\
+PCi\_j & real matrix & & \\
+CDELTn & real & & \\
+CUNITn & string & & \\
+CNAMEn & string (up to 68 characters) & & \\
+\hline
+\end{tabular}
+\end{sidewaystable}
 
@@
 \label{table:ctype}
-\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
-Codes for planets are assumed in \citet{calagreisenI}. We propose to add codes for classes of targets,
-more or less corresponding to PDS4 Target type attribute and to EPN-TAP target\_class parameter.}
+\caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.}
 \centering
@@
 Neptune & NE \\
-Satellites (other than the Moon) & ST \\
-Asteroids & AS \\
-Dwarf Planets & DW \\
-Comets & CO \\
+Planetary bodies & B \\
 \hline
@@
 the associated data array.
-PDS4 uses the attribute \texttt{data\_type}: corresponding 
+PDS4 uses the attribute \texttt{data\_type}: correspondent 
 values depend on the associated object class.}
@@
 elements along axis \texttt{n} of a data array.
-The \texttt{elements} attribute of the PDS4 \texttt{Axis\_Array} class provides
+The \texttt{elements} attribute of the \texttt{Axis\_Array} class provides
 the count of the number of elements along an array.}
@@
 The units of all FITS header keyword values should conform with the recommendations
-in the IAU Style Manual\footnote{https://www.iau.org/publications/proceedings\_rules/units/}.
+in the IAU Style Manual.
 For angular measurements given as floating point values, degrees are the recommended
@@
 For all classes containing Elements, PDS4 defines a \texttt{scaling\_factor}
-attribute.
-Both FITS and PDS4 default this value to $1$.}
+attribute. Both FITS and PDS4 defaulted this value to $1$.}
 
@@
 For all classes containing Elements, PDS4 defines a \texttt{value\_offset}
-attribute. Both FITS and PDS4 default this value to $0$.}
+attribute. Both FITS and PDS4 defaulted this value to $0$.}
 
 \item{\texttt{BLANK}.
-This keyword must be used only with integer data.
+This keyword must be used only in integer data.
 It contains an integer that specifies the value that is used to
@@
 It is assumed to be expressed in UTC but other systems are accepted as long as they
-are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
+are specified in the ISO string. 
 PDS4 provides a \texttt{Time\_Coordinates} class with attributes
 \textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
-stop date and time appropriate to the product being labeled, always in UTC.
+stop date and time appropriate to the product being labeled, always in UT
+coordinates.
 The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
-they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}
+they must always be formatted in UTC and Julian Days.}
+% * <chiara.marmo@u-psud.fr> 2018-02-05T10:20:13.335Z:
+% 
+% > Julian Days
+% From https://voparis-confluence.obspm.fr/display/VES/EPN-TAP+v2+parameter+description I see that ISO format is still in the race.... Maybe it would be better to homogenize PDS4 EPN-TAP and FITS?
+% 
+% ^.
 
@@
 It contains a character string identifying the instrument used to acquire the data.
-PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally 
-corresponds to this FITS keyword.
+PDS4 provides an \texttt{Instrument} class which attribute \texttt{Name} ideally corresponds
+to this FITS keywords.
 EPN-TAP provides with an \texttt{instrument\_name} parameter.}
@@
 ground-based observations.
-In the EPN-TAP scheme this is introduced by the \texttt{instrument\_host\_name}
-parameter\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}.}
+In the EPN-TAP scheme this will defined the \texttt{instrument\_host\_name} parameter.
+\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Spacecraft}}
 
 \item{\texttt{OBJECT}.
-It contains a character string giving a name for the observed
-object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
+It contains a character string giving a name for the observed object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
 The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to this
@@
 
-\item{\texttt{ORIGIN}.
-It contains string identifying the organization or institution responsible for
-creating the FITS file.}
-
+%
+%\item{\texttt{ORIGIN}.
+%It contains string identifying the organization or institution responsible for
+%creating the FITS file.
+%}
 %
@@
 \item{\texttt{REFERENC}.
-It contains a character string citing the bibliographic reference where the
-data are first described.
-It is recommended that either the Astrophysics Data System bibliographic
-databases\footnote{http://adswww.harvard.edu/}
+It contains a character string citing a reference where the data are published.
+It is recommended that either the Astrophysics Data System bibliographic databases\footnote{http://adswww.harvard.edu/}
 bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
 be used.
-PDS4 \texttt{External\_Reference} class and EPN-TAP \texttt{bib\_reference} have
-exactly the same meaning and use than the FITS keyword.}
+PDS4 \texttt{External\_Reference} class 
+EPN-TAP \texttt{bib\_reference} has exactly the same meaning and use than the FITS keyword.}
 
@@
 format, as GeoTIFF make them GIS ready.
-Given some new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
+Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
 and GeoTIFF are completely interchangeable.
@@
 We used ds9 to display the FITS image and were able to visualize coordinates in meters
-as in QGIS, a popular open source GIS software, but also in planetary latitude and longitude
-thanks to the alternate coordinate
+as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
 description (see figure \ref{fig:ds9qgis}, left side).
 \begin{figure}[ht!]
-\centerline{\includegraphics[height=17pc]{ess_marmo2}\includegraphics[height=17pc]{ess_marmo1}}
+\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
 \caption{left: Mercury South Pole FITS image displayed in ds9;
@@
 
-Assuming projected products are distributed in FITS format, a quick method to add GIS support for
+Assuming projected products distributed in FITS format, a quick method to add GIS support for
 FITS images is to use a GDAL-supported detached header.
@@
 (WKT) projection string.
-As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual Header
+As a result, the FITS raster loaded into QGIS thanks to its correspondent GDAL Virtual Header
 is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).
 
-In order to load FITS files in GIS software, it is recommended to have
+In order to straightforward load FITS files in GIS software, it is recommended to have
 body radii and linear coordinates filled in.
 Information about invalid values (NoData values) and minimum and maximum values are useful
-to correctly display the image dynamics.
+for correct display.
 
@@
 As stressed above, having planetary data distributed in FITS at any processing level is
-a way to avoid multiple conversions in the processing chain and to simplify reprocessing.
+a way to avoid multiple conversions in the processing chain and simplify reprocessing.
 In planetary science a first spatial calibrated product often consists in providing coordinates
@@
 
-As an example of non projected planetary data we have converted in FITS format some data
-acquired by the imaging spectrometer OMEGA on-board of Mars Express.
+As an example of non projected planetary data we have converted in FITS format data
+acquired by the spectroimager OMEGA on-board of Mars Express.
 OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
@@
 
-MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
+MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from solar system exploration missions.\\
+Up to now it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars, dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and can be integrated with other datasets thanks to its modular structure.\\
+Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is formatted as FITS, projected by interpolating the original observations to a grid regularly spaced in planetocentric coordinates (i.e., latitude and longitude).\\
+We used the planetary FITS standard described in this work in order to make the output file compliant with the great part of the FITS readers, DS9 and JS9 above all.
+The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt}, whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.\\
+Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based environment.\\
 

### 2018-02-21 ess_marmo.tex
Update on Overleaf.

--- 2018-02-16 ess_marmo.tex
@@
 standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
-surface investigations. We will  show  how FITS metadata can easily be transformed in the PDS4 metadata
+surface investigations. We will also show  how FITS metadata can easily be transformed in the PDS4 metadata
 archival model and lastly provide example implementations.
@@
 homogenize methods from acquisition, to visualization, while
-giving chances to optimize data processing across the planetary and astronomy domains.
+allowing for optimized data processing across the planetary and astronomy domains.
 
@@
 \subsection{FITS Format}
-FITS is an open digital standard, defined for data acquisition and archiving telescope data
-by astronomical observatories since the late 1970's.
+FITS is an open digital standard, created in the late 1970's for data acquisition and archiving telescope data
+by astronomical observatories.
 The International Astronomical Union approved FITS as the standard format
@@
 \label{ssec:geofits}
-To facilitate surface investigations, data taken of a planetary surface should be radiometrically calibrated and orthorectified (when topographic data is available) into a well-known map projection as required by Geographical Information Systems (GIS) applications.
+To facilitate surface investigations, data taken of a planetary surface should be radiometrically 
+calibrated and orthorectified (when topographic data is available) into a well-known map projection 
+as required by Geographical Information Systems (GIS) applications.
 In general, GIS applications excel in planetary data set integration and interoperability even
@@
 
-In addition, WCS resolves the ambiguity between east-handed and west-handed
+In addition, WCS resolves the ambiguity between east-positive and west-posited
 longitude systems, as the pixel conversion in world coordinates is set by an
@@
 coordinate system using several alternative WCS definitions (meters, degrees,
-East and West longitude, etc.)\citep{greisencalaII}.
+east and west longitude, etc.)\citep{greisencalaII}.
 
@@
 \begin{table}
-\label{table:proj}
 \caption{Projection names and codes already available in FITS suitable for
@@
 \end{tabular}
+\label{table:proj}
 \end{table}
@@
 \item{CTYPEn.
-Name of the coordinate represented by axis $n$, in “4–3” form: the first four characters
+Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
 specify the coordinate type, the fifth character is a '-' and the
@@
 \begin{table}
-\label{table:ctype}
 \caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
@@
 \end{tabular}
+\label{table:ctype}
 \end{table}
@@
 \begin{sidewaystable}
-\label{table:newkeys}
 \caption{Proposed keywords are related to 3D surface shape,
@@
 \end{tabular}
+\label{table:newkeys}
 \end{sidewaystable}
@@
 
-MATISSE\footnote{http://tools.asdc.asi.it/matisse.jsp} \citep{ZINZI2016}
+MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
+Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space
+Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from
+solar system exploration missions.
+Up to now it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
+dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
+remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and
+can be integrated with other datasets thanks to its modular structure.
+Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is
+formatted as FITS, projected by interpolating the original observations to a grid regularly spaced
+in planetocentric coordinates (i.e., latitude and longitude).
+We used the planetary FITS standard described in this work in order to make the output file compliant
+with the great part of the FITS readers, DS9 and JS9 above all.
+The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
+after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
+whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.
+Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end
+of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the
+tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based
+environment.
 

### 2018-02-22 ess_marmo.tex
Update on Overleaf.

@@
 The option to use FITS within the planetary surface domain could help to
-homogenize methods from acquisition, to visualization, while
+homogenize methods from acquisition to visualization, while
 allowing for optimized data processing across the planetary and astronomy domains.
@@
 \subsection{FITS Format}
-FITS is an open digital standard, created in the late 1970's for data acquisition and archiving telescope data
-by astronomical observatories.
+FITS is an open digital standard, created in the late 1970's for data acquisition, transfer and
+archiving of telescope data by astronomical observatories.
 The International Astronomical Union approved FITS as the standard format
@@
 keyword, containing the projection code and the coordinate system.
+% * <chiara.marmo@u-psud.fr> 2018-02-22T21:31:39.056Z:
+% 
+% Example or scheme needed in order to better understand the coordinate transformation flow
+% 
+% ^.
+
 The coordinate system must be defined with respect to a specific inertial reference frame.
@@
 Multiple coordinate system descriptions are allowed in the same header: this is a valuable
-property for planetary surface imagery, as angular ('deg') and linear ('m') description of
+property for planetary surface imagery, as angular (`deg') and linear (`m') description of
 the surface are often needed at the same time.
@@
 Name of the inertial reference frame with respect to which the coordinate
-system is defined (defaulted to \texttt{'ICRS'})}
+system is defined (defaulted to \texttt{`ICRS'})}
 \item{EQUINOX.
 The value field contains the equinox in years for the celestial
-coordinate system in which positions are expressed (defaulted to \texttt{'J2000'})}
+coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
 \item{WCSAXES.
@@
 Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
-specify the coordinate type, the fifth character is a '-' and the
+specify the coordinate type, the fifth character is a `-' and the
 remaining three characters specify an algorithm code for computing the world coordinate
@@
 & & & defines the approximate shape of a target body used in projection. \\
-& & & 'A' is usually in the equatorial plane. Always in meters.  \\
+& & & `A' is usually in the equatorial plane. Always in meters.  \\
 B\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the intermediate axis of \\
 & & & the ellipsoid that defines the approximate shape of a target body used in projection. \\
-& & & 'B' is usually in the equatorial plane. Always in meters.   \\
+& & & `B' is usually in the equatorial plane. Always in meters.   \\
 C\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the semi-minor axis of the \\
 & & & ellipsoid that defines the approximate shape of a target body used in projection. \\
-& & & 'C' is normal to the plane defined by 'A' and 'B'. Always in meters. \\
+& & & `C' is normal to the plane defined by `A' and `B'. Always in meters. \\
 OGCCODE	& string & reserved, optional & Value field contains a string describing the standard OGC code \\
@@
 For angular measurements given as floating point values, degrees are the recommended
-units (with the units, if specified, given as 'deg').
+units (with the units, if specified, given as `deg').
 PDS4 specifies in the class Element (of a vector, array, or any other class

### 2018-02-26 ess_marmo.tex
Update on Overleaf.

@@
 Number of axes in the WCS description.
-The value of WCSAXES may exceed the number of raster pixel axes.}
+The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
+generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
+additional time dependency is added to the coordinate description.}
 \item{WCSNAME.
@@
 For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
-are listed in Table \ref{table:ctype}.
-For linear coordinates we propose a similar scheme: xPX as Projected X, xPY as Projected Y,
-or yzPX and yzPY.
+correspond to the target body. We propose to standardize those codes as listed in Table
+\ref{table:ctype}.
+For projected coordinates in meters we propose a similar scheme: xPX as Projected X, xPY as
+Projected Y, or yzPX and yzPY.
 Alternative representation of WCS are already possible in FITS, to describe images in degrees
@@
 \label{sec:softdev}
+
+\subsection{FITS and GDAL}
 Basic conversion from and to FITS is already supported by GDAL.
@@
 
+\subsection{MATISSE}
 MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
@@
 
+\subsection{Aladin}
+Aladin is both a sky atlas and a portal to access data -- mainly images and
+catalogues -- available through the VO.
+It is able to load FITS files and to display images taking into account various
+WCS projections.
+Like DS9, Aladin is popular and broadly used in the part of the astronomical
+community manipulating images.
+So far it is also the reference generator and visualizer of Hierarchical
+Progressive Survey (HiPS).
+HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
+by the International Virtual Observatory Alliance (IVOA).
+It describes how to access to sky survey data stored at various spatial
+resolutions, offering a progressive view of possibly very large datasets.
+The purpose is to allows users to explore all-sky surveys without having to
+explicitly perform any query: the most important information is visible at low
+spatial resolutions while finer details appear by zooming in.
+In addition, an important property of HiPS is to preserve as much as possible
+the scientific content of the original dataset (few spatial distortions,
+flux conservation).
+
+In the framework of the VESPA project, the idea of using Aladin to display
+planetary data, especially planetary HiPS, has emerged.
+As a general tool, not dedicated to planetology, Aladin does not support PDS
+files.
+HiPS are usually generated from a collection of WCS tagged FITS files.
+%As an alternative, Aladin can use JPEG files together with additional metadata
+%files.
+%A metadata file -- one by JPEG file -- basically contains in a text format the
+%WCS part of a FITS header.
+
+Aladin\footnote{http://aladin.u-strasbg.fr/} and its HiPS generator have been updated to support
+several planetary conventions.
+First, while the sky is displayed from the inside of the unit sphere, planetary
+surfaces are observed -- and must be displayed -- from the exterior.
+It means that the direction of the east on the horizontal axis has to be
+inverted when displaying planetary data.
+Then, planetary missions usually use different projections when providing
+images of planets polar caps and equatorial region.
+Often the former are in the stereographic (STG) projection while the later
+is in the Plate Carr\'ee (CAR) projection. 
+This specificity has been taken into account into the HiPS generator.
+Also, in the future, Aladin will be able to recognize the FITS keywords defined
+in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
+For example, it will be able to take into account a planet's radii to compute
+distances on the surface of the body.
+Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
+barely used in extrasolar astronomy and are not supported in Aladin yet.
+An effort has been started to revamp and update the code of projections in
+Aladin.
+We expect all projections to be available in a forthcoming release.
+
 VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
-
-Aladin\footnote{http://aladin.u-strasbg.fr/}
 
@@
 
+\appendix
+\section{Planetary Header example}
+%An example of planetary header goes here.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-02-26 ess_marmo.tex
Update on Overleaf.

@@
 Flagstaff, AZ, USA}
-\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
-CNRS, Sorbonne Universit\'es, UPMC Univ. Paris 06, Univ. Paris Diderot,
-Sorbonne Paris Cit\'e, Meudon, France}
+\affiliation{3}{LESIA, Observatoire de Paris, Universit\'e PSL, CNRS, Sorbonne Universit\'e,
+Univ. Paris Diderot, Sorbonne Paris Cit\'e, 5 place Jules Janssen, 92195 Meudon, France}
 \affiliation{4}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
@@
 to simplify sharing data across the planetary and astronomy domains and promoting
-interoperability from data acquisition to final visualization.
+interoperability from raw data formatting to final visualization.
 \end{abstract}
@@
 The option to use FITS within the planetary surface domain could help to
-homogenize methods from acquisition to visualization, while
+homogenize methods from raw formatting to visualization, while
 allowing for optimized data processing across the planetary and astronomy domains.
@@
 archiving of telescope data by astronomical observatories.
+It had been used data exchange and archiving from several orbital telescopes.
 The International Astronomical Union approved FITS as the standard format
@@
 (VO).
-Finally, FITS data storage is compatible with the Planetary Data System (PDS)
-archiving specifications and supported by a large number of open libraries
+FITS data storage is compatible with the Planetary Data System (PDS)
+archiving specifications so that FITS files can be embedded in PDS datasets.
+Finally, it is supported by a large number of open libraries
 and software tools.
@@
 
-In addition, WCS resolves the ambiguity between east-positive and west-posited
+In addition, WCS resolves the ambiguity between east-positive and west-positive
 longitude systems, as the pixel conversion in world coordinates is set by an
@@
 The coordinate system must be defined with respect to a specific inertial reference frame.
-
+% * <chiara.marmo@u-psud.fr> 2018-02-26T11:33:56.098Z:
+% 
+% Please be more specific about the inertial frame
+% 
+% ^.
 Multiple coordinate system descriptions are allowed in the same header: this is a valuable
@@
 In this section WCS FITS keywords are inserted in the planetary surface context.
-When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
+When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving and distribution
 standardization purposes.
@@
 solar system exploration missions.
-Up to now it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
+Currently, it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
 dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
@@
 in planetocentric coordinates (i.e., latitude and longitude).
-We used the planetary FITS standard described in this work in order to make the output file compliant
-with the great part of the FITS readers, DS9 and JS9 above all.
+We used the planetary FITS standard described in this work in order to make the output file
+compliant with most FITS readers, especially DS9 and JS9.
 The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
@@
 
-Aladin\footnote{http://aladin.u-strasbg.fr/} and its HiPS generator have been updated to support
-several planetary conventions.
+Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
+been updated to support several planetary conventions.
 First, while the sky is displayed from the inside of the unit sphere, planetary
@@
 of astronomy experience. In addition to that WCS projection scheme is simpler than the
-historical one used in planetary applications. 
+historical Earth-based one used in planetary applications. 
 Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.

### 2018-02-27 ess_marmo.tex

@@
 standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
-surface investigations. We will also show  how FITS metadata can easily be transformed in the PDS4 metadata
-archival model and lastly provide example implementations.
+surface investigations. We will also show  how FITS metadata can easily be transformed
+in the PDS4 metadata archival model and lastly provide example implementations.
 More than imposing a formal data model, a FITS description for planetary data aims
@@
 \subsection{FITS Format}
-FITS is an open digital standard, created in the late 1970's for data acquisition, transfer and
-archiving of telescope data by astronomical observatories.
-It had been used data exchange and archiving from several orbital telescopes.
+FITS is an open digital standard, created in the late 1970's for data acquisition,
+transfer and archiving of telescope data by astronomical observatories.
+It had been used for data exchange and archiving from several orbital telescopes.
 The International Astronomical Union approved FITS as the standard format
@@
 and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
-In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed as FITS files with PDS3 labels.
-However, to be more efficiently used in planetary surface investigations, FITS metadata must
-be extended in order to take into account the size of the reference body and orientation as standardized by
-the IAU Working Group on Cartographic Coordinates and Rotational Elements
-(WGCCRE).
+In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed
+as FITS files with PDS3 labels.
+However, to be more efficiently used in planetary surface investigations, FITS metadata
+must be extended in order to take into account the size of the reference body and
+orientation as standardized by the IAU Working Group on Cartographic Coordinates and
+Rotational Elements (WGCCRE).
 This group reports triennially \citep{wgccre} on the preferred rotation rate,
@@
 \label{ssec:geofits}
-To facilitate surface investigations, data taken of a planetary surface should be radiometrically 
-calibrated and orthorectified (when topographic data is available) into a well-known map projection 
-as required by Geographical Information Systems (GIS) applications.
-In general, GIS applications excel in planetary data set integration and interoperability even
-though some have historically been anchored to Earth's spatial description.
-Often, robust interoperability in GIS applications is achieved by using the Geospatial Data Abstraction
-Library\footnote{http://www.gdal.org} (GDAL), released by the Open Source
-Geospatial Foundation.
+To facilitate surface investigations, data taken of a planetary surface should be
+radiometrically calibrated and orthorectified (when topographic data is available)
+into a well-known map projection as required by Geographical Information Systems (GIS)
+applications.
+In general, GIS applications excel in planetary data set integration and 
+interoperability even though some have historically been anchored to Earth's spatial
+description.
+Often, robust interoperability in GIS applications is achieved by using the Geospatial
+Data Abstraction Library\footnote{http://www.gdal.org} (GDAL), released by the Open
+Source Geospatial Foundation.
 GDAL is essentially a format translation library written in C++ for geospatial raster
@@
 We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
-\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
+\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator: those three projection are not used for archival or analysis.}
 \end{tabular}
@@
 % ^.
-Multiple coordinate system descriptions are allowed in the same header: this is a valuable
-property for planetary surface imagery, as angular (`deg') and linear (`m') description of
-the surface are often needed at the same time.
+Multiple coordinate system descriptions are allowed in the same header: this is a
+valuable property for planetary surface imagery, as angular (`deg') and linear (`m')
+description of the surface are often needed at the same time.
 Further below 3D keyword will be proposed for FITS metadata, in order to standardize
@@
 In this section WCS FITS keywords are inserted in the planetary surface context.
-When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving and distribution
-standardization purposes.
+When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
+and distribution standardization purposes.
 \begin{enumerate}
@@
 \item{CTYPEn.
-Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
-specify the coordinate type, the fifth character is a `-' and the
+Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four
+characters specify the coordinate type, the fifth character is a `-' and the
 remaining three characters specify an algorithm code for computing the world coordinate
 value (\ref{table:proj}).
-For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
-correspond to the target body. We propose to standardize those codes as listed in Table
-\ref{table:ctype}.
-For projected coordinates in meters we propose a similar scheme: xPX as Projected X, xPY as
-Projected Y, or yzPX and yzPY.
-Alternative representation of WCS are already possible in FITS, to describe images in degrees
-and in meters \citep{greisencalaII}.
-The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
+For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG,
+where x or yz correspond to the target body.
+We propose to standardize those codes as listed in Table \ref{table:ctype}.
+For projected coordinates in meters we propose a similar scheme: xPX as Projected X,
+xPY as Projected Y, or yzPX and yzPY.
+Alternative representation of WCS are already possible in FITS, to describe images in
+degrees and in meters \citep{greisencalaII}.
+The name of the alternative axis will be specified using the alternative CNAMEn
+keyword.}
 \item{CRVALn.
@@
 \item{CDi\_j.
-The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
-world coordinates.}
+The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the
+intermediate world coordinates.}
 \item{PCi\_j.
-The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
-coordinates (to be used with CDELT keywords).}
+The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate
+pixel coordinates (to be used with CDELT keywords).}
 \item{CDELTn.
@@
 matrix PCi\_j).
-In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
-attributes.
+In PDS4 they correspond to the \textit{pixel\_resolution\_x} and
+\texttt{pixel\_resolution\_y} attributes.
 In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
 \item{CUNITn.
-Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
-and to meters (m) for linear ones.}
+Units of the coordinate system, not mandatory, defaulted to degrees for angular
+coordinates and to meters (m) for linear ones.}
 \item{CNAMEn.
@@
 \caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
-Codes for planets are assumed in \citet{calagreisenI}. We propose to add codes for classes of targets,
-more or less corresponding to PDS4 Target type attribute and to EPN-TAP target\_class parameter.}
+Codes for planets are assumed in \citet{calagreisenI}.
+We propose to add codes for classes of targets,
+more or less corresponding to PDS4 Target type attribute and to EPN-TAP
+target\_class parameter.}
 \centering
@@
 On floating point data the IEEE NaN must be used to represent undefined values.  
-\texttt{PDS4 Special\_Constants?}}
+It is comparable with the \texttt{invalid\_constant} or \texttt{missing\_constant}
+in PDS4 \texttt{Special Constants} class.}
 
 \item{\texttt{DATAMAX}.
-It always contains a floating-point number, giving the maximum valid physical value represented
-by the array, excluding any IEEE special values.
+It always contains a floating-point number, giving the maximum valid physical value
+represented by the array, excluding any IEEE special values.
 PDS4 defines a \texttt{valid\_maximum} attribute in the \texttt{Special Constants} class
@@
 \item{\texttt{DATAMIN}.
-It always contains a floating-point number, giving the minimum valid physical value represented
-by the array, excluding any IEEE special values.
+It always contains a floating-point number, giving the minimum valid physical value
+represented by the array, excluding any IEEE special values.
 PDS4 defines a \texttt{valid\_minimum} attribute in the \texttt{Special Constants} class
@@
 stop date and time appropriate to the product being labeled, always in UTC.
-The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP scheme,
-they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}
+The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP
+scheme, they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}
 
@@
 The PDS4 \texttt{Instrument\_Host} class provides the description of the physical object
-mounting the instrument. Its \texttt{Name} attribute should correspond to the FITS keyword.
+mounting the instrument.
+Its \texttt{Name} attribute should correspond to the FITS keyword.
 PDS4 also provides for the definition of a \texttt{Telescope} class, in case of
@@
 object\footnote{https://naif.jpl.nasa.gov/pub/naif/toolkit\_docs/FORTRAN/req/naif\_ids.html\#Planets\%20and\%20Satellites}.
-The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to this
-FITS keyword.
+The PDS4 \texttt{Target} class contains the \texttt{name} attribute corresponding to
+this FITS keyword.
 EPN-TAP standard also provides with a \texttt{target\_name} parameter.
-Notice that PDS4 and EPN-TAP also ask for a \texttt{target\_type} attribute, which must be
-specified from an enumerated list.}
+Notice that PDS4 and EPN-TAP also ask for a \texttt{target\_type} attribute,
+which must be specified from an enumerated list.}
 
@@
 High level products are meant to be used for visual inspection or GIS analysis.
-They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
+They are distributed as projected rasters, often in
+GeoTIFF\footnote{http://geotiff.osgeo.org/}
 format, as GeoTIFF make them GIS ready.
-Given some new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
-and GeoTIFF are completely interchangeable.
+Given some new keywords introduced in FITS and WCS definitions adapted to planetary
+surface, FITS and GeoTIFF are completely interchangeable.
 As an example we have downloaded some global maps from USGS
@@
 We used ds9 to display the FITS image and were able to visualize coordinates in meters
-as in QGIS, a popular open source GIS software, but also in planetary latitude and longitude
-thanks to the alternate coordinate
+as in QGIS, a popular open source GIS software, but also in planetary latitude and
+longitude thanks to the alternate coordinate
 description (see figure \ref{fig:ds9qgis}, left side).
@@
 are perfectly superposed.
-Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
-Laboratory, Carnegie Science}
+Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied
+Physics Laboratory, Carnegie Science}
 \label{fig:ds9qgis}
@@
 
-Assuming projected products are distributed in FITS format, a quick method to add GIS support for
-FITS images is to use a GDAL-supported detached header.
+Assuming projected products are distributed in FITS format, a quick method to add GIS
+support for FITS images is to use a GDAL-supported detached header.
 GDAL Virtual Header files simply describe the internal structure of a FITS image, but
@@
 (WKT) projection string.
-As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual Header
-is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis}, right side).
+As a result, the FITS raster loaded into QGIS thanks to its corresponding GDAL Virtual
+Header is indistinguishable from the GeoTIFF layer (see figure \ref{fig:ds9qgis},
+right side).
 
@@
 body radii and linear coordinates filled in.
-Information about invalid values (NoData values) and minimum and maximum values are useful
-to correctly display the image dynamics.
+Information about invalid values (NoData values) and minimum and maximum values are
+useful to correctly display the image dynamics.
 
@@
 As stressed above, having planetary data distributed in FITS at any processing level is
-a way to avoid multiple conversions in the processing chain and to simplify reprocessing.
-In planetary science spatial information is first provided for specific points over the detector
-(the four corners of the detector, the center of each pixel, one or more pixel corners, etc.).
-To support in FITS metadata such look-up table coordinate representation the \texttt{TAB}
-algorithm has been defined in \citet{wcsspectral}.
-In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array, an indexing vector
-can be used to address coordinate array elements.
-Coordinates can then be sampled more or less coarsely depending on the behavior of the spatial
-reconstruction.
-Also, when the field is only partially covered by the planetary surface, coordinate array
-dimensions can be significantly reduced. 
+a way to avoid multiple conversions in the processing chain and to simplify
+reprocessing.
+In planetary science spatial information is first provided for specific points over
+the detector (the four corners of the detector, the center of each pixel,
+one or more pixel corners, etc.).
+To support in FITS metadata such look-up table coordinate representation the
+\texttt{TAB} algorithm has been defined in \citet{wcsspectral}.
+In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array,
+an indexing vector can be used to address coordinate array elements.
+Coordinates can then be sampled more or less coarsely depending on the behavior of the
+spatial reconstruction.
+Also, when the field is only partially covered by the planetary surface, coordinate
+array dimensions can be significantly reduced. 
 TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
@@
 acquired by the imaging spectrometer OMEGA on-board of Mars Express.
-OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA and PDS archives.
+OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA and PDS
+archives.
 Their structure is described in the OMEGA Experiment Archive 
@@
 \centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo3}\hspace{2pc}\includegraphics[height=17pc]{ess_marmo5}}
-\caption{left: OMEGA data structure (from OMEGA EAICD), science data and geometry description are
-distributed in different files; center: FITS Multi Digital Object scheme; as an example, we stored
-the spectral data in the first cube extension, the coordinate look-up table in the second
-extension, and other geometry parameters (incidence and emission angle, etc.) as images in the
+\caption{left: OMEGA data structure (from OMEGA EAICD), science data and geometry
+description are distributed in different files; center: FITS Multi Digital Object
+scheme; as an example, we stored the spectral data in the first cube extension,
+the coordinate look-up table in the second extension, and other geometry parameters
+(incidence and emission angle, etc.) as images in the
 following extensions; right: BSQ cube visualization with ds9.}
@@
 TAB description without indexing.
-In addition, as geometrical description depends on the observation filter, this allows us to
-distribute a file for each filter, containing science and geometry data.
-The spectral cubes have been converted from Band Interleaved per Line (BIL) to Band Sequential
-(BSQ) in order to make them easily understandable when visualized using standard FITS viewer
-(as ds9 in figure \ref{fig:omegascheme})\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/orb0413\_1.IR1.fits.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/omega2fits.py}.
+In addition, as geometrical description depends on the observation filter, this allows
+us to distribute a file for each filter, containing science and geometry data.
+The spectral cubes have been converted from Band Interleaved per Line (BIL) to Band
+Sequential (BSQ) in order to make them easily understandable when visualized using
+standard FITS viewer (as ds9 in figure \ref{fig:omegascheme})\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/orb0413\_1.IR1.fits.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/omega2fits.py}.
 
@@
 distribution, in the \texttt{wcslib-tools} in Debian) had been run on the output file,
-validating the WCS structure and checking the ability to convert to (\texttt{wcsware -w})
-and from (\texttt{wcsware -x}) pixel and planetary surface coordinates.
+validating the WCS structure and checking the ability to convert to
+(\texttt{wcsware -w}) and from (\texttt{wcsware -x}) pixel and planetary surface
+coordinates.
 
@@
 Akatsuki imaging data are already distributed in FITS format.
-Geometry information is distributed in two different cubes of data, containing respectively 
-values at the center of the pixel and at the lower left corner of the pixel.
-Tabular coordinate representation allows us to gather geometry in one table together with
-the described raster.
-In addition to that, in Akatsuki images Venus is sometime far from completely filling the field:
-tabular representation avoids to store unnecessary data. 
-In that case, coordinate indexes are necessary to identify which pixels contains the spatial
-information\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/ir1\_20160415\_070351\_097\_l2b\_v10\_out.fit.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/akatsukil2b.py}.
-
-Again, we used \texttt{wcsware} to validate the WCS structure and to test pixel to world /
-world to pixel coordinate conversion.
-Even though TAB projection is not fully implemented in FITS visualization tools, this representation
-has the advantage to store the detector geometrical information together with physical
-quantities measured by the detector itself in a standardized way.
+Geometry information is distributed in two different cubes of data, containing
+respectively values at the center of the pixel and at the lower left corner of the
+pixel.
+Tabular coordinate representation allows us to gather geometry in one table together
+with the described raster.
+In addition to that, in Akatsuki images Venus is sometime far from completely filling
+the field: tabular representation avoids to store unnecessary data. 
+In that case, coordinate indexes are necessary to identify which pixels contains the
+spatial information\footnote{The output result can be downloaded at the address https://github.com/cmarmo/convertofits/releases/download/v1.0-beta/ir1\_20160415\_070351\_097\_l2b\_v10\_out.fit.gz together with the python script used for conversion https://github.com/cmarmo/convertofits/blob/master/akatsukil2b.py}.
+
+Again, we used \texttt{wcsware} to validate the WCS structure and to test pixel to
+world / world to pixel coordinate conversion.
+Even though TAB projection is not fully implemented in FITS visualization tools,
+this representation has the advantage to store the detector geometrical information
+together with physical quantities measured by the detector itself in a standardized way.
 
@@
 Basic conversion from and to FITS is already supported by GDAL.
-However, dictionaries for precisely translate geospatial metadata are not implemented yet.
-The conversions described in section \ref{sssec:maps} have been performed using python scripts
-available from the VESPA github repository\footnote{https://github.com/epn-vespa/fits2vrt}.
-Those python developments benefit from the rich and well maintained Astropy and GDAL python APIs.
-They will be implemented in planetary VO services with the goal to bridge between GIS and VO services.
-The next step to VO and GIS tool integration will be to directly implement metadata translation
-into GDAL C++ library.
+However, dictionaries for precisely translate geospatial metadata are not implemented
+yet.
+The conversions described in section \ref{sssec:maps} have been performed using python
+scripts available from the VESPA github repository\footnote{https://github.com/epn-vespa/fits2vrt}.
+Those python developments benefit from the rich and well maintained Astropy and GDAL
+python APIs.
+They will be implemented in planetary VO services with the goal to bridge between GIS 
+and VO services.
+The next step to VO and GIS tool integration will be to directly implement metadata 
+translation into GDAL C++ library.
 
 \subsection{MATISSE}
-MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
-Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space
-Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from
-solar system exploration missions.
-Currently, it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
-dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
-remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and
+MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
+for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
+designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
+access, visualize and analyze data from solar system exploration missions.
+Currently, it provides access to public data from five different targets (i.e., Mercury,
+Venus, Mars, dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local
+storage or by querying remote dataset, such as VESPA, Planet Server and NASA JPL and
 can be integrated with other datasets thanks to its modular structure.
-Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is
-formatted as FITS, projected by interpolating the original observations to a grid regularly spaced
+One of th MATISSE 2D outputs is formatted as FITS, projected by interpolating the
+original observations to a grid regularly spaced
 in planetocentric coordinates (i.e., latitude and longitude).
-We used the planetary FITS standard described in this work in order to make the output file
-compliant with most FITS readers, especially DS9 and JS9.
-The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
+We used the planetary FITS standard described in this work in order to make the output
+file compliant with most FITS readers, especially DS9 and JS9.
+The files generated by MATISSE could be also opened with standard GIS software (e.g.,
+QGIS, ArcGIS) 
 after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
-whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.
-Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end
-of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the
-tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based
-environment.
+whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
+projection.
+Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
+online by the end of 2018) we are planning to make use of JS9 as the standard reader for
+the 2D online output of the tool, thus allowing users to perform advanced analysis in
+both 2D and 3D directly inside a web-based environment.
 
@@
 resolutions, offering a progressive view of possibly very large datasets.
-The purpose is to allows users to explore all-sky surveys without having to
-explicitly perform any query: the most important information is visible at low
-spatial resolutions while finer details appear by zooming in.
-In addition, an important property of HiPS is to preserve as much as possible
-the scientific content of the original dataset (few spatial distortions,
-flux conservation).
-
-In the framework of the VESPA project, the idea of using Aladin to display
-planetary data, especially planetary HiPS, has emerged.
-As a general tool, not dedicated to planetology, Aladin does not support PDS
-files.
+%The purpose is to allows users to explore all-sky surveys without having to
+%explicitly perform any query: the most important information is visible at low
+%spatial resolutions while finer details appear by zooming in.
+%In the framework of the VESPA project, the idea of using Aladin to display
+%planetary data, especially planetary HiPS, has emerged.
 HiPS are usually generated from a collection of WCS tagged FITS files.
@@
 %WCS part of a FITS header.
-
 Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
-been updated to support several planetary conventions.
-First, while the sky is displayed from the inside of the unit sphere, planetary
-surfaces are observed -- and must be displayed -- from the exterior.
-It means that the direction of the east on the horizontal axis has to be
-inverted when displaying planetary data.
-Then, planetary missions usually use different projections when providing
-images of planets polar caps and equatorial region.
-Often the former are in the stereographic (STG) projection while the later
-is in the Plate Carr\'ee (CAR) projection. 
-This specificity has been taken into account into the HiPS generator.
-Also, in the future, Aladin will be able to recognize the FITS keywords defined
+been updated to support planetary conventions.
+%Then, planetary missions usually use different projections when providing
+%images of planets polar caps and equatorial region.
+%Often the former are in the stereographic (STG) projection while the later
+%is in the Plate Carr\'ee (CAR) projection. 
+%This specificity has been taken into account into the HiPS generator.
+In the future, Aladin will be able to recognize the FITS keywords defined
 in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).

### 2018-02-27 ess_marmo.tex

@@
 \begin{keypoints}
-\item Inter-operable data formats 
-\item Planetary data standards
-\item Data processing and visualization
+\item We present an extension of FITS metadata description for planetary surface
+investigations. 
+\item A FITS description for planetary data aims
+to simplify sharing data across the planetary and astronomy domains.
+\item FITS open description metadata will promote interoperability from raw data
+formatting to final visualization.
 \end{keypoints}
@@
 % ^.
-In this paper we will describe how a small addition to the Flexible Image Transport System (FITS) 
-standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
-surface investigations. We will also show  how FITS metadata can easily be transformed
+In this paper we will describe how a small addition to the Flexible Image Transport
+System (FITS) standard, widely used in the astronomy investigations, will allow FITS
+to be more easily used in planetary surface investigations.
+We will also show how FITS metadata can easily be transformed
 in the PDS4 metadata archival model and lastly provide example implementations.
@@
 Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
-been updated to support planetary conventions.
+been updated to support planetary conventions (see Figure \ref{fig:aladin}).
 %Then, planetary missions usually use different projections when providing
@@
 %This specificity has been taken into account into the HiPS generator.
+
+\begin{figure}[ht!]
+\centerline{\includegraphics[height=20pc]{ess_marmo6}}
+\caption{Visualization of several level of detail on Mars with Aladin: from the
+global topographic map, as a HiPS layer, to a high resolution CRISM spectral cube
+in FITS format. All spectral analysis tools useful for astronomical spectra
+are then available for planetary data.}
+\label{fig:aladin}
+\end{figure}
+
 In the future, Aladin will be able to recognize the FITS keywords defined
@@
 distances on the surface of the body.
-Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
-barely used in extrasolar astronomy and are not supported in Aladin yet.
+Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of
+them are barely used in extrasolar astronomy and are not supported in Aladin yet.
 An effort has been started to revamp and update the code of projections in

### 2018-02-28 ess_marmo.tex

@@
 
+VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
+
 \subsection{MATISSE}
@@
 
-VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
-
 \section{Discussion and Perspectives}
 \label{sec:disc}
-FITS discussion in astronomy\citep{THOMAS2015133} 
-let's astronomy learn from planetary sciences too: accent must be put on interoperability
-.... blabla
-Planetary surface is lacking open and general software for massive processing and can benefit
-of astronomy experience. In addition to that WCS projection scheme is simpler than the
-historical Earth-based one used in planetary applications. 
-Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
-Reducing custom data format conversions during the processing chain but
+An extension to the FITS standard has been proposed to encompass planetary
+surface data description.
+Planetary surface investigation is lacking open and general software for massive data
+processing and can benefit from astronomy experience in this field.
+
+We are aware that FITS is today subject of questioning in astronomical applications
+\citep{THOMAS2015133}.
+We are not convinced that one format can solve all the issues faced from long-term
+archiving to on-line streaming.
+Planetary science research community has used a large variety of format customizing
+specific software.
+We want to stress here that using FITS could be an efficient compromise, reducing custom
+data format conversions during the processing chain and
 allowing easy conversion for final products following the needed application.
+
+FITS as an open and flexible data and metadata exchange and archiving format, could be
+inserted in the PDS4 data model scheme in order to address the issue of more complex
+metadata and data relationships.
+
+In order to improve interoperability and knowledge sharing between astronomy and
+planetary science, we will soon finalize the integration of FITS specifications in the
+GDAL library.
+In the framework of the VESPA project a review of the available and planned mission data
+archives is ongoing in order to evaluate their insertion in the planetary VO structure
+with a possible on-demand conversion in FITS format.
 

### 2018-02-28 ess_marmo.tex
Update on Overleaf.

@@
  S. Erard\affil{3},
- A. Zinzi\affil{4,5},
- M. Minin\affil{6},
- F.-X. Pineau\affil{7},
+ M. Minin\affil{4},
+ F.-X. Pineau\affil{5},
+ A. Zinzi\affil{6,7},
  B. Cecconi\affil{3},
- A. P. Rossi\affil{6}}
+ A. P. Rossi\affil{4}}
 
@@
 Univ. Paris Diderot, Sorbonne Paris Cit\'e, 5 place Jules Janssen, 92195 Meudon, France}
-\affiliation{4}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
-\affiliation{5}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
-\affiliation{6}{Jacobs University, Bremen, Germany}
-\affiliation{7}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
+\affiliation{4}{Jacobs University, Bremen, Germany}
+\affiliation{5}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}
+\affiliation{6}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
+\affiliation{7}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
 

### 2018-02-28 ess_marmo.tex
Update on Overleaf.

@@
 
-VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
+\subsection{FITS and QGIS}
+QGIS is a largely used open source geospatial analysis software.
+It is using GDAL for layer import: QGIS is then already capable to display FITS files,
+making use of a Virtual GDAL header, as described above.
+Once implemented in GDAL, FITS will officially become a supported format by QGIS.
+
+In addition to that, a QGIS plugin has been developed\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
+\citep{Minin} in order to download and display files exposed via EPN-TAP servers.
 

### 2018-03-01 ess_marmo.tex
Update on Overleaf.

@@
 \begin{keypoints}
-\item We present an extension of FITS metadata description for planetary surface
+\item We present an extension of FITS metadata for planetary surface
 investigations. 
 \item A FITS description for planetary data aims
-to simplify sharing data across the planetary and astronomy domains.
-\item FITS open description metadata will promote interoperability from raw data
+to simplify sharing data across the astronomy and planetary domains.
+\item An open FITS portrayal will promote interoperability from raw data
 formatting to final visualization.
@@
 standards.
-The option to use FITS within the planetary surface domain could help to
+The option to use FITS within the planetary surface domain can help to
 homogenize methods from raw formatting to visualization, while
@@
 value (\ref{table:proj}).
-For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG,
+For angular coordinates \citet{calagreisenI} provides the form xLN xLT or yzLN yzLT,
 where x or yz correspond to the target body.

### 2018-05-08 2018EA000388_Answers_1Reviewer.txt
Working on ESS first internal review.

2018EA000388 Answers to the first Reviewer

l. 43
Rephrased: see the new version of the manuscript

l. 57
We agree with the reviewer: the paragraph has been deleted

l. 68
Rephrased: see the new version of the manuscript

l. 68-69
IPDA informally discussed about proposing recommended data formats for archive.
As an IPDA official position is missing we cannot argue about FITS in that context.

l. 83-84

l. 102
GDAL lack of metadata description for FITS format: added in the manuscript

Table 1: note b

Table 2 : header
Mercury must be there... sorry for the oversight.
TNO and KBO are Asteroids: this is now clarified in the note.

### 2018-05-09 2018EA000388_Answers_1Reviewer.txt
Still about reviewer comments

@@
 l. 68-69
-IPDA informally discussed about proposing recommended data formats for archive.
+IPDA informally discussed about proposing recommended data formats for archive,
+but no conclusions are available at the moment.
 As an IPDA official position is missing we cannot argue about FITS in that context.
@@
 l. 83-84
+Rosetta data at PSA have been cited.
 
 l. 102
-GDAL lack of metadata description for FITS format: added in the manuscript
+GDAL lacks a metadata description/dictionary for FITS format: added in the manuscript
 
 Table 1: note b
+We agree with referee: the statement is confusing. Our goal here is to show that FITS metadata
+describe a number of projections used in Planetary surface data and the mapping to PDS4
+vocabulary is straightforward. The statement has been deleted.
+
+l. 144
+Planetary surface applications of standard keywords have been detailed. See the manuscript.
+
+l. 151
+WCSNAME is not the same than Table 1 column 1.
+WCSNAME is a human-readable short description of the coordinate system e.g. body-fixed rotating,
+-ocentric -ographic, non-rotating, inertia,... ). In PDS3 it was probably COORDINATE_SYSTEM_NAME + COORDINATE_SYSTEM_TYPE
+
+l. 152
+Rephrased: see the new version of the manuscript
+
+l. 175
+Rephrased: see the new version of the manuscript
+
+l. 177
+CUNIT is not mandatory in FITS standard: when absent, degrees are assumed.
+This is a valid assumption for planetary surfaces too. 
+Rephrased: see the new version of the manuscript
 

### 2018-05-11 2018EA000388_Answers_1Reviewer.txt
Almost done, only ORIGIN translation is missing.

@@
 TNO and KBO are Asteroids: this is now clarified in the note.
+
+l. 257
+ORIGIN?
+
+l.275-277
+Images have been modified to better show the equivalence of georeferenced TIFF or FITS.
+The paragraph has been modified accordingly.
+
+l. 279
+I have added a footnote to point the reader to a virtual header tutorial, in order to not to charge the
+text with overwhelming technical details.
+
+l. 307
+Fixed
+
+l. 347
+I have added a paragraph: GDAL implentation is in progress.
+
+l. 351
+Rephrased
+
+l. 389
+Rephrased
+
+l. 402
+Rephrased
+
+l. 404
+Rphrased

### 2018-05-11 ess_marmo.tex
Almost done, only ORIGIN translation is missing.

@@
 
-%% Choose from this list of Journals:
-%
-% JGR-Atmospheres
-% JGR-Biogeosciences
-% JGR-Earth Surface
-% JGR-Oceans
-% JGR-Planets
-% JGR-Solid Earth
-% JGR-Space Physics
-% Global Biochemical Cycles
-% Geophysical Research Letters
-% Paleoceanography
-% Radio Science
-% Reviews of Geophysics
-% Tectonics
-% Space Weather
-% Water Resource Research
-% Geochemistry, Geophysics, Geosystems
-% Journal of Advances in Modeling Earth Systems (JAMES)
-% Earth's Future
-% Earth and Space Science
-
 %% ------------------------------------------------------------------------ %%
@@
 
-% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
-% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
-
-% (include name and email addresses of the corresponding author.  More
-% than one corresponding author is allowed in this LaTeX file and for
-% publication; but only one corresponding author is allowed in our
-% editorial system.)
-
 %% Corresponding Author:
@@
 
-% Example: \correspondingauthor{First and Last Name}{email@address.edu}
-
-% Authors are individuals who have significantly contributed to the
-% research and preparation of the article. Group authors are allowed, if
-% each author in the group is separately identified in an appendix.)
-
-% \affiliation{1}{First Affiliation}
-% \affiliation{2}{Second Affiliation}
-% \affiliation{3}{Third Affiliation}
-% \affiliation{4}{Fourth Affiliation}
-
 %% Keypoints, final entry on title page.
-% Example: 
-% \begin{keypoints}
-% \item	List up to three key points (at least one is required)
-% \item	Key Points summarize the main points and conclusions of the article
-% \item	Each must be 100 characters or less with no special
-% characters or punctuation 
-% \end{keypoints}
-
 %% \begin{abstract} begins second page
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Track Changes:
-% To add words, \added{<word added>}
-% To delete words, \deleted{<word deleted>}
-% To replace words, \replace{<word to be replaced>}{<replacement word>}
-% To explain why change was made: \explain{<explanation>}
-
-% At the end of the document, use \listofchanges, which will list the
-% changes and the page and line number where the change was made.
-
-% When final version, \listofchanges will not produce anything,
-% \added{} word will be printed, \deleted{} will take away the word,
-% \replaced{}{} will print only the 2nd argument.
-% \explain will not print anything.
-
-% Optional argument:
-% You can also add additional information to be printed with the list
-% of changes, to indicate the initials of the person changing the text,
-% and the time and/or date of the change, or any other comment by using
-% the optional [] argument:
-% \added[AH, 3:30pm, Feb 18, 2016]{added term}
-% will yield 
-% [AH, 3:30pm, Feb 18, 2016] added term on page...
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 diverged as these related, but almost completely isolated domains evolved.
-% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
-% 
-% > Unfortunately, the technologies 
-% > and data formats used by researchers for planetary surface studies and the field of astronomy 
-% > diverged as these related, but almost completely isolated domains evolved.  
-% 
-% Is this too strong a statement?
-% 
-% ^.
+
 In this paper we will describe how a small addition to the Flexible Image Transport
@@
 Interoperability, openness of data formats, and shared processing techniques are 
-a necessity to efficiently extract scientific information from the data and 
-to guarantee the reproducibility of the scientific results.
+a necessity to allow efficient scientific analysis of the data and 
+to guarantee the reproducibility of the results.
 
@@
 
-This paper is organized as follows:
-\begin{enumerate}
-\item{a brief description of FITS format;}
-\item{a proposition for planetary surface metadata description in FITS;}
-\item{dictionaries between FITS, PDS metadata description and EPN-TAP2.0
-parameters;}
-\item{some examples of planetary dataset in FITS;}
-\item{current developments for making FITS more usable to handle data
-of planetary surfaces.}
-\end{enumerate}
-
 \section{FITS Format in Planetary Surface Investigation}
@@
 FITS is an open digital standard, created in the late 1970's for data acquisition,
-transfer and archiving of telescope data by astronomical observatories.
-It had been used for data exchange and archiving from several orbital telescopes.
+transfer and archiving of telescope data by astronomical observatories, where
+it has been used during the last thirty years.
+It had been adopted for data exchange and archiving from several orbital telescopes
+and spatial missions.
 The International Astronomical Union approved FITS as the standard format
@@
 Therefore, FITS is one of the standard formats implemented in the Virtual
-Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
-(VO).
+Observatory (VO).
 FITS data storage is compatible with the Planetary Data System (PDS)
@@
 and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
+Some Rosetta data, from NAVCAM and OSIRIS cameras, are distributed in FITS format at
+the ESA Planetary Science Archive\footnote{https://archives.esac.esa.int/psa} (PSA).
 In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed
@@
 GDAL is essentially a format translation library written in C++ for geospatial raster
-and vector data and, fortunately, it also offers basic support for the FITS format.
+and vector data and, fortunately, it also offers basic support for the FITS format,
+however missing a FITS metadata dictionary.
 
@@
 \begin{table}
-\caption{Projection names and codes already available in FITS suitable for
+\caption{Projection names and codes available in WCS FITS, suitable for
 planetary applications: the corresponding PDS4 terminology is
@@
 \hline
-FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
+FITS Name$^{a}$ & FITS code$^{a}$ & PDS4 name$^{b}$ \\ 
 \hline
@@
 \multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
-\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}.
-We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
-\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator: those three projection are not used for archival or analysis.}
+\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary}
 \end{tabular}
@@
 keyword, containing the projection code and the coordinate system.
-% * <chiara.marmo@u-psud.fr> 2018-02-22T21:31:39.056Z:
-% 
-% Example or scheme needed in order to better understand the coordinate transformation flow
-% 
-% ^.
-
 The coordinate system must be defined with respect to a specific inertial reference frame.
-% * <chiara.marmo@u-psud.fr> 2018-02-26T11:33:56.098Z:
-% 
-% Please be more specific about the inertial frame
-% 
-% ^.
 Multiple coordinate system descriptions are allowed in the same header: this is a
@@
 description of the surface are often needed at the same time.
-Further below 3D keyword will be proposed for FITS metadata, in order to standardize
+Further below 3D keywords will be proposed for FITS metadata, in order to standardize
 the conversion between angular and linear coordinates: alternate description will still
@@
 wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
-astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
+astropy\footnote{http://www.astropy.org/}, \citet{astropy}) and some FITS visualization
 software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
@@
 Name of the inertial reference frame with respect to which the coordinate
-system is defined (defaulted to \texttt{`ICRS'})}
+system is defined. Standard FITS defaults this keyword to \texttt{`ICRS'}.
+This is applicable to planetary data too as long as the planetary coordinate system
+parameters, i.e. the rotational elements, are defined with respect to \texttt{`ICRS'}.
+This is the case starting from the 2003 WGCCRE report \citep{wgccre2003}.
+Before the 1982 WGCCRE report \citep{wgccre1982} the inertial system was \texttt{`FK4'},
+after it was \texttt{`FK5'}.}
 \item{EQUINOX.
 The value field contains the equinox in years for the celestial
-coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
+coordinate system in which positions are expressed.
+Standard FITS defaults this keyword to \texttt{`J2000'}.
+This is applicable to planetary data starting from the 1982 WGCCRE report \citep{wgccre1982}.
+Before 1982 the standard equinox was \texttt{`B1950'}.}
 \item{WCSAXES.
@@
 generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
-additional time dependency is added to the coordinate description.}
+additional time or filter dependency is added to the coordinate description.}
 \item{WCSNAME.
@@
 non-rotating, inertial).
-EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
+EPN-TAP provides similar information through the \texttt{spatial\_coordinate\_description} parameter.}
 \item{CRPIXn.
@@
 matrix PCi\_j).
-In PDS4 they correspond to the \textit{pixel\_resolution\_x} and
-\texttt{pixel\_resolution\_y} attributes.
-In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
+They correspond to the \textit{pixel\_resolution\_x} and
+\texttt{pixel\_resolution\_y} attributes in PDS4, to the \texttt{spatial\_resolution} parameter
+in EPN-TAP.}
 \item{CUNITn.
-Units of the coordinate system, not mandatory, defaulted to degrees for angular
-coordinates and to meters (m) for linear ones.}
+Units of the coordinate system. This is a not mandatory keyword: when not present FITS standard
+assumes its value as degrees, the most common description in astronomical sky images.
+The same can be assumed for planetary surfaces: when linear coordinates are used SI units must
+be used (meters, m).}
 \item{CNAMEn.
@@
 Moon & SE \\
+Mercury & ME \\
 Venus & VE \\
@@
 Satellites (other than the Moon) & ST \\
-Asteroids & AS \\
+Asteroids$^{b}$ & AS \\
 Dwarf Planets & DW \\
@@
 \hline
-\multicolumn{2}{l}{$^{a}$From \citet{calagreisenI}}
+\multicolumn{2}{l}{$^{a}$From \citet{calagreisenI}} \\
+\multicolumn{2}{l}{$^{b}$Transneptunian (TNO) and Kuiper Belt (KBO) objects must be placed there}
 \end{tabular}
@@
 PDS4 provides a \texttt{Time\_Coordinates} class with attributes
-\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
+\texttt{start\_date\_time} and \texttt{stop\_date\_time} providing start and
 stop date and time appropriate to the product being labeled, always in UTC.
@@
 
-%
-%\item{\texttt{AUTHOR}}
-
 \item{\texttt{REFERENC}.
@@
 surface, FITS and GeoTIFF are completely interchangeable.
-As an example we have downloaded some global maps from USGS
-PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
-We used ds9 to display the FITS image and were able to visualize coordinates in meters
-as in QGIS, a popular open source GIS software, but also in planetary latitude and
-longitude thanks to the alternate coordinate
-description (see figure \ref{fig:ds9qgis}, left side).
+As an example we have downloaded a Mercury polar map from the USGS
+PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted it in FITS.
+The FITS image has been cropped in two smaller versions using the open source software
+SWarp\footnote{https://www.astromatic.net/software/swarp} \citep{swarp}: SWarp was able to
+automatically manage WCS planetary information.
+We used ds9 to display the three FITS images and were able to correctly align them.
+Linear coordinates in meters are visualized (see figure \ref{fig:ds9qgis}, left side)
+as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
+description.
 \begin{figure}[ht!]
 \centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
-\caption{left: Mercury South Pole FITS image displayed in ds9;
-right: GeoTIFF raster (displayed in yellow) and the virtual
-header linked to the FITS raster (displayed in red); the two layers
+\caption{left: Mercury South Pole FITS image displayed in ds9, alternate coordinates in meters
+are displayed;
+right: GeoTIFF raster (displayed in red) and the two virtual haeders
+linked to FITS rasters (displayed in green and blue); the three layers
 are perfectly superposed.
@@
 support for FITS images is to use a GDAL-supported detached header.
-GDAL Virtual Header files simply describe the internal structure of a FITS image, but
+GDAL Virtual Header\footnote{http://www.gdal.org/gdal\_vrttut.html}
+files simply describe the internal structure of a FITS image, but
 also the map projection and body size in a standardized "well known text"
@@
 acquired by the imaging spectrometer OMEGA on-board of Mars Express.
-OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA and PDS
-archives.
+OMEGA unprojected and uncalibrated data (level1b) are distributed by ESA-PSA \citep{BESSE2018131}.
 Their structure is described in the OMEGA Experiment Archive 
@@
 making use of a Virtual GDAL header, as described above.
+The complete implementation of FITS metadata in GDAL library is in progress.
 Once implemented in GDAL, FITS will officially become a supported format by QGIS.
@@
 access, visualize and analyze data from solar system exploration missions.
-Currently, it provides access to public data from five different targets (i.e., Mercury,
-Venus, Mars, dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local
-storage or by querying remote dataset, such as VESPA, Planet Server and NASA JPL and
-can be integrated with other datasets thanks to its modular structure.
-One of th MATISSE 2D outputs is formatted as FITS, projected by interpolating the
-original observations to a grid regularly spaced
-in planetocentric coordinates (i.e., latitude and longitude).
-We used the planetary FITS standard described in this work in order to make the output
+Currently, MATISSE provides access to public data from five different targets (i.e., Mercury,
+Venus, Mars, the dwarf planet Ceres and the asteroid Vesta).
+MATISSE 2D outputs are formatted as FITS, in equirectangular projection and in planetocentric coordinates.
+The planetary FITS standard described in this work has been implemented in MATISSE in order to make the output
 file compliant with most FITS readers, especially DS9 and JS9.
@@
 Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
-online by the end of 2018) we are planning to make use of JS9 as the standard reader for
+online by the end of 2018) JS9 will be the standard reader for
 the 2D online output of the tool, thus allowing users to perform advanced analysis in
@@
 \subsection{Aladin}
-Aladin is both a sky atlas and a portal to access data -- mainly images and
-catalogues -- available through the VO.
-It is able to load FITS files and to display images taking into account various
+Aladin is both a sky atlas and a portal to access data, mainly images and
+catalogues, available through the VO.
+Aladin is able to load FITS files and to display images taking into account various
 WCS projections.
-Like DS9, Aladin is popular and broadly used in the part of the astronomical
-community manipulating images.
 So far it is also the reference generator and visualizer of Hierarchical
@@
 resolutions, offering a progressive view of possibly very large datasets.
-%The purpose is to allows users to explore all-sky surveys without having to
-%explicitly perform any query: the most important information is visible at low
-%spatial resolutions while finer details appear by zooming in.
-%In the framework of the VESPA project, the idea of using Aladin to display
-%planetary data, especially planetary HiPS, has emerged.
 HiPS are usually generated from a collection of WCS tagged FITS files.
-%As an alternative, Aladin can use JPEG files together with additional metadata
-%files.
-%A metadata file -- one by JPEG file -- basically contains in a text format the
-%WCS part of a FITS header.
 Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
 been updated to support planetary conventions (see Figure \ref{fig:aladin}).
-%Then, planetary missions usually use different projections when providing
-%images of planets polar caps and equatorial region.
-%Often the former are in the stereographic (STG) projection while the later
-%is in the Plate Carr\'ee (CAR) projection. 
-%This specificity has been taken into account into the HiPS generator.
 
@@
 in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
-For example, it will be able to take into account a planet's radii to compute
-distances on the surface of the body.
 Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of
@@
 Aladin.
-We expect all projections to be available in a forthcoming release.
 
@@
 
-FITS as an open and flexible data and metadata exchange and archiving format, could be
-inserted in the PDS4 data model scheme in order to address the issue of more complex
-metadata and data relationships.
-
-In order to improve interoperability and knowledge sharing between astronomy and
-planetary science, we will soon finalize the integration of FITS specifications in the
-GDAL library.
+FITS is an open and flexible data and metadata exchange and archiving format, its use
+in the PDS4 data model scheme would solve the issue of complex
+metadata and data relationships separating data processing and the data archiving processes.
+
 In the framework of the VESPA project a review of the available and planned mission data
@@
 with a possible on-demand conversion in FITS format.
+The integration of FITS specifications in the GDAL library will be soon finalize with the goal to
+improve data processing efficiency in planetary science.
 
@@
 %
-%\section{Here Is Appendix Title}
-% will show
-% A: Here Is Appendix Title
-%
-%\appendix
-%\section{Here Is Appendix Title}
-%This is an Appendix section.
-
-%\begin{linenomath*}
-%\begin{equation}asdf\end{equation}
-%\end{linenomath*}
-
-\appendix
-\section{Planetary Header example}
-%An example of planetary header goes here.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %
-
-\bibliography{ess_marmo}
+%\bibliography{ess_marmo}
 
@@
 
-%\begin{thebibliography}{}
-
-% % \bibitem[{\textit{Bell and Munoz}}(2008)]{Boug10} Bell, A.~H., and
-% % Munoz, D.~P.  (2008). Activity in the superior colliculus reflects
-% % dynamic interactions between voluntary and involuntary influences on
-% % orienting behaviour. \textit{Eur. J. Neurosci.} 28, 1654--1660.
-
-% % \bibitem[{\textit{Corbetta et~al.}}(1991)]{Buiz07} Corbetta, M.,
-% % Miezin, F.~M., Dobmeyer, S., Shulman, G.~L., and Petersen, S.~E.
-% % (1991). Selective and divided attention during visual discriminations
-% % of shape, color, and speed: functional anatomy by positron emission
-% % tomography. \textit{J.~Neurosci.} 11, 2383--2402.
-
-% % \bibitem[{\textit{Borra et~al.}}(2014)]{Buiz98} Borra, E., Gerbella,
-% % M., Rozzi, S., Tonelli, S., and Luppino, G.  (2014). Projections to
-% % the superior colliculus from inferior parietal, ventral premotor, and
-% % ventrolateral prefrontal areas involved in controlling goal-directed
-% % hand actions in the macaque. \textit{Cereb. Cortex} 24, 1054--1065.
-
-% % \bibitem[{\textit{Dorris et~al.}}(1997)]{Fra10}
-% %  Dorris, M.~C.,
-% % Par\'{e}, M., and Munoz, D.~P.  (1997). Neuronal activity in monkey
-% % superior colliculus related to the initiation of saccadic eye
-% % movements. \textit{J.~Neurosci.} 17, 8566--8579.
-
-% % \bibitem[{\textit{Elsabbagh et~al.}}(2009)]{Ghel00} 
-% % Elsabbagh, M.,
-% % Volein, A., Holmboe, K., Tucker, L., Csibra, G., Baron-Cohen, S.,
-% % et~al.  (2009). Visual orienting in the early broader autism
-% % phenotype: disengagement. \textit{J.~Child Psychol. Psychiatry} 50,
-% % 637--642.
-
-% % \bibitem[{\textit{Fortin et~al.}}(1999)]{Gneit05} Fortin, S., Chabli,
-% % A., Dumont, I., Shumikhina, S., Itaya, S.~K., and Molotchnikoff, S.
-% % (1999). Maturation of visual receptive field properties in the rat
-% % superior colliculus. \textit{bibain Res. Dev. Brain Res.} 1112,
-% % 55--64.
-
-% % \bibitem[{\textit{Felsen and Mainen}}(2008)]{Gneit07b} Felsen, G., and
-% % Mainen, Z.~F.  (2008). Neural substrates of sensory-guided locomotor
-% % decisions in the rat superior colliculus. \textit{Neuron} 60,
-% % 137--148.
-
-% % \bibitem[{\textit{Gattass and Desimone}}(1996)]{Haid14} Gattass, R.,
-% % and Desimone, R.  (1996). Responses of cells in the superior
-% % colliculus during performance of a spatial attention task in the
-% % macaque. \textit{Rev. Bras. Biol.} 56, 257--279.
-
-% % \bibitem[{\textit{Goldberg and Wurtz}}(1972)]{Hami00} Goldberg, M.~E.,
-% % and Wurtz, R.~H.  (1972). Activity of superior colliculus in behaving
-% % monkey. II. Effect of attention on neuronal responses.
-% % \textit{J.~Neurophysiol.} 35, \hbox{560--574}.
-
-% % \bibitem[{\textit{Krauzlis}}(2003)]{Kendall} Krauzlis, R.~J.  (2003).
-% % Neuronal activity in the rostral superior colliculus related to the
-% % initiation of pursuit and saccadic eye movements.
-% % \textit{J.~Neurosci.} 23, 4333--4344.
-
-% % \bibitem[{\textit{Heesy}}(2009)]{Leit74} 
-% % Heesy, C.~P.  (2009). Seeing
-% % in stereo: the ecology and evolution of primate binocular vision and
-% % stereopsis. \textit{Evol. Anthropol.} 18, 21--35.
-
-% % \bibitem[{\textit{Hilbig et~al.}}(2000)]{Mann45} Hilbig, H., Bidmon,
-% % H.~J., Ettrich, P., and M\"{u}ller, A.  (2000). Projection neurons in
-% % the superficial layers of the superior colliculus in the rat: a
-% % topographic and quantitative morphometric analysis.
-% % \textit{Neuroscience} 96, 109--119.
-
-% % \bibitem[{\textit{Ignashchenkova et~al.}}(2004)]{Math76}
-% % Ignashchenkova, A., Dicke, P.~W., Haarmeier, T., and Their, P.
-% % (2004). Neuron-specific contribution of the superior colliculus to
-% % overt and covert shifts of attention. \textit{Nat. Neurosci.} 7,
-% % 56--64.
-
-% \bibitem[{\textit{Krauzlis et~al.}}(2013)]{Palm00} Krauzlis, R.~J.,
-% Lovejoy, L.~P., and Z\'{e}non, A.  (2013). Superior colliculus and
-% visual spatial attention. \textit{Annu. Rev. Neurosci.} 36, 165--182.
-
-% \bibitem[{\textit{Kustov and Robinson}}(1996)]{Papp09} Kustov, A.~A.,
-% and Robinson, D.~L.  (1996). Shared neural control of attentional
-% shifts and eye movements. \textit{Nature} 384, 74--77.
-
-% \bibitem[{\textit{Landry and Bryson}}(2004)]{Park08} Landry, R., and
-% Bryson, S.~E.  (2004). Impaired disengagement of attention in young
-% children with autism. \textit{J.~Child Psychol. Psychiatry} 45,
-% 1115--1122.
-
-% \bibitem[{\textit{Kobayashi et~al.}}(2003)]{R2013} Kobayashi, T.,
-% Tran, A.~H., Nishijo, H., Ono, T., and Matsumoto, G.  (2003).
-% Contribution of hippocampal place cell activity to learning and
-% formation of goal-directed navigation in rats. \textit{Neuroscience}
-% 117, 1025--1035.
-
-% \bibitem[{\textit{McHaffie et~al.}}(2005)]{Raft05} McHaffie, J.~G.,
-% Stanford, T.~R., Stein, B.~E., Coizet, V., and Redgrave, P.  (2005).
-% Subcortical loops through the basal ganglia. \textit{Trends Neurosci.}
-% 28, 401--407.
-
-% \bibitem[{\textit{McPeek and Keller}}(2004)]{Rich13} McPeek, R.~M.,
-% and Keller, E.~L.  (2004). Deficits in saccade target selection after
-% inactivation of superior colliculus. \textit{Nat. Neurosci.} 7,
-% 757--763.
-
-% \bibitem[{\textit{M\"{u}ller et~al.}}(2005)]{Scheu14} M\"{u}ller,
-% J.~R., Philiastides, M.~G., and Newsome, W.~T.  (2005).
-% Microstimulation of the superior colliculus focuses attention without
-% moving the eyes. \textit{Proc. Natl. Acad. Sci. U.S.A.} 102, 524--529.
-
-% \bibitem[{\textit{Munoz and Istvan}}(1998)]{ScheueBue13} Munoz, D.~P.,
-% and Istvan, P.~J.  (1998). Lateral inhibitory interactions in
-% the intermediate layers of the monkey superior colliculus.
-% \textit{J.~Neurophysiol.} 79, 1193--1209.
-
-%\end{thebibliography}
+\begin{thebibliography}{22}
+\providecommand{\natexlab}[1]{#1}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
+
+\bibitem[{\textit{{Archinal} et~al.}(2011)\textit{{Archinal}, {A'Hearn},
+  {Bowell}, {Conrad}, {Consolmagno}, {Courtin}, {Fukushima}, {Hestroffer},
+  {Hilton}, {Krasinsky}, {Neumann}, {Oberst}, {Seidelmann}, {Stooke}, {Tholen},
+  {Thomas}, and {Williams}}}]{wgccre}
+{Archinal}, B.~A., M.~F. {A'Hearn}, E.~{Bowell}, A.~{Conrad}, G.~J.
+  {Consolmagno}, R.~{Courtin}, T.~{Fukushima}, D.~{Hestroffer}, J.~L. {Hilton},
+  G.~A. {Krasinsky}, G.~{Neumann}, J.~{Oberst}, P.~K. {Seidelmann},
+  P.~{Stooke}, D.~J. {Tholen}, P.~C. {Thomas}, and I.~P. {Williams} (2011),
+  {Report of the IAU Working Group on Cartographic Coordinates and Rotational
+  Elements: 2009}, \textit{Celestial Mechanics and Dynamical Astronomy},
+  \textit{109}, 101--135, \doi{10.1007/s10569-010-9320-4}.
+
+\bibitem[{\textit{{Astropy Collaboration} et~al.}(2013)\textit{{Astropy
+  Collaboration}, {Robitaille}, {Tollerud}, {Greenfield}, {Droettboom}, {Bray},
+  {Aldcroft}, {Davis}, {Ginsburg}, {Price-Whelan}, {Kerzendorf}, {Conley},
+  {Crighton}, {Barbary}, {Muna}, {Ferguson}, {Grollier}, {Parikh}, {Nair},
+  {Unther}, {Deil}, {Woillez}, {Conseil}, {Kramer}, {Turner}, {Singer}, {Fox},
+  {Weaver}, {Zabalza}, {Edwards}, {Azalee Bostroem}, {Burke}, {Casey},
+  {Crawford}, {Dencheva}, {Ely}, {Jenness}, {Labrie}, {Lim}, {Pierfederici},
+  {Pontzen}, {Ptak}, {Refsdal}, {Servillat}, and {Streicher}}}]{astropy}
+{Astropy Collaboration}, T.~P. {Robitaille}, E.~J. {Tollerud}, P.~{Greenfield},
+  M.~{Droettboom}, E.~{Bray}, T.~{Aldcroft}, M.~{Davis}, A.~{Ginsburg}, A.~M.
+  {Price-Whelan}, W.~E. {Kerzendorf}, A.~{Conley}, N.~{Crighton}, K.~{Barbary},
+  D.~{Muna}, H.~{Ferguson}, F.~{Grollier}, M.~M. {Parikh}, P.~H. {Nair}, H.~M.
+  {Unther}, C.~{Deil}, J.~{Woillez}, S.~{Conseil}, R.~{Kramer}, J.~E.~H.
+  {Turner}, L.~{Singer}, R.~{Fox}, B.~A. {Weaver}, V.~{Zabalza}, Z.~I.
+  {Edwards}, K.~{Azalee Bostroem}, D.~J. {Burke}, A.~R. {Casey}, S.~M.
+  {Crawford}, N.~{Dencheva}, J.~{Ely}, T.~{Jenness}, K.~{Labrie}, P.~L. {Lim},
+  F.~{Pierfederici}, A.~{Pontzen}, A.~{Ptak}, B.~{Refsdal}, M.~{Servillat}, and
+  O.~{Streicher} (2013), {Astropy: A community Python package for astronomy},
+  \textit{Astronomy and Astrophysics}, \textit{558}, A33,
+  \doi{10.1051/0004-6361/201322068}.
+
+\bibitem[{\textit{{Bertin} et~al.}(2002)\textit{{Bertin}, {Mellier},
+  {Radovich}, {Missonnier}, {Didelon}, and {Morin}}}]{swarp}
+{Bertin}, E., Y.~{Mellier}, M.~{Radovich}, G.~{Missonnier}, P.~{Didelon}, and
+  B.~{Morin} (2002), {The TERAPIX Pipeline}, in \textit{Astronomical Data
+  Analysis Software and Systems XI}, \textit{Astronomical Society of the
+  Pacific Conference Series}, vol. 281, edited by D.~A. {Bohlender},
+  D.~{Durand}, and T.~H. {Handley}, p. 228.
+
+\bibitem[{\textit{Besse et~al.}(2018)\textit{Besse, Vallat, Barthelemy, Coia,
+  Costa, Marchi, Fraga, Grotheer, Heather, Lim, Martinez, Arviset, Barbarisi,
+  Docasal, Macfarlane, Rios, Saiz, and Vallejo}}]{BESSE2018131}
+Besse, S., C.~Vallat, M.~Barthelemy, D.~Coia, M.~Costa, G.~D. Marchi, D.~Fraga,
+  E.~Grotheer, D.~Heather, T.~Lim, S.~Martinez, C.~Arviset, I.~Barbarisi,
+  R.~Docasal, A.~Macfarlane, C.~Rios, J.~Saiz, and F.~Vallejo (2018), Esa's
+  planetary science archive: Preserve and present reliable scientific data
+  sets, \textit{Planetary and Space Science}, \textit{150}, 131 -- 140,
+  \doi{https://doi.org/10.1016/j.pss.2017.07.013}, enabling Open and
+  Interoperable Access to Planetary Science and Heliophysics Databases and
+  Tools.
+
+\bibitem[{\textit{{Bibring} et~al.}(2005)\textit{{Bibring}, {Langevin},
+  {Gendrin}, {Gondet}, {Poulet}, {Berth{\'e}}, {Soufflot}, {Arvidson},
+  {Mangold}, {Mustard}, {Drossart}, {OMEGA Team}, {Erard}, {Forni}, {Combes},
+  {Encrenaz}, {Fouchet}, {Merchiorri}, {Belluci}, {Altieri}, {Formisano},
+  {Bonello}, {Capaccioni}, {Cerroni}, {Coradini}, {Fonti}, {Kottsov},
+  {Ignatiev}, {Moroz}, {Titov}, {Zasova}, {Mangold}, {Pinet}, {Dout{\'e}},
+  {Schmitt}, {Sotin}, {Hauber}, {Hoffmann}, {Jaumann}, {Keller}, {Duxbury}, and
+  {Forget}}}]{omega}
+{Bibring}, J.-P., Y.~{Langevin}, A.~{Gendrin}, B.~{Gondet}, F.~{Poulet},
+  M.~{Berth{\'e}}, A.~{Soufflot}, R.~{Arvidson}, N.~{Mangold}, J.~{Mustard},
+  P.~{Drossart}, {OMEGA Team}, S.~{Erard}, O.~{Forni}, M.~{Combes},
+  T.~{Encrenaz}, T.~{Fouchet}, R.~{Merchiorri}, G.~{Belluci}, F.~{Altieri},
+  V.~{Formisano}, G.~{Bonello}, F.~{Capaccioni}, P.~{Cerroni}, A.~{Coradini},
+  S.~{Fonti}, V.~{Kottsov}, N.~{Ignatiev}, V.~{Moroz}, D.~{Titov}, L.~{Zasova},
+  M.~{Mangold}, P.~{Pinet}, S.~{Dout{\'e}}, B.~{Schmitt}, C.~{Sotin},
+  E.~{Hauber}, H.~{Hoffmann}, R.~{Jaumann}, U.~{Keller}, T.~{Duxbury}, and
+  F.~{Forget} (2005), {Mars Surface Diversity as Revealed by the OMEGA/Mars
+  Express Observations}, \textit{Science}, \textit{307}, 1576--1581,
+  \doi{10.1126/science.1108806}.
+
+\bibitem[{\textit{{Bonnarel} et~al.}(2000)\textit{{Bonnarel}, {Fernique},
+  {Bienaym{\'e}}, {Egret}, {Genova}, {Louys}, {Ochsenbein}, {Wenger}, and
+  {Bartlett}}}]{aladin}
+{Bonnarel}, F., P.~{Fernique}, O.~{Bienaym{\'e}}, D.~{Egret}, F.~{Genova},
+  M.~{Louys}, F.~{Ochsenbein}, M.~{Wenger}, and J.~G. {Bartlett} (2000), {The
+  ALADIN interactive sky atlas. A reference tool for identification of
+  astronomical sources}, \textit{Astronomy and Astrophysics Supplement},
+  \textit{143}, 33--40, \doi{10.1051/aas:2000331}.
+
+\bibitem[{\textit{{Calabretta} and {Greisen}}(2002)}]{calagreisenI}
+{Calabretta}, M.~R., and E.~W. {Greisen} (2002), {Representations of celestial
+  coordinates in FITS}, \textit{Astronomy and Astrophysics}, \textit{395},
+  1077--1122, \doi{10.1051/0004-6361:20021327}.
+
+\bibitem[{\textit{{Davies} et~al.}(1983)\textit{{Davies}, {Abalakin}, {Lieske},
+  {Seidelmann}, {Sinclair}, {Sinzi}, {Smith}, and {Tjuflin}}}]{wgccre1982}
+{Davies}, M.~E., V.~K. {Abalakin}, J.~H. {Lieske}, P.~K. {Seidelmann}, A.~T.
+  {Sinclair}, A.~M. {Sinzi}, B.~A. {Smith}, and Y.~S. {Tjuflin} (1983), {Report
+  of the IAU working group on cartographic coordinates and rotational elements
+  of the planets and satellites - 1982}, \textit{Celestial Mechanics},
+  \textit{29}, 309--321, \doi{10.1007/BF01228525}.
+
+\bibitem[{\textit{Erard et~al.}(2014)\textit{Erard, Cecconi, Sidaner, Berthier,
+  Henry, Molinaro, Giardino, Bourrel, Andr{\'e}, Gangloff, Jacquey, and
+  Topf}}]{ERARDepntap}
+Erard, S., B.~Cecconi, P.~L. Sidaner, J.~Berthier, F.~Henry, M.~Molinaro,
+  M.~Giardino, N.~Bourrel, N.~Andr{\'e}, M.~Gangloff, C.~Jacquey, and F.~Topf
+  (2014), The epn-tap protocol for the planetary science virtual observatory,
+  \textit{Astronomy and Computing}, \textit{7-8}, 52 -- 61,
+  \doi{https://doi.org/10.1016/j.ascom.2014.07.008}, special Issue on The
+  Virtual Observatory: I.
+
+\bibitem[{\textit{Erard et~al.}(2018)\textit{Erard, Cecconi, Sidaner, Rossi,
+  Capria, Schmitt, G\'enot, Andr\'e, Vandaele, Scherf, Hueso,
+  M{\"a}{\"a}tt{\"a}nen, Thuillot, Carry, Achilleos, Marmo, Santolik, Benson,
+  Fernique, Beigbeder, Millour, Rousseau, Andrieu, Chauvin, Minin, Ivanoski,
+  Longobardo, Bollard, Albert, Gangloff, Jourdane, Bouchemit, Glorian, Trompet,
+  Al-Ubaidi, Juaristi, Desmars, Guio, Delaa, Lagain, Soucek, and
+  Pisa}}]{vespapss}
+Erard, S., B.~Cecconi, P.~L. Sidaner, A.~Rossi, M.~Capria, B.~Schmitt,
+  V.~G\'enot, N.~Andr\'e, A.~Vandaele, M.~Scherf, R.~Hueso,
+  A.~M{\"a}{\"a}tt{\"a}nen, W.~Thuillot, B.~Carry, N.~Achilleos, C.~Marmo,
+  O.~Santolik, K.~Benson, P.~Fernique, L.~Beigbeder, E.~Millour, B.~Rousseau,
+  F.~Andrieu, C.~Chauvin, M.~Minin, S.~Ivanoski, A.~Longobardo, P.~Bollard,
+  D.~Albert, M.~Gangloff, N.~Jourdane, M.~Bouchemit, J.-M. Glorian, L.~Trompet,
+  T.~Al-Ubaidi, J.~Juaristi, J.~Desmars, P.~Guio, O.~Delaa, A.~Lagain,
+  J.~Soucek, and D.~Pisa (2018), Vespa: A community-driven virtual observatory
+  in planetary science, \textit{Planetary and Space Science}, \textit{150}, 65
+  -- 85, \doi{https://doi.org/10.1016/j.pss.2017.05.013}, enabling Open and
+  Interoperable Access to Planetary Science and Heliophysics Databases and
+  Tools.
+
+\bibitem[{\textit{{Greisen} and {Calabretta}}(2002)}]{greisencalaII}
+{Greisen}, E.~W., and M.~R. {Calabretta} (2002), {Representations of world
+  coordinates in FITS}, \textit{Astronomy and Astrophysics}, \textit{395},
+  1061--1075, \doi{10.1051/0004-6361:20021326}.
+
+\bibitem[{\textit{{Greisen, E. W.} et~al.}(2006)\textit{{Greisen, E. W.},
+  {Calabretta, M. R.}, {Valdes, F. G.}, and {Allen, S. L.}}}]{wcsspectral}
+{Greisen, E. W.}, {Calabretta, M. R.}, {Valdes, F. G.}, and {Allen, S. L.}
+  (2006), Representations of spectral coordinates in fits, \textit{Astronomy
+  and Astrophysics}, \textit{446}(2), 747--771,
+  \doi{10.1051/0004-6361:20053818}.
+
+\bibitem[{\textit{{Hare} et~al.}(2006)\textit{{Hare}, {Archinal}, {Plesea},
+  {Dobinson}, and {Curkendall}}}]{hare2006}
+{Hare}, T.~M., B.~A. {Archinal}, L.~{Plesea}, E.~{Dobinson}, and
+  D.~{Curkendall} (2006), {Standards Proposal to Support Planetary Coordinate
+  Reference Systems in Open Geospatial Web Services and Geospatial
+  Applications}, in \textit{37th Annual Lunar and Planetary Science
+  Conference}, \textit{Lunar and Planetary Inst.~Technical Report}, vol.~37,
+  edited by S.~{Mackwell} and E.~{Stansbery}.
+
+\bibitem[{\textit{McEwen et~al.}(2007)\textit{McEwen, Eliason, Bergstrom,
+  Bridges, Hansen, Delamere, Grant, Gulick, Herkenhoff, Keszthelyi, Kirk,
+  Mellon, Squyres, Thomas, and Weitz}}]{hirise}
+McEwen, A.~S., E.~M. Eliason, J.~W. Bergstrom, N.~T. Bridges, C.~J. Hansen,
+  W.~A. Delamere, J.~A. Grant, V.~C. Gulick, K.~E. Herkenhoff, L.~Keszthelyi,
+  R.~L. Kirk, M.~T. Mellon, S.~W. Squyres, N.~Thomas, and C.~M. Weitz (2007),
+  Mars reconnaissance orbiter's high resolution imaging science experiment
+  (hirise), \textit{Journal of Geophysical Research: Planets},
+  \textit{112}(E5), n/a--n/a, \doi{10.1029/2005JE002605}, e05S02.
+
+\bibitem[{\textit{{Minin} et~al.}(2018)\textit{{Minin}, {Rossi}, {Marco
+  Figuera}, {Unnithan}, Marmo, Walter, Demleitner, Le~Sidaner, Cecconi, S., and
+  Hare}}]{Minin}
+{Minin}, M., A.~{Rossi}, R.~{Marco Figuera}, V.~{Unnithan}, C.~Marmo,
+  S.~Walter, M.~Demleitner, P.~Le~Sidaner, B.~Cecconi, E.~S., and T.~Hare
+  (2018), {Bridging the gap between Geographical Information Systems and
+  Planetary Virtual Observatory}, \textit{this issue}.
+
+\bibitem[{\textit{Murchie et~al.}(2007)\textit{Murchie, Arvidson, Bedini,
+  Beisser, Bibring, Bishop, Boldt, Cavender, Choo, Clancy, Darlington,
+  Des~Marais, Espiritu, Fort, Green, Guinness, Hayes, Hash, Heffernan, Hemmler,
+  Heyler, Humm, Hutcheson, Izenberg, Lee, Lees, Lohr, Malaret, Martin,
+  McGovern, McGuire, Morris, Mustard, Pelkey, Rhodes, Robinson, Roush,
+  Schaefer, Seagrave, Seelos, Silverglate, Slavney, Smith, Shyong, Strohbehn,
+  Taylor, Thompson, Tossman, Wirzburger, and Wolff}}]{crism}
+Murchie, S., R.~Arvidson, P.~Bedini, K.~Beisser, J.-P. Bibring, J.~Bishop,
+  J.~Boldt, P.~Cavender, T.~Choo, R.~T. Clancy, E.~H. Darlington,
+  D.~Des~Marais, R.~Espiritu, D.~Fort, R.~Green, E.~Guinness, J.~Hayes,
+  C.~Hash, K.~Heffernan, J.~Hemmler, G.~Heyler, D.~Humm, J.~Hutcheson,
+  N.~Izenberg, R.~Lee, J.~Lees, D.~Lohr, E.~Malaret, T.~Martin, J.~A. McGovern,
+  P.~McGuire, R.~Morris, J.~Mustard, S.~Pelkey, E.~Rhodes, M.~Robinson,
+  T.~Roush, E.~Schaefer, G.~Seagrave, F.~Seelos, P.~Silverglate, S.~Slavney,
+  M.~Smith, W.-J. Shyong, K.~Strohbehn, H.~Taylor, P.~Thompson, B.~Tossman,
+  M.~Wirzburger, and M.~Wolff (2007), Compact reconnaissance imaging
+  spectrometer for mars (crism) on mars reconnaissance orbiter (mro),
+  \textit{Journal of Geophysical Research: Planets}, \textit{112}(E5),
+  n/a--n/a, \doi{10.1029/2006JE002682}, e05S03.
+
+\bibitem[{\textit{{Pence} et~al.}(2010)\textit{{Pence}, {Chiappetti}, {Page},
+  {Shaw}, and {Stobie}}}]{fitsver3}
+{Pence}, W.~D., L.~{Chiappetti}, C.~G. {Page}, R.~A. {Shaw}, and E.~{Stobie}
+  (2010), {Definition of the Flexible Image Transport System (FITS), version
+  3.0}, \textit{Astronomy and Astrophysics}, \textit{524}, A42,
+  \doi{10.1051/0004-6361/201015362}.
+
+\bibitem[{\textit{{Seidelmann} et~al.}(2005)\textit{{Seidelmann}, {Archinal},
+  {A'Hearn}, {Cruikshank}, {Hilton}, {Keller}, {Oberst}, {Simon}, {Stooke},
+  {Tholen}, and {Thomas}}}]{wgccre2003}
+{Seidelmann}, P.~K., B.~A. {Archinal}, M.~F. {A'Hearn}, D.~P. {Cruikshank},
+  J.~L. {Hilton}, H.~U. {Keller}, J.~{Oberst}, J.~L. {Simon}, P.~{Stooke},
+  D.~J. {Tholen}, and P.~C. {Thomas} (2005), {Report of the IAU/IAG Working
+  Group on Cartographic Coordinates and Rotational Elements: 2003},
+  \textit{Celestial Mechanics and Dynamical Astronomy}, \textit{91}, 203--215,
+  \doi{10.1007/s10569-004-3115-4}.
+
+\bibitem[{\textit{Snyder}(1987)}]{Snyder1987}
+Snyder, J.~P. (1987), Map projections: A working manual, \textit{Tech. rep.},
+  Washington, D.C., report.
+
+\bibitem[{\textit{Thomas et~al.}(2015)\textit{Thomas, Jenness, Economou,
+  Greenfield, Hirst, Berry, Bray, Gray, Muna, Turner, de~Val-Borro,
+  Santander-Vela, Shupe, Good, Berriman, Kitaeff, Fay, Laurino, Alexov, Landry,
+  Masters, Brazier, Schaaf, Edwards, Redman, Marsh, Streicher, Norris, Pascual,
+  Davie, Droettboom, Robitaille, Campana, Hagen, Hartogh, Klaes, Craig, and
+  Homeier}}]{THOMAS2015133}
+Thomas, B., T.~Jenness, F.~Economou, P.~Greenfield, P.~Hirst, D.~Berry,
+  E.~Bray, N.~Gray, D.~Muna, J.~Turner, M.~de~Val-Borro, J.~Santander-Vela,
+  D.~Shupe, J.~Good, G.~Berriman, S.~Kitaeff, J.~Fay, O.~Laurino, A.~Alexov,
+  W.~Landry, J.~Masters, A.~Brazier, R.~Schaaf, K.~Edwards, R.~Redman,
+  T.~Marsh, O.~Streicher, P.~Norris, S.~Pascual, M.~Davie, M.~Droettboom,
+  T.~Robitaille, R.~Campana, A.~Hagen, P.~Hartogh, D.~Klaes, M.~Craig, and
+  D.~Homeier (2015), Learning from fits: Limitations in use in modern
+  astronomical research, \textit{Astronomy and Computing}, \textit{12}, 133 --
+  145, \doi{https://doi.org/10.1016/j.ascom.2015.01.009}.
+
+\bibitem[{\textit{{Wells} et~al.}(1981)\textit{{Wells}, {Greisen}, and
+  {Harten}}}]{fitsorig}
+{Wells}, D.~C., E.~W. {Greisen}, and R.~H. {Harten} (1981), {FITS - a Flexible
+  Image Transport System}, \textit{Astronomy and Astrophysics Supplement
+  Series}, \textit{44}, 363.
+
+\bibitem[{\textit{Zinzi et~al.}(2016)\textit{Zinzi, Capria, Palomba, Giommi,
+  and Antonelli}}]{ZINZI2016}
+Zinzi, A., M.~Capria, E.~Palomba, P.~Giommi, and L.~Antonelli (2016), Matisse:
+  A novel tool to access, visualize and analyse data from planetary exploration
+  missions, \textit{Astronomy and Computing}, \textit{15}, 16 -- 28,
+  \doi{https://doi.org/10.1016/j.ascom.2016.02.006}.
+
+\end{thebibliography}
 
@@
 %
-%% ------------------------------------------------------------------------ %%
-%
-%  EQUATIONS
-%
-%% ------------------------------------------------------------------------ %%
-
-% Single-line equations are centered.
-% Equation arrays will appear left-aligned.
-
-Math coded inside display math mode \[ ...\]
- will not be numbered, e.g.,:
- \[ x^2=y^2 + z^2\]
-
- Math coded inside \begin{equation} and \end{equation} will
- be automatically numbered, e.g.,:
- \begin{equation}
- x^2=y^2 + z^2
- \end{equation}
-
-% To create multiline equations, use the
-% \begin{eqnarray} and \end{eqnarray} environment
-% as demonstrated below.
-\begin{eqnarray}
-  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
-        && + (y - y_{0}) \sin \Theta  \nonumber \\
-  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
-        && + (y - y_{0}) \cos \Theta.
-\end{eqnarray}
-
-%If you don't want an equation number, use the star form:
-%\begin{eqnarray*}...\end{eqnarray*}
-
-% Break each line at a sign of operation
-% (+, -, etc.) if possible, with the sign of operation
-% on the new line.
-
-% Indent second and subsequent lines to align with
-% the first character following the equal sign on the
-% first line.
-
-% Use an \hspace{} command to insert horizontal space
-% into your equation if necessary. Place an appropriate
-% unit of measure between the curly braces, e.g.
-% \hspace{1in}; you may have to experiment to achieve
-% the correct amount of space.
-
-%% ------------------------------------------------------------------------ %%
-%
-%  EQUATION NUMBERING: COUNTER
-%
-%% ------------------------------------------------------------------------ %%
-
-% You may change equation numbering by resetting
-% the equation counter or by explicitly numbering
-% an equation.
-
-% To explicitly number an equation, type \eqnum{}
-% (with the desired number between the brackets)
-% after the \begin{equation} or \begin{eqnarray}
-% command.  The \eqnum{} command will affect only
-% the equation it appears with; LaTeX will number
-% any equations appearing later in the manuscript
-% according to the equation counter.
-%
-
-% If you have a multiline equation that needs only
-% one equation number, use a \nonumber command in
-% front of the double backslashes (\\) as shown in
-% the multiline equation above.
-
-% If you are using line numbers, remember to surround
-% equations with \begin{linenomath*}...\end{linenomath*}
-
-%  To add line numbers to lines in equations:
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}

### 2018-05-11 ess_marmo.tex
Finalizing comments.

@@
 \caption{Codes to be used in CTYPE FITS keyword in order to identify the planetary body.
-Codes for planets are assumed in \citet{calagreisenI}.
+Codes for planets are implied in \citet{calagreisenI}.
 We propose to add codes for classes of targets,
@@
 It contains string identifying the organization or institution responsible for
-creating the FITS file.}
+creating the FITS file.
+No clear correspondence for this reserved FITS keyword is available in PDS4.
+This is because PDS4 is focused on referencing the authors of the data collection
+and its metadata curation, not the data producer itself.
+Yet, the content of the \texttt{ORIGIN} keyword can be inserted in a PDS4 comment.}
 

### 2018-05-11 ess_marmo.tex
Typo.

@@
 \centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
-\caption{left: Mercury South Pole FITS image displayed in ds9, alternate coordinates in meters
+\caption{left: Mercury South Pole FITS images displayed in ds9, alternate coordinates in meters
 are displayed;

### 2018-05-12 ess_marmo.tex
Typo again..
Update on Overleaf.

@@
 
-We are aware that FITS is today subject of questioning in astronomical applications
-\citep{THOMAS2015133}.
-We are not convinced that one format can solve all the issues faced from long-term
-archiving to on-line streaming.
-Planetary science research community has used a large variety of format customizing
-specific software.
-We want to stress here that using FITS could be an efficient compromise, reducing custom
-data format conversions during the processing chain and
-allowing easy conversion for final products following the needed application.
+We understand that some believe FITS has grown too old for newer astronomical
+applications \citep{THOMAS2015133}. We are also not convinced that one format
+can solve all the issues faced from long-term archiving to on-line streaming.
+But we want to stress that using FITS can be an efficient compromise.
+FITS can become the format that allows the astronomy and the  planetary science 
+research communities to more easily share data across domains, allowing for
+reduced data format conversions during processing and easy conversion for final
+archival products.
 

### 2018-08-22 ess_marmo.tex
Fixes minor comments from first reviewer.

@@
 Some Rosetta data, from NAVCAM and OSIRIS cameras, are distributed in FITS format at
-the ESA Planetary Science Archive\footnote{https://archives.esac.esa.int/psa} (PSA).
+the ESA Planetary Science Archive\footnote{http://psa.esa.int} (PSA).
 In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed
@@
 as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
-description.
+description keywords in the header.
 \begin{figure}[ht!]
@@
 
+The integration of FITS specifications in the GDAL library will be finalized soon.
 In the framework of the VESPA project a review of the available and planned mission data
@@
 with a possible on-demand conversion in FITS format.
-The integration of FITS specifications in the GDAL library will be soon finalize with the goal to
-improve data processing efficiency in planetary science.
 

### 2018-08-22 ess_marmo.tex
Adds acronym table

@@
 We will also show how FITS metadata can easily be transformed
-in the PDS4 metadata archival model and lastly provide example implementations.
+in the Planetary Data System (PDS4) metadata archival model and lastly provide example 
+implementations.
 More than imposing a formal data model, a FITS description for planetary data aims
@@
 
-This paper proposes to use Flexible Image Transport
-System\footnote{https://fits.gsfc.nasa.gov/}
-(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
+This paper proposes to use FITS\footnote{https://fits.gsfc.nasa.gov/}
+format \citep{fitsorig,fitsver3} in planetary surface investigations,
 and describes how FITS metadata can easily be inserted in the
@@
 and spatial missions.
-The International Astronomical Union approved FITS as the standard format
+The International Astronomical Union (IAU) approved FITS as the standard format
 for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
 Therefore, FITS is one of the standard formats implemented in the Virtual
-Observatory (VO).
+Observatory.
 FITS data storage is compatible with the Planetary Data System (PDS)
@@
 ensuring that cartographic endeavors are effectively comparable.
-In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
+In the framework of the Virtual European Solar and Planetary Access\footnote{http://europlanet-vespa.eu/} (VESPA)
 \citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
@@
 
-FITS World Coordinate System
+FITS World Coordinate System (WCS)
 representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
@@
 As an example, the HiRISE team has chosen to distribute HiRISE projected
-products using Equirectangular projection (CAR) on a local surface, approximated
+products using Equirectangular projection on a local surface, approximated
 by the sphere with radius equivalent to the local Mars spheroid radius.
@@
 remaining three characters specify an algorithm code for computing the world coordinate
-value (\ref{table:proj}).
+value (see Table \ref{table:proj}).
 For angular coordinates \citet{calagreisenI} provides the form xLN xLT or yzLN yzLT,
@@
 matrix PCi\_j).
-They correspond to the \textit{pixel\_resolution\_x} and
+They correspond to the \texttt{pixel\_resolution\_x} and
 \texttt{pixel\_resolution\_y} attributes in PDS4, to the \texttt{spatial\_resolution} parameter
@@
 The description of a particular coordinate for an axis in a particular WCS (up to 68
-characters), while the WCSNAMEa keyword names the particular WCS as a whole.
+characters), while the WCSNAMEn keyword names the particular WCS as a whole.
 Its default value will be blank.}
@@
 & & & `C' is normal to the plane defined by `A' and `B'. Always in meters. \\
-OGCCODE	& string & reserved, optional & Value field contains a string describing the standard OGC code \\
+OGCCODE	& string & reserved, optional & Value field contains a string describing the standard Open Geospatial Consortium (OGC) code \\
 & & & (if any) from \citet{hare2006} corresponding to the shape and projection used in the image. \\
@@
 unless another interpretation is clearly explained in the comment field.
-It is assumed to be expressed in UTC but other systems are accepted as long as they
+It is assumed to be expressed in Universal Time (UTC) but other systems are accepted as long as they
 are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
@@
 databases\footnote{http://adswww.harvard.edu/}
-bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
+bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org} (DOI)
 be used.
@@
 surface, FITS and GeoTIFF are completely interchangeable.
-As an example we have downloaded a Mercury polar map from the USGS
+As an example we have downloaded a Mercury polar map from the United States Geological Survey (USGS)
 PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted it in FITS.
@@
 after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
-whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
+whose output is a \texttt{.vrt} GDAL Virtual Header pointing to the FITS file with the appropriate
 projection.
@@
 We understand that some believe FITS has grown too old for newer astronomical
-applications \citep{THOMAS2015133}. We are also not convinced that one format
+applications \citep[e.g.][]{THOMAS2015133}. We are also not convinced that one format
 can solve all the issues faced from long-term archiving to on-line streaming.
@@
 % Acronyms
-% \begin{acronyms}
-% \acro{Acronym}
-% Definition here
-% \acro{EMOS}
-% Ensemble model output statistics 
-% \acro{ECMWF}
-% Centre for Medium-Range Weather Forecasts
-% \end{acronyms}
+\begin{acronyms}
+\acro{AMICA}
+Asteroid Multi-band Imaging Camera
+\acro{BIL}
+Band Interleaved per Line
+\acro{BSQ}
+Band SeQuential
+\acro{CRISM}
+Compact Reconnaissance Imaging Spectrometer for Mars
+\acro{DOI}
+Digital Object Identifier
+\acro{EAICD}
+Experiment Archive Interface Control Document
+\acro{EPN-TAP}
+EuroPlaNet Table Access Protocol
+\acro{ESA}
+European Space Agency
+\acro{FITS}
+Flexible Image Transport System
+\acro{GDAL}
+Geospatial Data Abstraction Library
+\acro{GeoTIFF}
+Geographical Tagged Image File Format
+\acro{GIS}
+Geographical Information Systems
+\acro{HiPS}
+Hierarchical Progressive Survey
+\acro{HiRISE}
+High Resolution Imaging Science Experiment
+\acro{IAU}
+International Astronomical Union
+\acro{IEEE}
+Institute of Electrical and Electronics Engineers
+\acro{IVOA}
+International Virtual Observatory Alliance
+\acro{ISO}
+International Organization for Standardization
+\acro{JAXA}
+Japan Aerospace Exploration Agency
+\acro{MATISSE}
+Multi-purpose Advanced Tool for the Instruments for the Solar System Exploration
+\acro{MEF}
+Multi Extension FITS
+\acro{NAVCAM}
+NAVigation CAMera
+\acro{NIRS}
+Near-Infrared Spectrometer
+\acro{OGC}
+Open Geospatial Consortium
+\acro{OMEGA}
+Observatoire pour la Min\'eralogie, l'Eau, les Glaces et l'Activit\'e
+\acro{OSIRIS}
+Optical, Spectroscopic, and Infrared Remote Imaging System
+\acro{PDS}
+Planetary Data System
+\acro{PDS3}
+Planetary Data System (version 3)
+\acro{PDS4}
+Planetary Data System (version 4)
+\acro{PSA}
+Planetary Space Archive
+\acro{SSDC-ASI}
+Space Science Data Center of the Italian Space Agency
+\acro{USGS}
+United States Geological Survey
+\acro{UTC}
+Universal Time Coordinated
+\acro{VESPA}
+Virtual European Solar and Planetary Access
+\acro{VO}
+Virtual Observatory
+\acro{WCS}
+World Coordinate System
+\acro{WGCCRE}
+Working Group on Cartographic Coordinates and Rotational Elements
+\acro{WKT}
+Well Known Text
+\end{acronyms}
 

### 2018-08-27 ess_marmo.tex
Why FITS and why now in the introduction.

@@
 archiving specifications so that FITS files can be embedded in PDS datasets.
-Finally, it is supported by a large number of open libraries
-and software tools.
+
+FITS format is supported by a large number of open libraries
+and software tools, developed in numerous languages.
+The reference C and Fortran library, CFITSIO\footnote{https://heasarc.gsfc.nasa.gov/docs/software/fitsio/fitsio.html},
+is maintained by the NASA's High Energy Astrophysics Science Archive Research Center,
+and it comes with a number of interfaces for other languages (C++, C\#, Perl, Python, MatLab, etc.).
+Independent libraries are available for example in JAVA, go, Python, R, IDL, and many
+other languages\footnote{A non exhaustive list is available at https://fits.gsfc.nasa.gov/fits\_libraries.html}.
+Software tools for visualization and analysis make use of those libraries or implement
+their own, having different levels of maturity and option completness\footnote{See https://fits.gsfc.nasa.gov/fits\_viewer.html}.
 
@@
 as FITS files with PDS3 labels.
-However, to be more efficiently used in planetary surface investigations, FITS metadata
+
+%However, INSERT HERE COMMENTS FROM REVIEWER #2.
+%\textit{3.
+A bigger picture question that this paper should address is, "why hasn't this already happened?" Again, FITS is over 30 years old.
+What barriers or disincentives have been or currently are in place for the planetary data community preventing or slowing a move toward FITS?
+%The paper should then describe why this proposed scheme reduces or bypasses the relevant barriers, or provides incentives to make the move.
+Is it just tools, as I note above?
+Or are there community reasons, e.g.
+that astronomical and planetary communities have been disjoint to such an extent that languages, standards, etc have never dovetailed?
+What are the authors' perspectives on such barriers/disincentives, and how the present paper smooths over or bypasses them?}
+
+However, solar system imagery data has been traditionally described using terrestrial based
+reference systems.
+Engineers and cartographers working in spatial missions are closer to Earth observation
+techniques and standards than the astronomy one.
+Furthermore, geospatial standardization in FITS dates back to the 2000s, when planetary
+surface research habits had already been installed.
+Where astronomers and planetary scientists work closer, in small bodies investigations,
+FITS format is already more popular.
+
+Why is changing?
+
+- Too many raster images : problem of processing efficiency.
+- Europe and Asia bring more astronomers to the surface planetary investigation.
+- More international collaboration.
+
+This paper:
+
+- specifically addresses the problem of interoperability between astronomy and terrestrial remote sensing tools
+- proposes a mapping between geospatial concepts and astronomical coordinates description
+- proposes to fill the 3d description missing in FITS metadata.
+
+To be more efficiently used in planetary surface investigations, FITS metadata
 must be extended in order to take into account the size of the reference body and

### 2018-08-27 ess_marmo.tex
Finalizes big picture question in FITS description.

@@
 
-%However, INSERT HERE COMMENTS FROM REVIEWER #2.
-%\textit{3.
-A bigger picture question that this paper should address is, "why hasn't this already happened?" Again, FITS is over 30 years old.
-What barriers or disincentives have been or currently are in place for the planetary data community preventing or slowing a move toward FITS?
-%The paper should then describe why this proposed scheme reduces or bypasses the relevant barriers, or provides incentives to make the move.
-Is it just tools, as I note above?
-Or are there community reasons, e.g.
-that astronomical and planetary communities have been disjoint to such an extent that languages, standards, etc have never dovetailed?
-What are the authors' perspectives on such barriers/disincentives, and how the present paper smooths over or bypasses them?}
-
-However, solar system imagery data has been traditionally described using terrestrial based
-reference systems.
-Engineers and cartographers working in spatial missions are closer to Earth observation
+However, solar system imagery data have been traditionally described using terrestrial based
+reference systems and remote sensing formats.
+This is particularly true for planetary surface investigations.
+Engineers and cartographers working in spatial missions are more familiar with Earth observation
 techniques and standards than the astronomy one.
-Furthermore, geospatial standardization in FITS dates back to the 2000s, when planetary
-surface research habits had already been installed.
-Where astronomers and planetary scientists work closer, in small bodies investigations,
-FITS format is already more popular.
-
-Why is changing?
-
-- Too many raster images : problem of processing efficiency.
-- Europe and Asia bring more astronomers to the surface planetary investigation.
-- More international collaboration.
-
-This paper:
-
-- specifically addresses the problem of interoperability between astronomy and terrestrial remote sensing tools
-- proposes a mapping between geospatial concepts and astronomical coordinates description
-- proposes to fill the 3d description missing in FITS metadata.
+Furthermore, the standardization of the spatial references in FITS dates back to the 2000s,
+when planetary surface research habits had already been installed.
+In small bodies investigations, where astronomers and planetary scientists have worked
+together for longer, FITS format is already more popular.
+
+With the impressive increasing of raster data from planetary missions, the efficient automatization
+of data processing is an issue.
+Metadata, historically optimized for archiving, are not always ready for efficient processing.
+FITS metadata schema can be a solution to this problem.
 
 To be more efficiently used in planetary surface investigations, FITS metadata
-must be extended in order to take into account the size of the reference body and
+must be thouroughly mapped to planetary geospatial concepts.
+Then, they must be extended in order to take into account the size of the reference body and
 orientation as standardized by the IAU Working Group on Cartographic Coordinates and
@@
 \label{sec:softdev}
+%\textit{2.
+This leads to my second point, which is that section 3 needs more explanation.
+At minimum, there should be an introduction to the section that describes why these specific tools are featured.
+If the paper provides more description up front of the current FITS tool ecosystem, this section 3 could then be contextualized as illustrating how the FITS tool ecosystem is still expanding, and/or that for the planetary data community to fully take advantage of FITS, more expansion is possible, and already in progress.}
+%\textit{3.
+A bigger picture question that this paper should address is, "why hasn't this already happened?" Again, FITS is over 30 years old.
+What barriers or disincentives have been or currently are in place for the planetary data community preventing or slowing a move toward FITS?
+%The paper should then describe why this proposed scheme reduces or bypasses the relevant barriers, or provides incentives to make the move.
+Is it just tools, as I note above?
+Or are there community reasons, e.g.
+that astronomical and planetary communities have been disjoint to such an extent that languages, standards, etc have never dovetailed?
+What are the authors' perspectives on such barriers/disincentives, and how the present paper smooths over or bypasses them?}
+%This paper:
+
+%- specifically addresses the problem of interoperability between astronomy and terrestrial remote sensing tools
+%- proposes a mapping between geospatial concepts and astronomical coordinates description
+%- proposes to fill the 3d description missing in FITS metadata.
 

### 2018-08-28 ess_marmo.tex
Finalizing comments of reviewer #2

@@
 This is particularly true for planetary surface investigations.
-Engineers and cartographers working in spatial missions are more familiar with Earth observation
-techniques and standards than the astronomy one.
+Engineers and cartographers working in the first spatial missions were more familiar with
+Earth observation techniques and standards than the astronomy one.
 Furthermore, the standardization of the spatial references in FITS dates back to the 2000s,
@@
 
-With the impressive increasing of raster data from planetary missions, the efficient automatization
-of data processing is an issue.
-Metadata, historically optimized for archiving, are not always ready for efficient processing.
+With the impressive increasing of raster data from planetary missions, the efficient
+automatization of data processing is an issue.
+Metadata, historically optimized for archiving, are not always ready for efficient
+processing.
 FITS metadata schema can be a solution to this problem.
@@
 \label{sec:softdev}
-%\textit{2.
-This leads to my second point, which is that section 3 needs more explanation.
-At minimum, there should be an introduction to the section that describes why these specific tools are featured.
-If the paper provides more description up front of the current FITS tool ecosystem, this section 3 could then be contextualized as illustrating how the FITS tool ecosystem is still expanding, and/or that for the planetary data community to fully take advantage of FITS, more expansion is possible, and already in progress.}
-%\textit{3.
-A bigger picture question that this paper should address is, "why hasn't this already happened?" Again, FITS is over 30 years old.
-What barriers or disincentives have been or currently are in place for the planetary data community preventing or slowing a move toward FITS?
-%The paper should then describe why this proposed scheme reduces or bypasses the relevant barriers, or provides incentives to make the move.
-Is it just tools, as I note above?
-Or are there community reasons, e.g.
-that astronomical and planetary communities have been disjoint to such an extent that languages, standards, etc have never dovetailed?
-What are the authors' perspectives on such barriers/disincentives, and how the present paper smooths over or bypasses them?}
-%This paper:
-
-%- specifically addresses the problem of interoperability between astronomy and terrestrial remote sensing tools
-%- proposes a mapping between geospatial concepts and astronomical coordinates description
-%- proposes to fill the 3d description missing in FITS metadata.
+In section \ref{ssec:geofits} the mapping between planetary metadata concepts and
+FITS metadata description has been addressed.
+In order to impulse the use of FITS in planetary sciences this section will provide
+some example efforts of implementation of the proposed mapping.
+Our goal is, on remote sensing side, to get FITS into habits of geospatial developers,
+on the other, to benefit planetary astronomers of FITS flexibility providing a processing
+efficient description of their data.
 
@@
 and VO services.
-The next step to VO and GIS tool integration will be to directly implement metadata 
-translation into GDAL C++ library.
+The next step will be to directly implement metadata translation into the GDAL library. 
+This work is already in progress, and new developments are available from the VESPA github
+repository\footnote{https://github.com/epn-vespa/gdal}.
 
@@
 making use of a Virtual GDAL header, as described above.
-The complete implementation of FITS metadata in GDAL library is in progress.
 Once implemented in GDAL, FITS will officially become a supported format by QGIS.

### 2018-08-29 ess_marmo.tex
Update on Overleaf.

@@
 However, solar system imagery data have been traditionally described using terrestrial based
-reference systems and remote sensing formats.
+geospatial descriptions and remote sensing formats.
 This is particularly true for planetary surface investigations.
@@
 \subsection{MATISSE}
-MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
+MATISSE\footnote{https://tools.ssdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
 for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool

### 2018-08-30 ess_marmo.tex
Clarifies introduction to Section 3.

@@
 some example efforts of implementation of the proposed mapping.
-Our goal is, on remote sensing side, to get FITS into habits of geospatial developers,
-on the other, to benefit planetary astronomers of FITS flexibility providing a processing
-efficient description of their data.
+On remote sensing side, our goal is to get FITS into habits of geospatial developers,
+improving GDAL and QGIS compatibility with FITS.
+On the other side, planetary astronomers will benefit of a FITS efficient description
+of their data, once the new keywords fully implemented in their traditional tools.
+We provide here two examples chosen within the Europlanet VESPA collaboration.
 

### 2018-08-30 2018EA000388R_Answers_Reviewers.txt
Last edits and answers to the reviewers.

Reviewer 1

> Page 4, the footnote 8 should rather be psa.esa.int (which is indeed
> redirected to the addressed mentioned). In the future, the psa.esa.int
> will certainly remain.

Fixed

> L278 -> headers

Fixed

-------------------------------------------------------------------------------

Reviewer 2

We want to thank the anonymous reviewer for his thoughtful comments.
Sections 2.1 and 3 have been enriched following his suggestions.

> 1. The FITS tool ecosystem - FITS has been around for a long time, and has a
> robust user base. Accordingly, the FITS tool ecosystem is very large and well
> established. This seems like a prime motivator for moving toward FITS.
> This paper notes on lines 72-73 that there is a "large number of open libraries
> and software tools," but does not say much more. It would be useful to note
> that FITS tools exist in numerous languages, including, but not limited to, C,
> C++, IDL, Java, MATLAB, Fortran, R, & Python.

Section 2.1 has been expanded (see paragraph starting at L68 in trackchanges.pdf),
references to available tools have been added in footnotes.

> 2. This leads to my second point, which is that section 3 needs more explanation.
> At minimum, there should be an introduction to the section that describes why
> these specific tools are featured. If the paper provides more description up front
> of the current FITS tool ecosystem, this section 3 could then be contextualized as
> illustrating how the FITS tool ecosystem is still expanding, and/or that for the
> planetary data community to fully take advantage of FITS, more expansion is possible,
> and already in progress.

A small introduction to Section 3 has been added, in particular clarifying, we hope,
the choice of the described tools.

> 3. A bigger picture question that this paper should address is, "why hasn't this
> already happened?" Again, FITS is over 30 years old. What barriers or disincentives
> have been or currently are in place for the planetary data community preventing or
> slowing a move toward FITS? The paper should then describe why this proposed scheme
> reduces or bypasses the relevant barriers, or provides incentives to make the move.
> Is it just tools, as I note above? Or are there community reasons, e.g. that
> astronomical and planetary communities have been disjoint to such an extent that
> languages, standards, etc have never dovetailed? What are the authors' perspectives
> on such barriers/disincentives, and how the present paper smooths over or bypasses them?

Section 2.1 has been expanded (see paragraph starting at L86 in trackchanges.pdf).
We hope this syntetic answer will be enough to clarify our point of view.
A more detailed discussion have been omitted due to overall length.

> 4. Editorial comment - the paper is heavy on acronyms, some of which are not spelled
> out at all, e.g. VRT, some of which are only spelled out in the titles of papers in the
> bibliography but not in the text (VESPA, CRISM, HiRISE).
> All acronyms should be spelled out, and/or a table of acronyms should be added.

Fixed.

### 2018-08-30 ess_marmo.tex
Last edits and answers to the reviewers.

@@
 as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
-description keywords in the header.
+description.
 \begin{figure}[ht!]
@@
 are displayed;
-right: GeoTIFF raster (displayed in red) and the two virtual haeders
+right: GeoTIFF raster (displayed in red) and the two virtual headers
 linked to FITS rasters (displayed in green and blue); the three layers
